---
title: "normalisation + clustering"
format: html
editor: visual
---

```{r}
# Loading relevant libraries 
library(tidyverse) # includes ggplot2, for data visualisation. dplyr, for data manipulation.
library(Seurat)
library(scuttle)
```

```{r}
poolA_et_B <- readRDS("data_filtered_with_adt_qc_outlierdrop.rds")
```

## Normalisation

normalisation “LogNormalize” that normalizes the feature expression measurements for each cell by the total expression, multiplies this by a scale factor (10,000 by default), and log-transforms the result.

```{r}
poolA_et_B <- NormalizeData(poolA_et_B) 
```

## Identification of highly-variable genes

```{r}
poolA_et_B <- FindVariableFeatures(poolA_et_B, nfeatures = 2000)

# Identify the top 10 Highly Variable Genes (HVGs) poolA et B
top10_poolA_et_B <- head(VariableFeatures(poolA_et_B), 10)
top10_plot_poolA_et_B <- VariableFeaturePlot(poolA_et_B)
LabelPoints(plot = top10_plot_poolA_et_B, points = top10_poolA_et_B, repel = TRUE)
```

### nbr gene/cell

```{r}
# Summary table of genes per cell by Sample
VlnPlot(poolA_et_B, features = "nFeature_RNA", group.by = "orig.ident")

# To get the average number per pool:
#tapply(poolA_et_B$nFeature_RNA, poolA_et_B$pool, mean) [cite: 12]
```

### nbr de cell/condition

```{r}
poolA_et_B@meta.data %>%
  count(response) %>%
  ggplot(aes(x = response, y = n, fill = response)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  labs(
    x = "réponse au ttt",
    y = "Nombre de cellules"
  )
```

### nbr de cell/patient

```{r}
poolA_et_B@meta.data %>%
  count(patient) %>%
  ggplot(aes(x = patient, y = n, fill = patient)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  labs(
    x = "patient",
    y = "Nombre de cellules"
  )
```

# re Split the RNA assay by your batch/sample column

```{r}
poolA_et_B <- split(poolA_et_B, f = poolA_et_B$orig.ident)
```

## scaling

```{r}
poolA_et_B <- ScaleData(poolA_et_B)
#view(poolA_et_B@assays$RNA)
```

## PCA

voir si on monte pas à 30

```{r}
poolA_et_B <- RunPCA(poolA_et_B, features = VariableFeatures(poolA_et_B))

#choisir le nbr de PC

ElbowPlot(poolA_et_B, ndims = 50)

# Visualize PCA
VizDimLoadings(poolA_et_B, dims = 1:3, reduction = "pca")
DimPlot(poolA_et_B, reduction = "pca")
```

## Integration

```{r}
# (keeping layers separate)
poolA_et_B <- IntegrateLayers(
  object = poolA_et_B, 
  method = CCAIntegration, # Good for different donors
  orig.reduction = "pca", 
  new.reduction = "integrated.cca",
  verbose = FALSE
)
```

## clustering

tester plsr resolution 

```{r}
DefaultAssay(poolA_et_B)
poolA_et_B <-  FindNeighbors(poolA_et_B, dim = 1:30)
poolA_et_B <- FindClusters(poolA_et_B, resolution = 0.4)

#poolA et B integrated
poolA_et_B <-  FindNeighbors(poolA_et_B, reduction = "integrated.cca", dim = 1:30, k.param = 20)
poolA_et_B <- FindClusters(poolA_et_B, resolution = 0.4, cluster.name = "cca_clusters")
```

## UMAP

```{r}
poolA_et_B <- RunUMAP(poolA_et_B, dims = 1:30)
DimPlot(poolA_et_B, reduction = "umap", label = TRUE)

##poolA et B integrated
poolA_et_B <- RunUMAP(poolA_et_B, reduction = "integrated.cca", dims = 1:30, reduction.name = "umap.cca")
umap_cluster <- DimPlot(poolA_et_B, reduction = "umap.cca", label = TRUE) 
ggsave("umap_cluster.jpeg",umap_cluster )
```

## t-sne

```{r}
poolA_et_B <- RunTSNE(poolA_et_B, dims = 1:20)
DimPlot(poolA_et_B, reduction = "tsne", label = TRUE)

#poolA et B integrated
poolA_et_B <- RunTSNE(poolA_et_B, reduction = "integrated.cca", dims = 1:20, reduction.name = "tnse.cca")
DimPlot(poolA_et_B,  reduction = "tnse.cca", label = TRUE)
```

### umap 1 couleur/patient

```{r}
umap_patient <- DimPlot(poolA_et_B, reduction = "umap.cca", group.by = "patient")
ggsave( "umap_global_coloré_par_patient.jpeg",umap_patient)

umap_per_patient <- DimPlot(poolA_et_B, reduction = "umap.cca", split.by = "orig.ident", ncol = 4)
ggsave("umap_per_patient.jpeg",umap_per_patient, width = 15, height = 10)
```

```{r}
# Transformer 'timepoint' en facteur avec l'ordre souhaité
poolA_et_B$timepoint <- factor(poolA_et_B$timepoint, levels = c("pre", "post"))

umap_response_per_timepoint <- DimPlot(poolA_et_B, reduction = "umap.cca", group.by = "response", split.by = "timepoint")
ggsave( "umap_response_per_timepoint_0_4.jpeg",umap_response_per_timepoint, width = 15, height = 10)
```

```{r}
#évolution des timepoints par groupe
umap_timepoint_per_response <- DimPlot(poolA_et_B, 
        reduction = "umap.cca", 
        group.by = "timepoint", 
        split.by = "response") + 
  theme_bw() +
  ggtitle("Trajectoire des cellules par groupe de réponse")
ggsave( "umap_timepoint_per_response_0_4.jpeg", umap_timepoint_per_response, width = 15, height = 10)
```

### differente resolution

```{r}
#res 0.5
poolA_et_B <- FindClusters(poolA_et_B, resolution = 0.5, cluster.name = "cca_clusters", k.param = 20)
poolA_et_B <- RunUMAP(poolA_et_B, reduction = "integrated.cca", dims = 1:30, reduction.name = "umap.cca")
p1 <-DimPlot(poolA_et_B, reduction = "umap.cca", label = TRUE)

#poolA et B integrated
#poolA_et_B <- RunTSNE(poolA_et_B, reduction = "integrated.cca", dims = 1:20, reduction.name = "tnse.cca")
#DimPlot(poolA_et_B,  reduction = "tnse.cca", label = TRUE)

#res 0.6
poolA_et_B <- FindClusters(poolA_et_B, resolution = 0.6, cluster.name = "cca_clusters", k.param = 20)
poolA_et_B <- RunUMAP(poolA_et_B, reduction = "integrated.cca", dims = 1:30, reduction.name = "umap.cca")
p2 <- DimPlot(poolA_et_B, reduction = "umap.cca", label = TRUE)

#poolA et B integrated
#poolA_et_B <- RunTSNE(poolA_et_B, reduction = "integrated.cca", dims = 1:20, reduction.name = "tnse.cca")
#DimPlot(poolA_et_B,  reduction = "tnse.cca", label = TRUE)

#res 0.7
poolA_et_B <- FindClusters(poolA_et_B, resolution = 0.7, cluster.name = "cca_clusters", k.param = 20)
poolA_et_B <- RunUMAP(poolA_et_B, reduction = "integrated.cca", dims = 1:30, reduction.name = "umap.cca")
p3 <- DimPlot(poolA_et_B, reduction = "umap.cca", label = TRUE)
#poolA et B integrated
#poolA_et_B <- RunTSNE(poolA_et_B, reduction = "integrated.cca", dims = 1:20, reduction.name = "tnse.cca")
#DimPlot(poolA_et_B,  reduction = "tnse.cca", label = TRUE)

#res 0.8
poolA_et_B <- FindClusters(poolA_et_B, resolution = 0.8, cluster.name = "cca_clusters", k.param = 20)
poolA_et_B <- RunUMAP(poolA_et_B, reduction = "integrated.cca", dims = 1:30, reduction.name = "umap.cca")
p4 <- DimPlot(poolA_et_B, reduction = "umap.cca", label = TRUE)
#poolA et B integrated
#poolA_et_B <- RunTSNE(poolA_et_B, reduction = "integrated.cca", dims = 1:20, reduction.name = "tnse.cca")
#DimPlot(poolA_et_B,  reduction = "tnse.cca", label = TRUE)

#res 0.9
poolA_et_B <- FindClusters(poolA_et_B, resolution = 0.9, cluster.name = "cca_clusters_0_9", k.param = 20)
poolA_et_B <- RunUMAP(poolA_et_B, reduction = "integrated.cca", dims = 1:30, reduction.name = "umap.cca_0_9")
p5 <- DimPlot(poolA_et_B, reduction = "umap.cca_0_9", label = TRUE)
#poolA et B integrated
#poolA_et_B <- RunTSNE(poolA_et_B, reduction = "integrated.cca", dims = 1:20, reduction.name = "tnse.cca")
#DimPlot(poolA_et_B,  reduction = "tnse.cca", label = TRUE)

#res 1
poolA_et_B <- FindClusters(poolA_et_B, resolution = 1, cluster.name = "cca_clusters", k.param = 20)
poolA_et_B <- RunUMAP(poolA_et_B, reduction = "integrated.cca", dims = 1:30, reduction.name = "umap.cca")
p6 <- DimPlot(poolA_et_B, reduction = "umap.cca", label = TRUE)
#poolA et B integrated
#poolA_et_B <- RunTSNE(poolA_et_B, reduction = "integrated.cca", dims = 1:20, reduction.name = "tnse.cca")
#DimPlot(poolA_et_B,  reduction = "tnse.cca", label = TRUE)

#res 1.1
poolA_et_B <- FindClusters(poolA_et_B, resolution = 1.1, cluster.name = "cca_clusters", k.param = 20)
poolA_et_B <- RunUMAP(poolA_et_B, reduction = "integrated.cca", dims = 1:30, reduction.name = "umap.cca")
p7 <- DimPlot(poolA_et_B, reduction = "umap.cca", label = TRUE)
#poolA et B integrated
#poolA_et_B <- RunTSNE(poolA_et_B, reduction = "integrated.cca", dims = 1:20, reduction.name = "tnse.cca")
#DimPlot(poolA_et_B,  reduction = "tnse.cca", label = TRUE)

#res 1.2
poolA_et_B <- FindClusters(poolA_et_B, resolution = 1.2, cluster.name = "cca_clusters", k.param = 20)
poolA_et_B <- RunUMAP(poolA_et_B, reduction = "integrated.cca", dims = 1:30, reduction.name = "umap.cca")
p8 <- DimPlot(poolA_et_B, reduction = "umap.cca", label = TRUE)
#poolA et B integrated
#poolA_et_B <- RunTSNE(poolA_et_B, reduction = "integrated.cca", dims = 1:20, reduction.name = "tnse.cca")
#DimPlot(poolA_et_B,  reduction = "tnse.cca", label = TRUE)

resolution <- p1 + p2 + p3 + p4 + p5 + p6 + p7 + p8

ggsave( "resolution.png", resolution , width = 20, height = 18 )
```

### nbr de cell/condition

```{r}
poolA_et_B@meta.data %>%
  count(response) %>%
  ggplot(aes(x = response, y = n, fill = response)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  labs(
    x = "réponse au ttt",
    y = "Nombre de cellules"
  )
```

### nbr de cell/patient

```{r}
poolA_et_B@meta.data %>%
  count(patient) %>%
  ggplot(aes(x = patient, y = n, fill = patient)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  labs(
    x = "patient",
    y = "Nombre de cellules"
  )
```

```{r}
#res 0.8
poolA_et_B <- FindClusters(poolA_et_B, resolution = 0.8, cluster.name = "cca_clusters_0_8", k.param = 20)
poolA_et_B <- RunUMAP(poolA_et_B, reduction = "integrated.cca", dims = 1:30, reduction.name = "umap.cca_0_8")
DimPlot(poolA_et_B, reduction = "umap.cca", label = TRUE)
#poolA et B integrated
#poolA_et_B <- RunTSNE(poolA_et_B, reduction = "integrated.cca", dims = 1:20, reduction.name = "tnse.cca")
#DimPlot(poolA_et_B,  reduction = "tnse.cca", label = TRUE)
```


```{r}
# Transformer 'timepoint' en facteur avec l'ordre souhaité
poolA_et_B$timepoint <- factor(poolA_et_B$timepoint, levels = c("pre", "post"))

umap_response_per_timepoint <- DimPlot(poolA_et_B, reduction = "umap.cca_0_8", group.by = "response", split.by = "timepoint")
ggsave( "umap_response_per_timepoint_0_8.jpeg",umap_response_per_timepoint, width = 15, height = 10)
```

```{r}
#évolution des timepoints par groupe
umap_timepoint_per_response <- DimPlot(poolA_et_B, 
        reduction = "umap.cca_0_8", 
        group.by = "timepoint", 
        split.by = "response") + 
  theme_bw() +
  ggtitle("Trajectoire des cellules par groupe de réponse")
ggsave( "umap_timepoint_per_response_0_8.jpeg", umap_timepoint_per_response, width = 15, height = 10)
```

#save
```{r}
saveRDS(poolA_et_B, file = "data_filtered_with_adt_qc_outlierdrop_norm_dimred.rds")
```
