---
title: "ADT analysis"
format: html
editor: visual
---


```{r}
# Loading relevant libraries 
library(tidyverse) # includes ggplot2, for data visualisation. dplyr, for data manipulation.
library(Seurat)
library(scuttle)
```

```{r}
poolA_et_B <-  readRDS("020_010_042_025_008_034_095_017_filtered_with_adt_qc_outlierdrop_norm_dimred_annot_sex_phase.rds")
```

# ADT

```{r}
DefaultAssay(poolA_et_B)

DefaultAssay(poolA_et_B) <- "ADT"
DefaultAssay(poolA_et_B)

# Extract a list of features measured in the ADT assay
rownames(poolA_et_B[["ADT"]])
```

## Normalisation

normalisation “LogNormalize” that normalizes the feature expression measurements for each cell by the total expression, multiplies this by a scale factor (10,000 by default), and log-transforms the result.

```{r}
poolA_et_B <- NormalizeData(poolA_et_B) 
```

## Identification of highly-variable genes

```{r}
#poolA et B
poolA_et_B <- FindVariableFeatures(poolA_et_B, nfeatures = 2000)

# Identify the top 10 Highly Variable Genes (HVGs) poolA et B
top10_poolA_et_B <- head(VariableFeatures(poolA_et_B), 10)
top10_plot_poolA_et_B <- VariableFeaturePlot(poolA_et_B)
LabelPoints(plot = top10_plot_poolA_et_B, points = top10_poolA_et_B, repel = TRUE)
```

### nbr gene/cell

```{r}
# Summary table of genes per cell by Sample
VlnPlot(poolA_et_B, features = "nFeature_ADT", group.by = "orig.ident")

FeaturePlot(poolA_et_B, features = "CD14-ADT")

# To get the average number per pool:
#tapply(poolA_et_B$nFeature_RNA, poolA_et_B$pool, mean) [cite: 12]
```

## scaling

```{r}
poolA_et_B <- ScaleData(poolA_et_B)
#view(poolA_et_B@assays$ADT)
```

## PCA

voir si on monte pas à 30

```{r}
#poolA et B
poolA_et_B <- RunPCA(poolA_et_B, features = VariableFeatures(poolA_et_B),reduction.name = "adt.pca")

#choisir le nbr de PC

ElbowPlot(poolA_et_B, ndims = 50)

# Visualize PCA
VizDimLoadings(poolA_et_B, dims = 1:3, reduction = "adt.pca")
DimPlot(poolA_et_B, reduction = "adt.pca")
```


## clustering

tester plsr resolution 

```{r}
#poolA et B integrated
poolA_et_B <-  FindNeighbors(poolA_et_B, reduction = "adt.pca", dim = 1:16)
poolA_et_B <- FindClusters(poolA_et_B, resolution = 0.4, cluster.name = "adt.clusters")
```

## UMAP

```{r}
##poolA et B integrated
poolA_et_B <- RunUMAP(poolA_et_B, reduction = "adt.pca", dims = 1:16, reduction.name = "adt.umap")
adt_umap_cluster <- DimPlot(poolA_et_B, reduction = "adt.umap", label = TRUE)
ggsave( "adt_umap_cluster.jpeg",adt_umap_cluster)
```

## t-sne

```{r}
#poolA et B integrated
poolA_et_B <- RunTSNE(poolA_et_B, reduction = "adt.pca", dims = 1:16, reduction.name = "adt.tnse")
DimPlot(poolA_et_B,  reduction = "adt.tnse", label = TRUE)
```

### differente resolution

```{r}
#res 0.5
poolA_et_B <- FindClusters(poolA_et_B, resolution = 0.5, cluster.name = "adt.clusters")
poolA_et_B <- RunUMAP(poolA_et_B, reduction = "adt.pca", dims = 1:16, reduction.name = "umap")
p1_adt <- DimPlot(poolA_et_B, reduction = "adt.umap", label = TRUE)

#res 0.6
poolA_et_B <- FindClusters(poolA_et_B, resolution = 0.6, cluster.name = "adt.clusters")
poolA_et_B <- RunUMAP(poolA_et_B, reduction = "adt.pca", dims = 1:16, reduction.name = "adt.umap")
p2_adt <- DimPlot(poolA_et_B, reduction = "adt.umap", label = TRUE)

#res 0.7
poolA_et_B <- FindClusters(poolA_et_B, resolution = 0.7, cluster.name = "adt.clusters")
poolA_et_B <- RunUMAP(poolA_et_B, reduction = "adt.pca", dims = 1:16, reduction.name = "adt.umap")
p3_adt <- DimPlot(poolA_et_B, reduction = "adt.umap", label = TRUE)

#res 0.8
poolA_et_B <- FindClusters(poolA_et_B, resolution = 0.8, cluster.name = "adt.clusters")
poolA_et_B <- RunUMAP(poolA_et_B, reduction = "adt.pca", dims = 1:16, reduction.name = "adt.umap")
p4_adt <- DimPlot(poolA_et_B, reduction = "adt.umap", label = TRUE)

#res 0.9
poolA_et_B <- FindClusters(poolA_et_B, resolution = 0.9, cluster.name = "adt.clusters")
poolA_et_B <- RunUMAP(poolA_et_B, reduction = "adt.pca", dims = 1:16, reduction.name = "adt.umap")
p5_adt <- DimPlot(poolA_et_B, reduction = "adt.umap", label = TRUE)

#res 1
poolA_et_B <- FindClusters(poolA_et_B, resolution = 1, cluster.name = "adt.clusters")
poolA_et_B <- RunUMAP(poolA_et_B, reduction = "adt.pca", dims = 1:16, reduction.name = "adt.umap")
p6_adt <- DimPlot(poolA_et_B, reduction = "adt.umap", label = TRUE)

#res 1.1
poolA_et_B <- FindClusters(poolA_et_B, resolution = 1.1, cluster.name = "adt.clusters")
poolA_et_B <- RunUMAP(poolA_et_B, reduction = "adt.pca", dims = 1:16, reduction.name = "adt.umap")
p7_adt <- DimPlot(poolA_et_B, reduction = "adt.umap", label = TRUE)

#res 1.2
poolA_et_B <- FindClusters(poolA_et_B, resolution = 1.2, cluster.name = "adt.clusters")
poolA_et_B <- RunUMAP(poolA_et_B, reduction = "adt.pca", dims = 1:16, reduction.name = "adt.umap")
p8_adt <- DimPlot(poolA_et_B, reduction = "adt.umap", label = TRUE)


resolution_adt <- p1_adt + p2_adt + p3_adt + p4_adt + p5_adt + p6_adt + p7_adt + p8_adt

ggsave( "resolution_adt.png", resolution_adt , width = 20, height = 18 )
```

## batch effect check

```{r}
# Check if samples (OB1-OB8) are mixing or separated
p1 <- DimPlot(poolA_et_B, reduction = "adt.umap", group.by = "orig.ident") + ggtitle("UMAP by Sample ID")

# Check if there is a "Pool" effect (PoolA vs PoolB)
# Note: Since you merged poolA and poolB with add.cell.ids, 
# you might need to create a 'pool' column if it's not already in metadata.
# Safer way to assign Pool metadata
poolA_et_B$pool <- ifelse(poolA_et_B$orig.ident %in% c("AML_OB1", "AML_OB2", "AML_OB3", "AML_OB4"), "PoolA", "PoolB")
poolA_et_B$pool <- ifelse(grepl("PoolA", rownames(poolA_et_B@meta.data)), "PoolA", "PoolB")

p2 <- DimPlot(poolA_et_B, reduction = "adt.umap", group.by = "pool") + ggtitle("UMAP by Pool")

p1 + p2

umap_per_patient <- DimPlot(poolA_et_B, reduction = "adt.umap", split.by = "orig.ident", ncol = 4)
ggsave("umap_per_patient.jpeg",umap_per_patient)
```

## join layer

```{r}
# Join layers first
poolA_et_B <- JoinLayers(poolA_et_B)

```

## find marker

```{r}
markers_adt_poolA_et_B <-  FindAllMarkers(
  poolA_et_B,
  only.pos = TRUE,
  min.pct = 0.25,
  logfc.threshold = 0.25
)

write.csv(markers_adt_poolA_et_B, "markers_adt_poolA_et_B_filtered.csv")
```

```{r}
library(dplyr)

top10_adt <- markers_adt_poolA_et_B %>%
  group_by(cluster) %>%
  slice_max(n = 10, order_by = avg_log2FC)

DoHeatmap(poolA_et_B, features = top10_adt$gene)
```

## évolution intra-patient

```{r}
DimPlot(poolA_et_B, reduction = "adt.umap", group.by = "timepoint", split.by = "patient")

DimPlot(poolA_et_B, group.by = "response", split.by = "patient")
```

### proportions par cluster

```{r}
prop.table(table(
  Idents(poolA_et_B),
  poolA_et_B$timepoint,
  poolA_et_B$patient
), margin = 2)
```

### comparer responder vs refractory

```{r}
DimPlot(poolA_et_B, reduction = "adt.umap", group.by = "response")
```

```{r}

#évolution des timepoints par groupe
umap_timepoint_per_response_adt <- DimPlot(poolA_et_B, 
        reduction = "adt.umap", 
        group.by = "timepoint", 
        split.by = "response") + 
  theme_bw() +
  ggtitle("Trajectoire des cellules par groupe de réponse")
ggsave( "umap_timepoint_per_response_adt.jpeg",umap_timepoint_per_response_adt)
```

```{r}
# Transformer 'timepoint' en facteur avec l'ordre souhaité
poolA_et_B$timepoint <- factor(poolA_et_B$timepoint, levels = c("pre", "post"))

umap_response_per_timepoint_adt <- DimPlot(poolA_et_B, reduction = "adt.umap", group.by = "response", split.by = "timepoint")
ggsave( "umap_response_per_timepoint_adt.jpeg",umap_response_per_timepoint_adt)
```

```{r}
library(patchwork) # to put plots side by side

p1 <- DimPlot(poolA_et_B, reduction = "adt.umap", group.by = "response") + ggtitle("Response")
p2 <- DimPlot(poolA_et_B, reduction = "adt.umap", group.by = "seurat_clusters", label = TRUE) + ggtitle("Clusters")

p1 + p2
```

## proportions de cell

```{r}
prop.table(table(
  Idents(poolA_et_B),
  poolA_et_B$response
), margin = 2)
```

## gene commun p1 et p4
```{r}
# ==============================================================================
# 3. EXTRACTION DE LA SIGNATURE COMMUNE (P1 & P4)
# ==============================================================================

# Marqueurs pour P1 (Réfractaire) vs Répondeurs (P2, P3)
markers_P1_adt <- FindMarkers(poolA_et_B, ident.1 = "P1", group.by = "patient", 
                          ident.2 = c("P2", "P3"), only.pos = TRUE)

# Marqueurs pour P4 (Réfractaire) vs Répondeurs (P2, P3)
markers_P4_adt <- FindMarkers(poolA_et_B, ident.1 = "P4", group.by = "patient", 
                          ident.2 = c("P2", "P3"), only.pos = TRUE)

# Intersection : Gènes significatifs (p_val_adj < 0.05) communs aux deux
sig_P1_adt <- rownames(markers_P1_adt[markers_P1_adt$p_val_adj < 0.05, ])
sig_P4_adt <- rownames(markers_P4_adt[markers_P4_adt$p_val_adj < 0.05, ])
genes_communs_adt <- intersect(sig_P1, sig_P4)

# Création du tableau final trié par puissance d'expression (Log2FC)
liste_resistance_adt <- markers_P1_adt[genes_communs_adt, ] %>% arrange(desc(avg_log2FC))

# ==============================================================================
# 4. VISUALISATION FINALE
# ==============================================================================

# Afficher les 10 gènes les plus discriminants
top30_genes_adt <- head(rownames(liste_resistance_adt), 30)

DotPlot(poolA_et_B, features = top30_genes_adt, group.by = "patient") + 
  ggtitle("Top 30 : Signature de Résistance Partagée (P1 & P4)")

# Export de la liste en CSV pour ton rapport
write.csv(liste_resistance_adt, "Signature_Resistance_AML_Commune_filtered_adt.csv")
```

### par cluster

```{r}
Idents(poolA_et_B) <- "seurat_clusters"

markers_cluster <- FindMarkers(
  subset(poolA_et_B, idents = 3),
  group.by = "response",
  ident.1 = "refractory",
  ident.2 = "responder"
)
```

### tableau de comptage

```{r}
table(poolA_et_B$patient, poolA_et_B$timepoint)
table(poolA_et_B$response, poolA_et_B$timepoint)
```

```{r}
DefaultAssay(poolA_et_B) <- "ADT"
p1 <- FeaturePlot(poolA_et_B, "CD64-ADT", cols = c("lightgrey", "darkgreen"), reduction = "adt.umap") + ggtitle("CD64 protein")
DefaultAssay(poolA_et_B) <- "RNA"
p2 <- FeaturePlot(poolA_et_B, "CD64-ADT", reduction = "umap.cca") + ggtitle("CD64 RNA")

# place plots side-by-side
p1 | p2

p22 <- FeaturePlot(poolA_et_B, features = c("HLADR-ADT",  "CD93-ADT" ,  "CD117-ADT",  "CD36-ADT",   "CD56-ADT",   "CD38-ADT", "CD7-ADT",    "CD14-ADT"  , "CD64-ADT",   "CD45RA-ADT", "CD34-ADT",   "CD33-ADT"  , "CD13-ADT" ,  "CD90-ADT"  , "CD97-ADT"  , "CD366-ADT" , "CD371-ADT"),reduction = "umap.cca")
ggsave("featureplot_adt.jpeg", p22, width = 20, height = 20)
```