---
title: "ADT analysis"
format: html
editor: visual
---

```{r}
# Loading relevant libraries 
library(tidyverse) # includes ggplot2, for data visualisation. dplyr, for data manipulation.
library(Seurat)
library(scuttle)
set.seed(123)
```

```{r}
poolA_et_B <-  readRDS("020_010_042_025_008_034_095_017_filtered_with_adt_qc_outlierdrop_norm_dimred_annot_sex_phase.rds")
```

# ADT

```{r}
DefaultAssay(pool2)
#Idents(poolA_et_B) <- "seurat_clusters"
DefaultAssay(pool2) <- "ADT"
DefaultAssay(pool2)

# Extract a list of features measured in the ADT assay
rownames(pool2[["ADT"]])
```

## Normalisation

normalisation “LogNormalize” that normalizes the feature expression measurements for each cell by the total expression, multiplies this by a scale factor (10,000 by default), and log-transforms the result.

```{r}
pool2 <- NormalizeData(pool2) 
```

## Identification of highly-variable genes

```{r}
#poolA et B
pool2 <- FindVariableFeatures(pool2, nfeatures = 2000)

# Identify the top 10 Highly Variable Genes (HVGs) poolA et B
top10_poolA_et_B <- head(VariableFeatures(poolA_et_B), 10)
top10_plot_poolA_et_B <- VariableFeaturePlot(poolA_et_B)
LabelPoints(plot = top10_plot_poolA_et_B, points = top10_poolA_et_B, repel = TRUE)
```

### nbr gene/cell

```{r}
# Summary table of genes per cell by Sample
VlnPlot(pool2, features = "nFeature_ADT", group.by = "orig.ident")

FeaturePlot(poolA_et_B, features = "CD14-ADT")

# To get the average number per pool:
#tapply(poolA_et_B$nFeature_RNA, poolA_et_B$pool, mean) [cite: 12]
```

## scaling

```{r}
pool2 <- ScaleData(pool2)
#view(poolA_et_B@assays$ADT)
```

## PCA

voir si on monte pas à 30

```{r}
#poolA et B
pool2 <- RunPCA(pool2, features = VariableFeatures(pool2),reduction.name = "adt.pca")

#choisir le nbr de PC

ElbowPlot(pool2, ndims = 50,reduction = "adt.pca")

# Visualize PCA
VizDimLoadings(pool2, dims = 1:3, reduction = "adt.pca")
DimPlot(pool2, reduction = "adt.pca")
```

## clustering

tester plsr resolution

```{r}
#poolA et B
pool2 <-  FindNeighbors(pool2, reduction = "adt.pca", dim = 1:16)
pool2 <- FindClusters(pool2, resolution = 0.4, cluster.name = "adt.clusters")
```

## UMAP

```{r}
##poolA et B integrated
pool2 <- RunUMAP(pool2, reduction = "adt.pca", dims = 1:16, reduction.name = "adt.umap")
DimPlot(pool2, reduction = "adt.umap", label = TRUE)
#ggsave( "adt_umap_cluster.jpeg",adt_umap_cluster)
```

## t-sne

```{r}
#poolA et B integrated
poolA_et_B <- RunTSNE(poolA_et_B, reduction = "adt.pca", dims = 1:16, reduction.name = "adt.tnse")
DimPlot(poolA_et_B,  reduction = "adt.tnse", label = TRUE)
```

### differente resolution

```{r}
#res 0.5
poolA_et_B <- FindClusters(poolA_et_B, resolution = 0.5, cluster.name = "adt.clusters")
poolA_et_B <- RunUMAP(poolA_et_B, reduction = "adt.pca", dims = 1:16, reduction.name = "umap")
p1_adt <- DimPlot(poolA_et_B, reduction = "adt.umap", label = TRUE)

#res 0.6
poolA_et_B <- FindClusters(poolA_et_B, resolution = 0.6, cluster.name = "adt.clusters")
poolA_et_B <- RunUMAP(poolA_et_B, reduction = "adt.pca", dims = 1:16, reduction.name = "adt.umap")
p2_adt <- DimPlot(poolA_et_B, reduction = "adt.umap", label = TRUE)

#res 0.7
poolA_et_B <- FindClusters(poolA_et_B, resolution = 0.7, cluster.name = "adt.clusters")
poolA_et_B <- RunUMAP(poolA_et_B, reduction = "adt.pca", dims = 1:16, reduction.name = "adt.umap")
p3_adt <- DimPlot(poolA_et_B, reduction = "adt.umap", label = TRUE)

#res 0.8
poolA_et_B <- FindClusters(poolA_et_B, resolution = 0.8, cluster.name = "adt.clusters")
poolA_et_B <- RunUMAP(poolA_et_B, reduction = "adt.pca", dims = 1:16, reduction.name = "adt.umap")
p4_adt <- DimPlot(poolA_et_B, reduction = "adt.umap", label = TRUE)

#res 0.9
poolA_et_B <- FindClusters(poolA_et_B, resolution = 0.9, cluster.name = "adt.clusters")
poolA_et_B <- RunUMAP(poolA_et_B, reduction = "adt.pca", dims = 1:16, reduction.name = "adt.umap")
p5_adt <- DimPlot(poolA_et_B, reduction = "adt.umap", label = TRUE)

#res 1
poolA_et_B <- FindClusters(poolA_et_B, resolution = 1, cluster.name = "adt.clusters")
poolA_et_B <- RunUMAP(poolA_et_B, reduction = "adt.pca", dims = 1:16, reduction.name = "adt.umap")
p6_adt <- DimPlot(poolA_et_B, reduction = "adt.umap", label = TRUE)

#res 1.1
poolA_et_B <- FindClusters(poolA_et_B, resolution = 1.1, cluster.name = "adt.clusters")
poolA_et_B <- RunUMAP(poolA_et_B, reduction = "adt.pca", dims = 1:16, reduction.name = "adt.umap")
p7_adt <- DimPlot(poolA_et_B, reduction = "adt.umap", label = TRUE)

#res 1.2
poolA_et_B <- FindClusters(poolA_et_B, resolution = 1.2, cluster.name = "adt.clusters")
poolA_et_B <- RunUMAP(poolA_et_B, reduction = "adt.pca", dims = 1:16, reduction.name = "adt.umap")
p8_adt <- DimPlot(poolA_et_B, reduction = "adt.umap", label = TRUE)


resolution_adt <- p1_adt + p2_adt + p3_adt + p4_adt + p5_adt + p6_adt + p7_adt + p8_adt

ggsave( "resolution_adt.png", resolution_adt , width = 20, height = 18 )
```

## batch effect check

```{r}
# Check if samples (OB1-OB8) are mixing or separated
p1 <- DimPlot(poolA_et_B, reduction = "adt.umap", group.by = "orig.ident") + ggtitle("UMAP by Sample ID")

# Check if there is a "Pool" effect (PoolA vs PoolB)
# Note: Since you merged poolA and poolB with add.cell.ids, 
# you might need to create a 'pool' column if it's not already in metadata.
# Safer way to assign Pool metadata
poolA_et_B$pool <- ifelse(poolA_et_B$orig.ident %in% c("AML_OB1", "AML_OB2", "AML_OB3", "AML_OB4"), "PoolA", "PoolB")
poolA_et_B$pool <- ifelse(grepl("PoolA", rownames(poolA_et_B@meta.data)), "PoolA", "PoolB")

p2 <- DimPlot(poolA_et_B, reduction = "adt.umap", group.by = "pool") + ggtitle("UMAP by Pool")

p1 + p2

umap_per_patient <- DimPlot(poolA_et_B, reduction = "adt.umap", split.by = "orig.ident", ncol = 4)
ggsave("umap_per_patient.jpeg",umap_per_patient)
```

## join layer

```{r}
# Join layers first
pool2 <- JoinLayers(pool2)

```

## find marker

```{r}
markers_adt_poolA_et_B <-  FindAllMarkers(
  poolA_et_B,
  only.pos = TRUE,
  min.pct = 0.25,
  logfc.threshold = 0.25
)

write.csv(markers_adt_poolA_et_B, "markers_adt_poolA_et_B_filtered.csv")
```

```{r}
library(dplyr)

top10_adt <- markers_adt_poolA_et_B %>%
  group_by(cluster) %>%
  slice_max(n = 10, order_by = avg_log2FC)

DoHeatmap(poolA_et_B, features = top10_adt$gene)
```

## évolution intra-patient

```{r}
DimPlot(poolA_et_B, reduction = "adt.umap", group.by = "timepoint", split.by = "patient")

DimPlot(poolA_et_B, group.by = "response", split.by = "patient")
```

### proportions par cluster

```{r}
prop.table(table(
  Idents(poolA_et_B),
  poolA_et_B$timepoint,
  poolA_et_B$patient
), margin = 2)
```

### comparer responder vs refractory

```{r}
DimPlot(poolA_et_B, reduction = "adt.umap", group.by = "response")
```

```{r}

#évolution des timepoints par groupe
umap_timepoint_per_response_adt <- DimPlot(poolA_et_B, 
        reduction = "adt.umap", 
        group.by = "timepoint", 
        split.by = "response") + 
  theme_bw() +
  ggtitle("Trajectoire des cellules par groupe de réponse")
ggsave( "umap_timepoint_per_response_adt.jpeg",umap_timepoint_per_response_adt)
```

```{r}
# Transformer 'timepoint' en facteur avec l'ordre souhaité
poolA_et_B$timepoint <- factor(poolA_et_B$timepoint, levels = c("pre", "post"))

umap_response_per_timepoint_adt <- DimPlot(poolA_et_B, reduction = "adt.umap", group.by = "response", split.by = "timepoint")
ggsave( "umap_response_per_timepoint_adt.jpeg",umap_response_per_timepoint_adt)
```

```{r}
library(patchwork) # to put plots side by side

p1 <- DimPlot(poolA_et_B, reduction = "adt.umap", group.by = "response") + ggtitle("Response")
p2 <- DimPlot(poolA_et_B, reduction = "adt.umap", group.by = "seurat_clusters", label = TRUE) + ggtitle("Clusters")

p1 + p2
```

## proportions de cell

```{r}
prop.table(table(
  Idents(poolA_et_B),
  poolA_et_B$response
), margin = 2)
```

## gene commun p1 et p4

```{r}
# ==============================================================================
# 3. EXTRACTION DE LA SIGNATURE COMMUNE (P1 & P4)
# ==============================================================================

# Marqueurs pour P1 (Réfractaire) vs Répondeurs (P2, P3)
markers_P1_adt <- FindMarkers(poolA_et_B, ident.1 = "P1", group.by = "patient", 
                          ident.2 = c("P2", "P3"), only.pos = TRUE)

# Marqueurs pour P4 (Réfractaire) vs Répondeurs (P2, P3)
markers_P4_adt <- FindMarkers(poolA_et_B, ident.1 = "P4", group.by = "patient", 
                          ident.2 = c("P2", "P3"), only.pos = TRUE)

# Intersection : Gènes significatifs (p_val_adj < 0.05) communs aux deux
sig_P1_adt <- rownames(markers_P1_adt[markers_P1_adt$p_val_adj < 0.05, ])
sig_P4_adt <- rownames(markers_P4_adt[markers_P4_adt$p_val_adj < 0.05, ])
genes_communs_adt <- intersect(sig_P1, sig_P4)

# Création du tableau final trié par puissance d'expression (Log2FC)
liste_resistance_adt <- markers_P1_adt[genes_communs_adt, ] %>% arrange(desc(avg_log2FC))

# ==============================================================================
# 4. VISUALISATION FINALE
# ==============================================================================

# Afficher les 10 gènes les plus discriminants
top30_genes_adt <- head(rownames(liste_resistance_adt), 30)

DotPlot(poolA_et_B, features = top30_genes_adt, group.by = "patient") + 
  ggtitle("Top 30 : Signature de Résistance Partagée (P1 & P4)")

# Export de la liste en CSV pour ton rapport
write.csv(liste_resistance_adt, "Signature_Resistance_AML_Commune_filtered_adt.csv")
```

### par cluster

```{r}
Idents(poolA_et_B) <- "seurat_clusters"

markers_cluster <- FindMarkers(
  subset(poolA_et_B, idents = 3),
  group.by = "response",
  ident.1 = "refractory",
  ident.2 = "responder"
)
```

### tableau de comptage

```{r}
table(poolA_et_B$patient, poolA_et_B$timepoint)
table(poolA_et_B$response, poolA_et_B$timepoint)
```

```{r}
DefaultAssay(pool_focus) <- "ADT"
p1 <- FeaturePlot(pool_focus, "HLADR-ADT", cols = c("lightgrey", "darkgreen"), reduction = "umap.cca_focus2_0.8") + ggtitle("HLA-DR protein")
DefaultAssay(pool_focus) <- "RNA"
p2 <- FeaturePlot(pool_focus, "HLA-DRA", reduction = "umap.cca_focus2_0.8") + ggtitle("HLA-DR RNA")

# place plots side-by-side
p1 | p2

CD13, CD33, CD14, FCGR3A : mye mature
CD34, CD117, HLA-DR variable, MPO selon cas <- mye immature

Idents(poolA_et_B_downsampled) <- "orig.ident"
p22 <- FeaturePlot(pool_focus2, features = markers_adt, reduction = "umap.cca0", cols = c("grey",  "red"))
print(p22)
ggsave("featureplot_adt_focus.jpeg", p22, width = 25, height = 25)
```

```{r}
library(tidyr)
library(dplyr)
library(ggplot2)

# 1. Agrégat pseudo-bulk (Vérifié avec la colonne 'timepoint')
DefaultAssay(pool_focus2) <- "ADT"
adt_pseudo <- AggregateExpression(pool_focus2, 
                                  group.by = c("patient_response", "timepoint"), 
                                  return.seurat = TRUE)

# Transformation avec les bons séparateurs
df_adt <- as.data.frame(t(GetAssayData(adt_pseudo, layer = "data"))) %>%
  select(any_of(markers_adt)) %>%
  mutate(sample_id = rownames(.)) %>%
  # On coupe au tiret OU à l'underscore
  separate(sample_id, into = c("patient", "status", "timepoint"), sep = "[_-]") %>%
  # On s'assure que la réponse est bien un facteur pour les couleurs
  mutate(response = factor(status, levels = c("refractory", "responder"))) %>%
  pivot_longer(cols = any_of(markers_adt), names_to = "marker", values_to = "expression")

# Forcer l'ordre temporel
df_adt$timepoint <- factor(df_adt$timepoint, levels = c("pre", "post"))

# --- Graphique 1 : Moyenne par Réponse (Time × Response) ---
padt <- ggplot(df_adt, aes(x = timepoint, y = expression, color = response, group = response)) +
  stat_summary(fun = mean, geom = "line", linewidth = 1.2) +
  stat_summary(fun = mean, geom = "point", size = 3) +
  facet_wrap(~ marker, scales = "free_y", ncol = 4) +
  theme_bw() +
  scale_color_manual(values = c("refractory" = "red", "responder" = "blue")) +
  labs(title = "Évolution ADT : Groupes", y = "Expression Normalisée", x = "Temps")

# 1. On s'assure que la colonne patient_label existe et contient les bons noms
df_adt <- df_adt %>%
  mutate(patient_label = paste(patient, status, sep = "_"))

# 2. Graphique avec vos couleurs spécifiques
padt2 <- ggplot(df_adt, aes(x = timepoint, y = expression, 
                                      color = patient_label, 
                                      group = patient_label)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~ marker, scales = "free_y", ncol = 4) +
  theme_bw() +
  # Application de votre palette personnalisée
  scale_color_manual(values = c(
    "P1_refractory" = "#C91837", 
    "P4_refractory" = "#9A68A4",
    "P2_responder"  = "#5386B3", 
    "P3_responder"  = "#44A75D"
  )) +
  labs(title = "Évolution ADT : Patients Individuels", 
       y = "Expression Normalisée", 
       x = "Temps",
       color = "Patient")



# Sauvegarde
ggsave("Interaction_plot_adt_group_poolfocus2.jpeg", padt, width = 10, height = 8)
ggsave("Interaction_plot_adt_patient_poolfocus2.jpeg", padt2, width = 10, height = 8)

```

"HLADR-ADT", "CD93-ADT", "CD117-ADT", "CD36-ADT", "CD56-ADT", "CD38-ADT", "CD7-ADT", "CD14-ADT", "CD64-ADT", "CD45RA-ADT",
"CD34-ADT", "CD33-ADT", "CD13-ADT", "CD90-ADT", "CD97-ADT", "CD366-ADT", "CD371-ADT" 

, split.by = "response_timepoint"

#pool focus
```{r}
Idents(pool2) <- "response_timepoint"
p22 <- FeaturePlot(pool2, features = c("HLADR-ADT", "CD93-ADT", "CD117-ADT", "CD36-ADT", "CD56-ADT", "CD38-ADT", "CD7-ADT", "CD14-ADT", "CD64-ADT", "CD45RA-ADT",
"CD34-ADT", "CD33-ADT", "CD13-ADT", "CD90-ADT", "CD97-ADT", "CD366-ADT", "CD371-ADT"  ),  reduction = "umap.integrated", pt.size = 0.3)
print(p22)
p22 <- p22 & scale_colour_viridis(option = "A")
#ggsave("p22.jpeg", p22, height = 35, width = 13)

Idents(pool_pre_all) <- "response_timepoint"
p22 <- FeaturePlot(pool_pre_all, features = c("HLADR-ADT", "CD93-ADT", "CD117-ADT", "CD36-ADT", "CD56-ADT", "CD38-ADT", "CD7-ADT", "CD14-ADT", "CD64-ADT", "CD45RA-ADT",
"CD34-ADT", "CD33-ADT", "CD13-ADT", "CD90-ADT", "CD97-ADT", "CD366-ADT", "CD371-ADT"  ),  reduction = "umap.integrated", pt.size = 0.3)
print(p22)
p22 <- p22 & scale_colour_viridis(option = "A")

FeaturePlot(pool_pre_all, features = "CD34-ADT",  reduction = "umap.integrated", pt.size = 0.3,cols =  c("grey85","gray88","gray", "red4") ) 
```

# limma
```{r}
DefaultAssay(pool2) <- "ADT"
library(limma)

# Facteurs
patient  <- factor(c("P1","P1","P4","P4","P2","P2","P3","P3"))
time     <- factor(c("pre","post","pre","post","pre","post","pre","post"),
                   levels = c("pre","post"))
response <- factor(c("refractory","refractory","refractory","refractory","responder","responder","responder","responder"),
                   levels = c("refractory","responder"))

# Design avec interaction — patient bloqué
design <- model.matrix(~ 0 + patient + time * response)
# Colonnes créées :
#   timepost        = effet traitement chez réfractaires
#   responseresp    = différence de baseline répondeurs
#   timepost:responseresponder  ← LE TERME D'INTERACTION

#contrast <- makeContrasts(
 # respondervsrefractory = responder - refractory,
 # levels = design
#)

# 1. Aggregate counts by patient and timepoint
# This creates a 'pseudo-bulk' object
pseudo_bulk <- AggregateExpression(pool_focus, 
                                   group.by = c("patient", "timepoint"), 
                                   return.seurat = TRUE)

# 2. Extract the matrix (using the RNA assay)
# Use 'counts' if you plan to use voom(), or 'data' if already normalized
exprs_matrix <- as.matrix(GetAssayData(pseudo_bulk, layer = "data", assay = "ADT"))

# 3. Check the column names
# They should match the order of your 'design' matrix factors
colnames(exprs_matrix)

fit <- lmFit(exprs_matrix, design)
fit <- eBayes(fit)

# Extraire l'interaction avec le nom correct
res_interaction <- topTable(fit,
                            coef = "timepost:responseresponder",
                            adjust.method = "none",
                            number = Inf)

# Vérifier les premiers gènes
head(res_interaction)

# Gènes avec interaction significative
sig_interact <- res_interaction[
+   abs(res_interaction$logFC) > 0.2 & res_interaction$P.Value < 0.05, ]

write.csv(res_interaction, "limma6_gene_unfiltered_20_05.csv")
```

#plot limma

```{r}
#read csv
limma_no_filtred <- read_delim("limma6_gene_unfiltered_20_05.csv")

#filtered a 0.3
sig_genes <- limma_no_filtred[
  abs(limma_no_filtred$logFC) > 0.3]

#créer 2 liste pour les genes up reg chez les repondeur
up_genes <- sig_genes[sig_genes$logFC > 0.3, ]
#et un chez refractaire
down_genes <- sig_genes[sig_genes$logFC < -0.3, ]  
top_genes <- head(sig_genes$...1, 20)  #modifier le nom ...1 selon le nom de colonne sinon empty 

library(reshape2)

exprs_matrix <- as.matrix(GetAssayData(pool_focus, assay = "ADT", layer = "data"))
expr_subset <- exprs_matrix[top_genes, ]

#transpose les colonnes
df <- melt(expr_subset)
colnames(df) <- c("gene", "sample", "expression")

meta <- pool_focus@meta.data 
df$patient  <- meta[match(df$sample, rownames(meta)), "patient_response"]
df$time     <- meta[match(df$sample, rownames(meta)), "timepoint"]
df$response <- meta[match(df$sample, rownames(meta)), "response"] 

library(ggplot2)

p24 <- ggplot(df, aes(x = time, y = expression, color = patient, group = patient)) +
  scale_color_manual(values = c("P1_refractory" = "#C91837"  , "P2_responder" = "#5386B3" , "P3_responder" =  "#44A75D"  , "P4_refractory" =  "#9A68A4"))+
  stat_summary(fun = mean, geom = "line", linewidth = 1.5) +
  stat_summary(fun = mean, geom = "point", size = 3) +
  facet_wrap(~ gene, scales = "free_y") +
  theme_bw() +
  ggtitle("Interaction plot: time × patient")

p24 + expand_limits(y = 0)
print(p24)
#ggsave("Interaction_plot2_time_×_patient.jpeg", p24, width = 9, height = 7)

p25 <- ggplot(df, aes(x = time, y = expression, color = response, group = response, ylim(0))) +
  scale_color_manual(values = c("responder" = "blue", "refractory" = "red"))+
  stat_summary(fun = mean, geom = "line", linewidth = 1.5) +
  stat_summary(fun = mean, geom = "point", size = 3) +
  facet_wrap(~ gene, scales = "free_y") +
  theme_bw() +
  ggtitle("Interaction plot: time × patient")

p25
p25 + expand_limits(y = 0)
#ggsave("Interaction_plot2_time_×_response.jpeg", p25, width = 9, height = 7)
DefaultAssay(pool2) <- "RNA"
```




```{r}
library(ggplot2)

# On génère le plot
p22 <- FeaturePlot(
  pool2, 
  features = c("HLADR-ADT", "CD93-ADT", "CD117-ADT", "CD36-ADT", "CD56-ADT", "CD38-ADT", 
               "CD7-ADT", "CD14-ADT", "CD64-ADT", "CD45RA-ADT", "CD34-ADT", "CD33-ADT", 
               "CD13-ADT", "CD90-ADT", "CD97-ADT", "CD366-ADT", "CD371-ADT"),  
  reduction = "umap.integrated", 
  pt.size = 0.3, 
  combine = TRUE # Important pour pouvoir utiliser & après
) + scale_x_reverse() + scale_y_reverse() + coord_flip()

# On force l'échelle de couleur à se recalculer pour CHAQUE gène individuellement
p22 <- p22 & scale_colour_viridis(option = "A")

print(p22)

ggsave("p22_2.jpeg",p22, width = 15, height = 15)
#p22 <- p22 & scale_colour_viridis(option = "B")

#print(p22)

#p22 <- p22 & scale_colour_viridis(option = "C")

#print(p22)

#p22 <- p22 & scale_colour_viridis(option = "D")

#print(p22)

#p22 <- p22 & scale_colour_viridis(option = "E")

#print(p22)

#p22 <- p22 & scale_colour_viridis(option = "F")

#print(p22)

#p22 <- p22 & scale_colour_viridis(option = "G") 

#print(p22)

#p22 <- p22 & scale_colour_viridis(option = "H")

#print(p22)

p23 <- FeaturePlot(
  pool_pre_all, 
  features = c("HLADR-ADT", "CD93-ADT", "CD117-ADT", "CD36-ADT", "CD56-ADT", "CD38-ADT", 
               "CD7-ADT", "CD14-ADT", "CD64-ADT", "CD45RA-ADT", "CD34-ADT", "CD33-ADT", 
               "CD13-ADT", "CD90-ADT", "CD97-ADT", "CD366-ADT", "CD371-ADT"),  
  reduction = "umap.integrated", 
  pt.size = 0.3, 
  combine = TRUE # Important pour pouvoir utiliser & après
) + scale_x_reverse() + scale_y_reverse() + coord_flip()

# On force l'échelle de couleur à se recalculer pour CHAQUE gène individuellement
p23 <- p23 & scale_colour_viridis(option = "A")

print(p23)
ggsave("p23.jpeg",p23, width = 15, height = 15)
```

```{r}
p22 <- FeaturePlot(
  pool_focus2, 
  features = c("HLADR-ADT", "CD93-ADT", "CD117-ADT", "CD36-ADT", "CD56-ADT", "CD38-ADT", 
               "CD7-ADT", "CD14-ADT", "CD64-ADT", "CD45RA-ADT", "CD34-ADT", "CD33-ADT", 
               "CD13-ADT", "CD90-ADT", "CD97-ADT", "CD366-ADT", "CD371-ADT"),  
  reduction = "umap.cca01", 
  pt.size = 0.3, 
  cols = c("grey", "green", "red"),
  min.cutoff = "q9", # Ajuste ici (ex: q5 ou q10)
  max.cutoff = "q95"  # Pour éviter que 1 seule cellule hyper-positive ne rende tout le reste gris
)
print(p22)
```

```{r}
library(ggplot2)
p22 <- p22 & scale_colour_gradientn(colours = c("grey", "green", "red"))
print(p22)
```


```{r}
FeaturePlot(pool_focus, features = c("CD34-ADT", "CD366-ADT"),reduction = "umap.cca", blend = TRUE)
FeaturePlot(pool_focus, features = c("CD34-ADT", "CD7-ADT"),reduction = "umap.cca_focus2_0.8", blend = TRUE)
FeaturePlot(pool_focus, features = c("CD34-ADT", "CD371-ADT"),reduction = "umap.cca", blend = TRUE)
FeaturePlot(pool_focus, features = c("CD366-ADT", "CD371-ADT"),reduction = "umap.cca", blend = TRUE)
FeaturePlot(pool_focus, features = c("CD34-ADT", "CD366-ADT", "CD371-ADT"), blend = TRUE)
leukemic_cells <- subset(pool_focus, subset = CD34-ADT > 1 & CD366-ADT > 1)
```

```{r}
DefaultAssay(pool_focus) <- "ADT"
pool_focus$lam_sus <- with(FetchData(pool_focus, vars = c("CD34", "CD117", "CD33", "CD7", "CD56", "CD14", "CD64", "HLADR")), ifelse( (CD34 >1 & CD117 >1 & CD33 >1) | (CD34 >1 & CD7>1) | (CD34 >1 & CD56>1) | (CD14>1 & CD64>1 & CD33 >1), "suspect", "non_suspect"))
```

```{r}
FeaturePlot(pool_focus2, features = c("CD34-ADT", "CD117-ADT"), blend = TRUE,reduction = "umap.cca0")
FeaturePlot(pool_focus2, features = c("CD34-ADT", "CD90-ADT"), blend = TRUE,reduction = "umap.cca0")
FeaturePlot(pool_focus2, features = c("CD117-ADT", "CD90-ADT"), blend = TRUE,reduction = "umap.cca0")

FeaturePlot(pool_focus2, features = c("CD45RA-ADT", "CD90-ADT"), blend = TRUE,reduction = "umap.cca0")
FeaturePlot(pool_focus2, features = c("CD123", "CD90-ADT"), blend = TRUE)


FeaturePlot(pool_focus2, features = c("CD34-ADT", "CD117-ADT", "CD90-ADT"), blend = TRUE)

```


#save

```{r}
saveRDS(poolA_et_B, file = "020_010_042_025_008_034_095_017_filtered_with_adt_qc_outlierdrop_norm_dimred_annot_sex_phase_adt.rds")
```
