---
title: "scRNA seq AML 2ndaire"
format: html
editor: visual
---

# Setting up environment:

```{r}
# Clean environment
rm(list = ls(all.names = TRUE)) # will clear all objects including hidden objects
gc() # free up memory and report the memory usage
options(max.print = .Machine$integer.max, scipen = 999, stringsAsFactors = F, dplyr.summarise.inform = F) # avoid truncated output in R console and scientific notation
```

```{r}
# Loading relevant libraries 
library(tidyverse) # includes ggplot2, for data visualisation. dplyr, for data manipulation.
library(Seurat)
```

```{r}
# Set input path
set.seed(42)
```

#import data
```{r}
# Import data 
#poolA
#020_010_D
inputMatrixDirOB1 <- "O:/Equipes_PF_Services/EQ_Duprez/EQ_Duprez/Current Lab Members/Chellal Alyssa/cellranger_output_poolA/poolA_output/outs/per_sample_outs/020_010_D/count/sample_raw_feature_bc_matrix"

if(!dir.exists(inputMatrixDirOB1)){
  stop(paste("ERREUR : Le dossier n'existe pas :", inputMatrixDirOB1))
}

cat("Reading 10X matrix...\n")
OB1 <- Read10X(
  data.dir = inputMatrixDirOB1,
  gene.column = 1,        # utiliser 2 si features.tsv a gene symbol
  unique.features = FALSE,
  strip.suffix = FALSE
)

str(OB1) # Check the multiple modalities (list of matrixes) - we're interested in Gene expression
cts_OB1 <- OB1$`Gene Expression`

#020_010_F
inputMatrixDirOB2 <- "O:/Equipes_PF_Services/EQ_Duprez/EQ_Duprez/Current Lab Members/Chellal Alyssa/cellranger_output_poolA/poolA_output/outs/per_sample_outs/020_010_F/count/sample_raw_feature_bc_matrix"

if(!dir.exists(inputMatrixDirOB2)){
  stop(paste("ERREUR : Le dossier n'existe pas :", inputMatrixDirOB2))
}

cat("Reading 10X matrix...\n")
OB2 <- Read10X(
  data.dir = inputMatrixDirOB2,
  gene.column = 1,        # utiliser 2 si features.tsv a gene symbol
  unique.features = FALSE,
  strip.suffix = FALSE
)

str(OB2) # Check the multiple modalities (list of matrixes) - we're interested in Gene expression
cts_OB2 <- OB2$`Gene Expression`

#042_025_D
inputMatrixDirOB3 <- "O:/Equipes_PF_Services/EQ_Duprez/EQ_Duprez/Current Lab Members/Chellal Alyssa/cellranger_output_poolA/poolA_output/outs/per_sample_outs/042_025_D/count/sample_raw_feature_bc_matrix"

if(!dir.exists(inputMatrixDirOB3)){
  stop(paste("ERREUR : Le dossier n'existe pas :", inputMatrixDirOB3))
}

cat("Reading 10X matrix...\n")
OB3 <- Read10X(
  data.dir = inputMatrixDirOB3,
  gene.column = 1,        # utiliser 2 si features.tsv a gene symbol
  unique.features = FALSE,
  strip.suffix = FALSE
)

str(OB3) # Check the multiple modalities (list of matrixes) - we're interested in Gene expression
cts_OB3 <- OB3$`Gene Expression`

#042_025_F
inputMatrixDirOB4 <- "O:/Equipes_PF_Services/EQ_Duprez/EQ_Duprez/Current Lab Members/Chellal Alyssa/cellranger_output_poolA/poolA_output/outs/per_sample_outs/042_025_F/count/sample_raw_feature_bc_matrix"

if(!dir.exists(inputMatrixDirOB4)){
  stop(paste("ERREUR : Le dossier n'existe pas :", inputMatrixDirOB4))
}

cat("Reading 10X matrix...\n")
OB4 <- Read10X(
  data.dir = inputMatrixDirOB4,
  gene.column = 1,        # utiliser 2 si features.tsv a gene symbol
  unique.features = FALSE,
  strip.suffix = FALSE
)

str(OB4) # Check the multiple modalities (list of matrixes) - we're interested in Gene expression
cts_OB4 <- OB4$`Gene Expression`
```

```{r}
# Import data 
#poolB
#008_034_D

inputMatrixDirOB5 <- "O:/Equipes_PF_Services/EQ_Duprez/EQ_Duprez/Current Lab Members/Chellal Alyssa/cellranger_output_poolB/poolB_output/outs/per_sample_outs/008_034_D/count/sample_raw_feature_bc_matrix"

if(!dir.exists(inputMatrixDirOB5)){
  stop(paste("ERREUR : Le dossier n'existe pas :", inputMatrixDirOB5))
}

cat("Reading 10X matrix...\n")
OB5 <- Read10X(
  data.dir = inputMatrixDirOB5,
  gene.column = 1,        # utiliser 2 si features.tsv a gene symbol
  unique.features = FALSE,
  strip.suffix = FALSE
)

str(OB5) # Check the multiple modalities (list of matrixes) - we're interested in Gene expression
cts_OB5 <- OB5$`Gene Expression`

#008_034_F
inputMatrixDirOB6 <- "O:/Equipes_PF_Services/EQ_Duprez/EQ_Duprez/Current Lab Members/Chellal Alyssa/cellranger_output_poolB/poolB_output/outs/per_sample_outs/008_034_F/count/sample_raw_feature_bc_matrix"

if(!dir.exists(inputMatrixDirOB6)){
  stop(paste("ERREUR : Le dossier n'existe pas :", inputMatrixDirOB6))
}

cat("Reading 10X matrix...\n")
OB6 <- Read10X(
  data.dir = inputMatrixDirOB6,
  gene.column = 1,        # utiliser 2 si features.tsv a gene symbol
  unique.features = FALSE,
  strip.suffix = FALSE
)

str(OB6) # Check the multiple modalities (list of matrixes) - we're interested in Gene expression
cts_OB6 <- OB6$`Gene Expression`


#095_017_D
inputMatrixDirOB7 <- "O:/Equipes_PF_Services/EQ_Duprez/EQ_Duprez/Current Lab Members/Chellal Alyssa/cellranger_output_poolB/poolB_output/outs/per_sample_outs/095_017_D/count/sample_raw_feature_bc_matrix"

if(!dir.exists(inputMatrixDirOB7)){
  stop(paste("ERREUR : Le dossier n'existe pas :", inputMatrixDirOB7))
}

cat("Reading 10X matrix...\n")
OB7 <- Read10X(
  data.dir = inputMatrixDirOB7,
  gene.column = 1,        # utiliser 2 si features.tsv a gene symbol
  unique.features = FALSE,
  strip.suffix = FALSE
)

str(OB7) # Check the multiple modalities (list of matrixes) - we're interested in Gene expression
cts_OB7 <- OB7$`Gene Expression`


#095_017_F
inputMatrixDirOB8 <- "O:/Equipes_PF_Services/EQ_Duprez/EQ_Duprez/Current Lab Members/Chellal Alyssa/cellranger_output_poolB/poolB_output/outs/per_sample_outs/095_017_F/count/sample_raw_feature_bc_matrix"

if(!dir.exists(inputMatrixDirOB8)){
  stop(paste("ERREUR : Le dossier n'existe pas :", inputMatrixDirOB8))
}

cat("Reading 10X matrix...\n")
OB8 <- Read10X(
  data.dir = inputMatrixDirOB8,
  gene.column = 1,        # utiliser 2 si features.tsv a gene symbol
  unique.features = FALSE,
  strip.suffix = FALSE
)

str(OB8) # Check the multiple modalities (list of matrixes) - we're interested in Gene expression
cts_OB8 <- OB8$`Gene Expression`
```

#create seurat object
```{r}
#poolA
OB1 <- CreateSeuratObject(
  counts = cts_OB1,
  project = "AML_OB1",
  min.cells = 3,
  min.features = 200
)

saveRDS(OB1, file = "020_010_D.rds")

OB2 <- CreateSeuratObject(
  counts = cts_OB2,
  project = "AML_OB2",
  min.cells = 3,
  min.features = 200
)

saveRDS(OB2, file = "020_010_F.rds")

OB3 <- CreateSeuratObject(
  counts = cts_OB3,
  project = "AML_OB3",
  min.cells = 3,
  min.features = 200
)

saveRDS(OB3, file = "042_025_D.rds")

OB4 <- CreateSeuratObject(
  counts = cts_OB4,
  project = "AML_OB4",
  min.cells = 3,
  min.features = 200
)

saveRDS(OB4, file = "042_025_F.rds")

```

```{r}
#poolB
OB5 <- CreateSeuratObject(
  counts = cts_OB5,
  project = "AML_OB5",
  min.cells = 3,
  min.features = 200
)

saveRDS(OB5, file = "008_034_D.rds")

OB6 <- CreateSeuratObject(
  counts = cts_OB6,
  project = "AML_OB6",
  min.cells = 3,
  min.features = 200
)

saveRDS(OB6, file = "008_034_F.rds")

OB7 <- CreateSeuratObject(
  counts = cts_OB7,
  project = "AML_OB7",
  min.cells = 3,
  min.features = 200
)

saveRDS(OB7, file = "095_017_D.rds")

OB8 <- CreateSeuratObject(
  counts = cts_OB8,
  project = "AML_OB8",
  min.cells = 3,
  min.features = 200
)

saveRDS(OB8, file = "095_017_F.rds")

```

#merge RDS
```{r}
#merge poolA
poolA <-
  c(
    "C:/myfiles/df_Q1.rds",
    "C:/otherfiles/df_Q2.rds"
  )

rds_combo <- rdsfiles %>%
  map_dfr(readRDS)

saveRDS(poolA, file = "020_010_042_025.rds")

```

```{r}
#merge poolB
poolB <-
  c(
    "C:/myfiles/df_Q1.rds",
    "C:/otherfiles/df_Q2.rds"
  )

rds_combo <- rdsfiles %>%
  map_dfr(readRDS)

saveRDS(poolB, file = "008_034_095_017.rds")

```

```{r}
#merge poolA + poolB
poolA_et_B <-
  c(
    "C:/myfiles/df_Q1.rds",
    "C:/otherfiles/df_Q2.rds"
  )

rds_combo <- rdsfiles %>%
  map_dfr(readRDS)

saveRDS(poolA_et_B, file = "020_010_042_025_008_034_095_017.rds")

```

#QC (quality control)
```{r}
#poolA
poolA[["percent.mt"]] <- PercentageFeatureSet(seu, pattern = "^MT-")
VlnPlot(poolA, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"))
FeatureScatter(poolA, "nCount_RNA", "percent.mt")

#poolB
poolB[["percent.mt"]] <- PercentageFeatureSet(seu, pattern = "^MT-")
VlnPlot(poolB, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"))
FeatureScatter(poolB, "nCount_RNA", "percent.mt")

#poolA et B
poolA_et_B[["percent.mt"]] <- PercentageFeatureSet(seu, pattern = "^MT-")
VlnPlot(poolA_et_B, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"))
FeatureScatter(poolA_et_B, "nCount_RNA", "percent.mt")
```

#filtering
```{r}
#poolA
poolA <- subset(
  poolA,
  subset = nFeature_RNA > 300 &
    nFeature_RNA < 2500 &
    percent.mt < 10
)

#poolB
poolB <- subset(
  poolB,
  subset = nFeature_RNA > 300 &
    nFeature_RNA < 2500 &
    percent.mt < 10
)

#poolA et B
poolA_et_B <- subset(
  poolA_et_B,
  subset = nFeature_RNA > 300 &
    nFeature_RNA < 2500 &
    percent.mt < 10
)
```

#normalisation
“LogNormalize” that normalizes the feature expression measurements for each cell by the total expression, multiplies this by a scale factor (10,000 by default), and log-transforms the result.
```{r}
#poolA
poolA <- NormalizeData(poolA) 

#poolB
poolB <- NormalizeData(poolB) 

#poolA et B
poolA_et_B <- NormalizeData(poolA_et_B) 
```

#Identification of highly-variable genes
```{r}
#poolA
poolA <- FindVariableFeatures(poolA, nfeatures = 2000)

# Identify the top 10 Highly Variable Genes (HVGs) poolA
top10_poolA <- head(VariableFeatures(poolA), 10)
top10_plot_poolA <- VariableFeaturePlot(poolA)
LabelPoints(plot = top10_plot_poolA, points = top10_poolA, repel = TRUE)

#poolB
poolB <- FindVariableFeatures(poolB, nfeatures = 2000)

# Identify the top 10 Highly Variable Genes (HVGs) poolB
top10_poolB <- head(VariableFeatures(poolB), 10)
top10_plot_poolB <- VariableFeaturePlot(poolB)
LabelPoints(plot = top10_plot_poolB, points = top10_poolB, repel = TRUE)

#poolA et B
poolA_et_B <- FindVariableFeatures(poolA_et_B, nfeatures = 2000)

# Identify the top 10 Highly Variable Genes (HVGs) poolA et B
top10_poolA_et_B <- head(VariableFeatures(poolA_et_B), 10)
top10_plot_poolA_et_B <- VariableFeaturePlot(poolA_et_B)
LabelPoints(plot = top10_plot_poolA_et_B, points = top10_poolA_et_B, repel = TRUE)
```

#scaling
```{r}
#poolA
poolA <- ScaleData(poolA)
view(poolA@assays$RNA)

#poolB
poolB <- ScaleData(poolB)
view(poolB@assays$RNA)

#poolA et B
poolA_et_B <- ScaleData(poolA_et_B)
view(poolA_et_B@assays$RNA)

```

#PCA
```{r}
#poolA
poolA <- RunPCA(poolA, features = VariableFeatures(poolA))

#choisir le nbr de PC

ElbowPlot(poolA)

# Visualize PCA
VizDimLoadings(poolA, dims = 1:2, reduction = "pca")
DimPlot(poolA, reduction = "pca")

#poolB
poolB <- RunPCA(poolB, features = VariableFeatures(poolB))

#choisir le nbr de PC

ElbowPlot(poolB)

# Visualize PCA
VizDimLoadings(poolB, dims = 1:2, reduction = "pca")
DimPlot(poolB, reduction = "pca")

#poolA et B
poolA_et_B <- RunPCA(poolA_et_B, features = VariableFeatures(poolA_et_B))

#choisir le nbr de PC

ElbowPlot(poolA_et_B)

# Visualize PCA
VizDimLoadings(poolA_et_B, dims = 1:2, reduction = "pca")
DimPlot(poolA_et_B, reduction = "pca")
```

#clustering
```{r}
#poolA
poolA <-  FindNeighbors(poolA, dim = 1:30)
poolA <- FindClusters(poolA, resolution = 0.4)

#poolB
poolB <-  FindNeighbors(poolB, dim = 1:30)
poolB <- FindClusters(poolB, resolution = 0.4)

#poolA et B
poolA_et_B <-  FindNeighbors(poolA_et_B, dim = 1:30)
poolA_et_B <- FindClusters(poolA_et_B, resolution = 0.4)

```

#UMAP
```{r}
#poolA
poolA <- RunUMAP(poolA, dims = 1:30)
DimPlot(poolA, reduction = "umap", label = TRUE)

#poolB
poolB <- RunUMAP(poolB, dims = 1:30)
DimPlot(poolB, reduction = "umap", label = TRUE)

#poolA et B
poolA_et_B <- RunUMAP(poolA_et_B, dims = 1:30)
DimPlot(poolA_et_B, reduction = "umap", label = TRUE)
```

#save
```{r}
#poolA
saveRDS(poolA, file = "020_010_042_025_qc_drop_adt_integ_clean.rds")

#poolB
saveRDS(poolB, file = "008_034_095_017_qc_drop_adt_integ_clean.rds")

#poolA et B
saveRDS(poolA_et_B, file = "020_010_042_025_008_034_095_017_qc_drop_adt_integ_clean.rds")

```

#annotation
```{r}
#poolA

#poolB

#poolA et B

```

#save final
```{r}
#poolA
saveRDS(poolA, file = "020_010_042_025_qc_drop_adt_integ_clean_annot.rds")

#poolB
saveRDS(poolB, file = "008_034_095_017_qc_drop_adt_integ_clean_annot.rds")

#poolA et B
saveRDS(poolA_et_B, file = "020_010_042_025_008_034_095_017_qc_drop_adt_integ_clean_annot.rds")

```
