---
title: "scRNA seq AML 2ndaire"
format: html
editor: visual
---

# Setting up environment:

```{r}
# Clean environment
rm(list = ls(all.names = TRUE)) # will clear all objects including hidden objects
gc() # free up memory and report the memory usage
options(max.print = .Machine$integer.max, scipen = 999, stringsAsFactors = F, dplyr.summarise.inform = F) # avoid truncated output in R console and scientific notation
```

```{r}
# Loading relevant libraries 
library(tidyverse) # includes ggplot2, for data visualisation. dplyr, for data manipulation.
library(Seurat)
```

```{r}
# Set input path
set.seed(42)
```

# import data

## poolA

```{r}
#020_010_D refractaire
inputMatrixDirOB1 <- "O:/Equipes_PF_Services/EQ_Duprez/EQ_Duprez/Current Lab Members/Chellal Alyssa/cellranger_output_poolA/poolA_output/outs/per_sample_outs/020_010_D/count/sample_raw_feature_bc_matrix"

if(!dir.exists(inputMatrixDirOB1)){
  stop(paste("ERREUR : Le dossier n'existe pas :", inputMatrixDirOB1))
}

cat("Reading 10X matrix...\n")
OB1 <- Read10X(
  data.dir = inputMatrixDirOB1,
  gene.column = 2,        # utiliser 2 si features.tsv a gene symbol
  unique.features = TRUE,
  strip.suffix = FALSE
)

str(OB1) # Check the multiple modalities (list of matrixes) - we're interested in Gene expression
cts_OB1 <- OB1$`Gene Expression`

#020_010_F
inputMatrixDirOB2 <- "O:/Equipes_PF_Services/EQ_Duprez/EQ_Duprez/Current Lab Members/Chellal Alyssa/cellranger_output_poolA/poolA_output/outs/per_sample_outs/020_010_F/count/sample_raw_feature_bc_matrix"

if(!dir.exists(inputMatrixDirOB2)){
  stop(paste("ERREUR : Le dossier n'existe pas :", inputMatrixDirOB2))
}

cat("Reading 10X matrix...\n")
OB2 <- Read10X(
  data.dir = inputMatrixDirOB2,
  gene.column = 2,        # utiliser 2 si features.tsv a gene symbol
  unique.features = TRUE,
  strip.suffix = FALSE
)

str(OB2) # Check the multiple modalities (list of matrixes) - we're interested in Gene expression
cts_OB2 <- OB2$`Gene Expression`

#042_025_D repondeur
inputMatrixDirOB3 <- "O:/Equipes_PF_Services/EQ_Duprez/EQ_Duprez/Current Lab Members/Chellal Alyssa/cellranger_output_poolA/poolA_output/outs/per_sample_outs/042_025_D/count/sample_raw_feature_bc_matrix"

if(!dir.exists(inputMatrixDirOB3)){
  stop(paste("ERREUR : Le dossier n'existe pas :", inputMatrixDirOB3))
}

cat("Reading 10X matrix...\n")
OB3 <- Read10X(
  data.dir = inputMatrixDirOB3,
  gene.column = 2,        # utiliser 2 si features.tsv a gene symbol
  unique.features = TRUE,
  strip.suffix = FALSE
)

str(OB3) # Check the multiple modalities (list of matrixes) - we're interested in Gene expression
cts_OB3 <- OB3$`Gene Expression`

#042_025_F
inputMatrixDirOB4 <- "O:/Equipes_PF_Services/EQ_Duprez/EQ_Duprez/Current Lab Members/Chellal Alyssa/cellranger_output_poolA/poolA_output/outs/per_sample_outs/042_025_F/count/sample_raw_feature_bc_matrix"

if(!dir.exists(inputMatrixDirOB4)){
  stop(paste("ERREUR : Le dossier n'existe pas :", inputMatrixDirOB4))
}

cat("Reading 10X matrix...\n")
OB4 <- Read10X(
  data.dir = inputMatrixDirOB4,
  gene.column = 2,        # utiliser 2 si features.tsv a gene symbol
  unique.features = TRUE,
  strip.suffix = FALSE
)

str(OB4) # Check the multiple modalities (list of matrixes) - we're interested in Gene expression
cts_OB4 <- OB4$`Gene Expression`
```

## poolB

```{r}
#008_034_D repondant

inputMatrixDirOB5 <- "O:/Equipes_PF_Services/EQ_Duprez/EQ_Duprez/Current Lab Members/Chellal Alyssa/cellranger_output_poolB/poolB_output/outs/per_sample_outs/008_034_D/count/sample_raw_feature_bc_matrix"

if(!dir.exists(inputMatrixDirOB5)){
  stop(paste("ERREUR : Le dossier n'existe pas :", inputMatrixDirOB5))
}

cat("Reading 10X matrix...\n")
OB5 <- Read10X(
  data.dir = inputMatrixDirOB5,
  gene.column = 2,        # utiliser 2 si features.tsv a gene symbol
  unique.features = TRUE,
  strip.suffix = FALSE
)

str(OB5) # Check the multiple modalities (list of matrixes) - we're interested in Gene expression
cts_OB5 <- OB5$`Gene Expression`

#008_034_F
inputMatrixDirOB6 <- "O:/Equipes_PF_Services/EQ_Duprez/EQ_Duprez/Current Lab Members/Chellal Alyssa/cellranger_output_poolB/poolB_output/outs/per_sample_outs/008_034_F/count/sample_raw_feature_bc_matrix"

if(!dir.exists(inputMatrixDirOB6)){
  stop(paste("ERREUR : Le dossier n'existe pas :", inputMatrixDirOB6))
}

cat("Reading 10X matrix...\n")
OB6 <- Read10X(
  data.dir = inputMatrixDirOB6,
  gene.column = 2,        # utiliser 2 si features.tsv a gene symbol
  unique.features = TRUE,
  strip.suffix = FALSE
)

str(OB6) # Check the multiple modalities (list of matrixes) - we're interested in Gene expression
cts_OB6 <- OB6$`Gene Expression`


#095_017_D refractaire
inputMatrixDirOB7 <- "O:/Equipes_PF_Services/EQ_Duprez/EQ_Duprez/Current Lab Members/Chellal Alyssa/cellranger_output_poolB/poolB_output/outs/per_sample_outs/095_017_D/count/sample_raw_feature_bc_matrix"

if(!dir.exists(inputMatrixDirOB7)){
  stop(paste("ERREUR : Le dossier n'existe pas :", inputMatrixDirOB7))
}

cat("Reading 10X matrix...\n")
OB7 <- Read10X(
  data.dir = inputMatrixDirOB7,
  gene.column = 2,        # utiliser 2 si features.tsv a gene symbol
  unique.features = TRUE,
  strip.suffix = FALSE
)

str(OB7) # Check the multiple modalities (list of matrixes) - we're interested in Gene expression
cts_OB7 <- OB7$`Gene Expression`


#095_017_F
inputMatrixDirOB8 <- "O:/Equipes_PF_Services/EQ_Duprez/EQ_Duprez/Current Lab Members/Chellal Alyssa/cellranger_output_poolB/poolB_output/outs/per_sample_outs/095_017_F/count/sample_raw_feature_bc_matrix"

if(!dir.exists(inputMatrixDirOB8)){
  stop(paste("ERREUR : Le dossier n'existe pas :", inputMatrixDirOB8))
}

cat("Reading 10X matrix...\n")
OB8 <- Read10X(
  data.dir = inputMatrixDirOB8,
  gene.column = 2,        # utiliser 2 si features.tsv a gene symbol
  unique.features = TRUE,
  strip.suffix = FALSE
)

str(OB8) # Check the multiple modalities (list of matrixes) - we're interested in Gene expression
cts_OB8 <- OB8$`Gene Expression`
```

# create seurat object

## poolA

```{r}
OB1 <- CreateSeuratObject(
  counts = cts_OB1,
  project = "AML_OB1",
  min.cells = 3,
  min.features = 200
)

saveRDS(OB1, file = "020_010_D.rds")

head(rownames(OB1))

OB2 <- CreateSeuratObject(
  counts = cts_OB2,
  project = "AML_OB2",
  min.cells = 3,
  min.features = 200
)

saveRDS(OB2, file = "020_010_F.rds")

head(rownames(OB2))

OB3 <- CreateSeuratObject(
  counts = cts_OB3,
  project = "AML_OB3",
  min.cells = 3,
  min.features = 200
)

saveRDS(OB3, file = "042_025_D.rds")

head(rownames(OB3))

OB4 <- CreateSeuratObject(
  counts = cts_OB4,
  project = "AML_OB4",
  min.cells = 3,
  min.features = 200
)

saveRDS(OB4, file = "042_025_F.rds")

head(rownames(OB4))

```

## poolB

```{r}
OB5 <- CreateSeuratObject(
  counts = cts_OB5,
  project = "AML_OB5",
  min.cells = 3,
  min.features = 200
)

saveRDS(OB5, file = "008_034_D.rds")

head(rownames(OB5))

OB6 <- CreateSeuratObject(
  counts = cts_OB6,
  project = "AML_OB6",
  min.cells = 3,
  min.features = 200
)

saveRDS(OB6, file = "008_034_F.rds")

head(rownames(OB6))

OB7 <- CreateSeuratObject(
  counts = cts_OB7,
  project = "AML_OB7",
  min.cells = 3,
  min.features = 200
)

saveRDS(OB7, file = "095_017_D.rds")

head(rownames(OB7))

OB8 <- CreateSeuratObject(
  counts = cts_OB8,
  project = "AML_OB8",
  min.cells = 3,
  min.features = 200
)

saveRDS(OB8, file = "095_017_F.rds")

head(rownames(OB8))

```

# merge RDS

## merge poolA

```{r}
# Merge OB1 (x) with the others (y)
poolA <- merge(
  x = OB1, 
  y = c(OB2, OB3, OB4), 
  add.cell.ids = c("OB1", "OB2", "OB3", "OB4"), 
  project = "PoolA"
)

# Check the new dimensions
# Should be ~38k genes and the sum of all cells
print(dim(poolA))

saveRDS(poolA, file = "020_010_042_025.rds")

```

## merge poolB

```{r}
# Merge OB1 (x) with the others (y)
poolB <- merge(
  x = OB5, 
  y = c(OB6, OB7, OB8), 
  add.cell.ids = c("OB5", "OB6", "OB7", "OB8"), 
  project = "PoolB"
)

# Check the new dimensions
# Should be ~38k genes and the sum of all cells
print(dim(poolB))

saveRDS(poolB, file = "008_034_095_017.rds")

```

## merge poolA + poolB

```{r}
# Merge poolA and poolB
# 'add.cell.ids' is critical here to avoid duplicate cell names across pools
poolA_et_B <- merge(
  x = poolA, 
  y = poolB, 
  add.cell.ids = c("PoolA", "PoolB"), 
  project = "Full_AML_Cohort"
)

# 1. Check the final cell count
ncol(poolA_et_B)
print(dim(poolA_et_B))
saveRDS(poolA_et_B, file = "020_010_042_025_008_034_095_017.rds")

```

# Ajouter les metadata biologiques

```{r}
poolA_et_B$patient <- NA
poolA_et_B$response <- NA
poolA_et_B$timepoint <- NA
```

```{r}
# Assignation patient
poolA_et_B$patient[poolA_et_B$orig.ident %in% c("AML_OB1","AML_OB2")] <- "P1"
poolA_et_B$patient[poolA_et_B$orig.ident %in% c("AML_OB3","AML_OB4")] <- "P2"
poolA_et_B$patient[poolA_et_B$orig.ident %in% c("AML_OB5","AML_OB6")] <- "P3"
poolA_et_B$patient[poolA_et_B$orig.ident %in% c("AML_OB7","AML_OB8")] <- "P4"

# Assignation réponse
poolA_et_B$response[poolA_et_B$patient %in% c("P1","P4")] <- "refractory"
poolA_et_B$response[poolA_et_B$patient %in% c("P2","P3")] <- "responder"

# Assignation timepoint
poolA_et_B$timepoint[poolA_et_B$orig.ident %in% c("AML_OB1","AML_OB3","AML_OB5","AML_OB7")] <- "pre"
poolA_et_B$timepoint[poolA_et_B$orig.ident %in% c("AML_OB2","AML_OB4","AML_OB6","AML_OB8")] <- "post"

```

```{r}
# Vérifier
table(poolA_et_B$patient, poolA_et_B$timepoint)
table(poolA_et_B$response, poolA_et_B$timepoint)
```

# QC (quality control)

```{r}
#poolA
poolA[["percent.mt"]] <- PercentageFeatureSet(poolA, pattern = "^MT-")
VlnPlot(poolA, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"))
FeatureScatter(poolA, "nCount_RNA", "percent.mt")

#poolB
poolB[["percent.mt"]] <- PercentageFeatureSet(poolB, pattern = "^MT-")
VlnPlot(poolB, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"))
FeatureScatter(poolB, "nCount_RNA", "percent.mt")

#poolA et B
poolA_et_B[["percent.mt"]] <- PercentageFeatureSet(poolA_et_B, pattern = "^MT-")
VlnPlot(poolA_et_B, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"))
FeatureScatter(poolA_et_B, "nCount_RNA", "percent.mt")
```

```{r}
#poolA
poolA[["percent.ribo"]] <- PercentageFeatureSet(poolA, pattern = "^RPS|^RPL")

#pool B
poolB[["percent.ribo"]] <- PercentageFeatureSet(poolB, pattern = "^RPS|^RPL")

#poolA et B
poolA_et_B[["percent.ribo"]] <- PercentageFeatureSet(poolA_et_B, pattern = "^RPS|^RPL")
```

# filtering

```{r}
#poolA
poolA <- subset(
  poolA,
  subset = nFeature_RNA > 300 &
    nFeature_RNA < 2500 &
    percent.mt < 10
)
VlnPlot(poolA, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"))
FeatureScatter(poolA, "nCount_RNA", "percent.mt")

#poolB
poolB <- subset(
  poolB,
  subset = nFeature_RNA > 300 &
    nFeature_RNA < 2500 &
    percent.mt < 10
)
VlnPlot(poolB, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"))
FeatureScatter(poolB, "nCount_RNA", "percent.mt")

#poolA et B
poolA_et_B <- subset(
  poolA_et_B,
  subset = nFeature_RNA > 300 &
    nFeature_RNA < 2500 &
    percent.mt < 10
)

VlnPlot(poolA_et_B, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"))
FeatureScatter(poolA_et_B, "nCount_RNA", "percent.mt")
```

#normalisation “LogNormalize” that normalizes the feature expression measurements for each cell by the total expression, multiplies this by a scale factor (10,000 by default), and log-transforms the result.

```{r}
#poolA
poolA <- NormalizeData(poolA) 

#poolB
poolB <- NormalizeData(poolB) 

#poolA et B
poolA_et_B <- NormalizeData(poolA_et_B) 
```


# Identification of highly-variable genes

```{r}
#poolA
poolA <- FindVariableFeatures(poolA, nfeatures = 2000)

# Identify the top 10 Highly Variable Genes (HVGs) poolA
top10_poolA <- head(VariableFeatures(poolA), 10)
top10_plot_poolA <- VariableFeaturePlot(poolA)
LabelPoints(plot = top10_plot_poolA, points = top10_poolA, repel = TRUE)

#poolB
poolB <- FindVariableFeatures(poolB, nfeatures = 2000)

# Identify the top 10 Highly Variable Genes (HVGs) poolB
top10_poolB <- head(VariableFeatures(poolB), 10)
top10_plot_poolB <- VariableFeaturePlot(poolB)
LabelPoints(plot = top10_plot_poolB, points = top10_poolB, repel = TRUE)

#poolA et B
poolA_et_B <- FindVariableFeatures(poolA_et_B, nfeatures = 2000)

# Identify the top 10 Highly Variable Genes (HVGs) poolA et B
top10_poolA_et_B <- head(VariableFeatures(poolA_et_B), 10)
top10_plot_poolA_et_B <- VariableFeaturePlot(poolA_et_B)
LabelPoints(plot = top10_plot_poolA_et_B, points = top10_poolA_et_B, repel = TRUE)
```


# nbr gene/cell

```{r}
# Summary table of genes per cell by Sample
VlnPlot(poolA_et_B, features = "nFeature_RNA", group.by = "orig.ident")

# To get the average number per pool:
#tapply(poolA_et_B$nFeature_RNA, poolA_et_B$pool, mean) [cite: 12]
```

## nbr de cell/condition 
```{r}
poolA_et_B@meta.data %>%
  count(response) %>%
  ggplot(aes(x = response, y = n, fill = response)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  labs(
    x = "réponse au ttt",
    y = "Nombre de cellules"
  )
```

# scaling

```{r}
#poolA
poolA <- ScaleData(poolA)
#view(poolA@assays$RNA)

#poolB
poolB <- ScaleData(poolB)
#view(poolB@assays$RNA)

#poolA et B
poolA_et_B <- ScaleData(poolA_et_B)
#view(poolA_et_B@assays$RNA)

```

# PCA

```{r}
#poolA
poolA <- RunPCA(poolA, features = VariableFeatures(poolA))

#choisir le nbr de PC

ElbowPlot(poolA, ndims = 50)

# Visualize PCA
VizDimLoadings(poolA, dims = 1:2, reduction = "pca")
DimPlot(poolA, reduction = "pca")

#poolB
poolB <- RunPCA(poolB, features = VariableFeatures(poolB))

#choisir le nbr de PC

ElbowPlot(poolB, ndims = 50)

# Visualize PCA
VizDimLoadings(poolB, dims = 1:2, reduction = "pca")
DimPlot(poolB, reduction = "pca")

#poolA et B
poolA_et_B <- RunPCA(poolA_et_B, features = VariableFeatures(poolA_et_B))

#choisir le nbr de PC

ElbowPlot(poolA_et_B, ndims = 50)

# Visualize PCA
VizDimLoadings(poolA_et_B, dims = 1:2, reduction = "pca")
DimPlot(poolA_et_B, reduction = "pca")
```

# cell cycle

```{r}
library(scran)
library(BiocParallel)

cell_cyle <- cyclone(
  poolA_et_B, 
  pairs = pairs, 
  gene.names = rownames(poolA_et_B),
  BPPARAM = SerialParam()
)

head(cell_cyle)
head(cell_cyle$scores)
head(cell_cyle$phases)

# Add the phase predictions to your metadata
poolA_et_B$G1_score <- assignments$score$G1
poolA_et_B$S_score <- assignments$score$S
poolA_et_B$G2M_score <- assignments$score$G2M
poolA_et_B$phase <- assignments$phases

# Visualize: Are the Refractory cells "sleeping" (G1)?
VlnPlot(poolA_et_B, features = "G1_score", group.by = "response")
```

```{r}
poolA_et_B@meta.data %>%
  count(cell_cyle) %>%
  ggplot(aes(x = cell_cyle, y = n, fill = cell_cyle)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_bw() +
  labs(
    x = "cell cycle",
    y = "nbr de cell"
  )
```

```{r}
poolA_et_B@meta.data %>%
  count(patient ,cell_cycle) %>%
  ggplot(aes(x = patient, y = n, fill = cell_cyle)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_bw() +
  labs(
    x = "samples",
    y = "nbr de cell",
    fill = "cell_cycle"
  )
```

# Integration

```{r}
# (keeping layers separate)
poolA_et_B <- IntegrateLayers(
  object = poolA_et_B, 
  method = CCAIntegration, # Good for different donors
  orig.reduction = "pca", 
  new.reduction = "integrated.cca",
  verbose = FALSE
)
```

# clustering

```{r}
#poolA
poolA <-  FindNeighbors(poolA, dim = 1:30)
poolA <- FindClusters(poolA, resolution = 0.4)

#poolB
poolB <-  FindNeighbors(poolB, dim = 1:30)
poolB <- FindClusters(poolB, resolution = 0.4)

#poolA et B
poolA_et_B <-  FindNeighbors(poolA_et_B, dim = 1:30)
poolA_et_B <- FindClusters(poolA_et_B, resolution = 0.4)

```

# UMAP

```{r}
#poolA
poolA <- RunUMAP(poolA, dims = 1:30)
DimPlot(poolA, reduction = "umap", label = TRUE)

#poolB
poolB <- RunUMAP(poolB, dims = 1:30)
DimPlot(poolB, reduction = "umap", label = TRUE)

#poolA et B
poolA_et_B <- RunUMAP(poolA_et_B, dims = 1:30)
DimPlot(poolA_et_B, reduction = "umap", label = TRUE)
```


# batch effect check

```{r}

# Check if samples (OB1-OB8) are mixing or separated
#p1 <- DimPlot(poolA_et_B, reduction = "umap", group.by = "orig.ident") + 
      #ggtitle("UMAP by Sample ID")

# Check if there is a "Pool" effect (PoolA vs PoolB)
# Note: Since you merged poolA and poolB with add.cell.ids, 
# you might need to create a 'pool' column if it's not already in metadata.
# Safer way to assign Pool metadata
#poolA_et_B$pool <- ifelse(poolA_et_B$orig.ident %in% c("AML_OB1", "AML_OB2", "AML_OB3", "AML_OB4"), "PoolA", "PoolB")
#poolA_et_B$pool <- ifelse(grepl("PoolA", rownames(poolA_et_B@meta.data)), "PoolA", "PoolB")

#p2 <- DimPlot(poolA_et_B, reduction = "umap", group.by = "pool") + 
     # ggtitle("UMAP by Pool")

#p1 + p2
```

# join layer

```{r}
# --- poolA ---
# Join layers first
poolA <- JoinLayers(poolA)

# --- poolB ---
# Join layers first
poolB <- JoinLayers(poolB)

# --- poolA et B ---
# Join layers first
poolA_et_B <- JoinLayers(poolA_et_B)

```


# find marker

```{r}
#poolA
markers_poolA <-  FindAllMarkers(
  poolA,
  only.pos = TRUE,
  min.pct = 0.25,
  logfc.threshold = 0.25
)

#poolB
markers_poolB <-  FindAllMarkers(
  poolB,
  only.pos = TRUE,
  min.pct = 0.25,
  logfc.threshold = 0.25
)

#poolA et B
markers_poolA_et_B <-  FindAllMarkers(
  poolA_et_B,
  only.pos = TRUE,
  min.pct = 0.25,
  logfc.threshold = 0.25
)
```

```{r}
library(dplyr)

top10 <- markers_poolA_et_B %>%
  group_by(cluster) %>%
  slice_max(n = 10, order_by = avg_log2FC)

DoHeatmap(poolA_et_B, features = top10$gene)
```

```{r}
VlnPlot(poolA_et_B, features = c("TKT", "MAL", "GZMH", "ADGRL3", "TRDC", "BICDL1", "LINC02506", "LINC01819", "IFIT1", "XACT","TNFAIP6", "IGLC3", "LINC02446", "ENSG00000283458", "ENSG00000236304", "LINC01169", "PACSIN1"))
RidgePlot(poolA_et_B, features = c("TKT", "MAL", "GZMH", "ADGRL3", "TRDC", "BICDL1", "LINC02506", "LINC01819", "IFIT1", "XACT","TNFAIP6", "IGLC3", "LINC02446", "ENSG00000283458", "ENSG00000236304", "LINC01169", "PACSIN1"))
DotPlot(poolA_et_B, features = c("TKT", "MAL", "GZMH", "ADGRL3", "TRDC", "BICDL1", "LINC02506", "LINC01819", "IFIT1", "XACT","TNFAIP6", "IGLC3", "LINC02446", "ENSG00000283458", "ENSG00000236304", "LINC01169", "PACSIN1"))
```

```{r}
VlnPlot(poolA_et_B, features = top10$gene)
RidgePlot(poolA_et_B, features = top10$gene)
```

```{r}
unique(poolA_et_B$orig.ident)
```


# sex classification

```{r}
library(org.Hs.eg.db)
library(cellXY)

# 1. Extraire les counts bruts
# JoinLayers est crucial ici pour avoir une seule matrice
poolA_et_B <- JoinLayers(poolA_et_B)
counts_bruts <- GetAssayData(poolA_et_B, assay = "RNA", layer = "counts")

# 2. Conversion en matrice standard (Dense Matrix)
# C'est l'étape qui règle l'erreur "S4 to vector"
counts_matrix <- as.matrix(counts_bruts)

# 3. Vérification des noms de gènes
# cellXY a besoin de SYMBOLS (ex: XIST, pas ENSG000...)
# Si vos gènes sont déjà des Symbols, passez à l'étape 4.
# Sinon, utilisez ce bloc pour convertir :
# ann <- select(org.Hs.eg.db, keys=rownames(counts_matrix), 
#               columns=c("SYMBOL"), keytype="ENSEMBL")
# m <- match(rownames(counts_matrix), ann$ENSEMBL)
# rownames(counts_matrix) <- ann$SYMBOL[m]

# 4. Lancer la classification
# On donne la matrice 'counts_matrix' directement, pas l'objet Seurat
sex_preds <- classifySex(counts_matrix, genome = "Hs")

# 5. Vérifier et Intégrer
table(sex_preds$prediction)

# Ajouter au metadata Seurat
poolA_et_B$predicted_sex <- sex_preds$prediction

# 6. Le Boxplot de validation (comme dans l'exemple)
boxplot(counts_matrix["XIST", ] ~ poolA_et_B$predicted_sex, 
        main="Validation par XIST", ylab="Expression Log", col=c("pink", "lightblue"))
```


```{r}
library(ggplot2)

df_sex <- as.data.frame(table(poolA_et_B$patient, poolA_et_B$predicted_sex))
colnames(df_sex) <- c("Patient", "Sexe", "Nombre_de_Cellules")

ggplot(df_sex, aes(x = Patient, y = Nombre_de_Cellules, fill = Sexe)) +
  geom_bar(stat = "identity", position = "fill") + # "fill" pour avoir des pourcentages (0 à 100%)
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("Female" = "pink", "Male" = "lightblue")) +
  theme_minimal() +
  labs(title = "Répartition du Sexe par Patient", 
       y = "Proportion de cellules", 
       x = "Identifiant Patient")
```


```{r}
library(ggplot2)

# EXtraction de la colonne sex en vecteur simple
sex_vector <- as.character(poolA_et_B$predicted_sex)

# creation d'un count table et convertion en data frame
sex_counts <- as.data.frame(table(sex_vector))
colnames(sex_counts) <- c("Sexe", "Nombre_de_cellules")

ggplot(sex_counts, aes(x = Sexe, y = Nombre_de_cellules, fill = Sexe)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("Female" = "pink", "Male" = "lightblue")) +
  theme_minimal() +
  labs(title = "Répartition Globale du Sexe", 
       subtitle = "Nombre total de cellules identifiées",
       x = "Sexe Prédit", 
       y = "Nombre de cellules")
```

# filtration des gene sex 
```{r}
# Comparer Réfractaire vs Répondeur en ignorant l'effet du sexe
# On utilise l'argument 'latent.vars' pour corriger statistiquement le biais
clean_markers <- FindMarkers(
  poolA_et_B, 
  group.by = "response", 
  ident.1 = "refractory", 
  ident.2 = "responder",
  latent.vars = "predicted_sex", # CECI est la clé pour "effacer" le bruit du sexe
  test.use = "LR" # Nécessaire quand on utilise latent.vars
)

# Regardez le nouveau top 10
head(clean_markers, 10)

# Mise à jour du Heatmap
DoHeatmap(poolA_et_B, 
          features = clean_markers, 
          group.by = "response", 
          size = 4) + 
  scale_fill_gradientn(colors = c("blue", "white", "red")) +
  ggtitle("Signature de Résistance (Corrigée du Sexe)")
```

## évolution intra-patient

```{r}
DimPlot(poolA_et_B, group.by = "timepoint", split.by = "patient")
```

## proportions par cluster

```{r}
prop.table(table(
  Idents(poolA_et_B),
  poolA_et_B$timepoint,
  poolA_et_B$patient
), margin = 2)
```

## comparer responder vs refractory

```{r}
DimPlot(poolA_et_B, group.by = "response")
```

```{r}
library(patchwork) # to put plots side by side

p1 <- DimPlot(poolA_et_B, reduction = "umap", group.by = "response") + ggtitle("Response")
p2 <- DimPlot(poolA_et_B, reduction = "umap", group.by = "seurat_clusters", label = TRUE) + ggtitle("Clusters")

p1 + p2
```

## proportions de cell

```{r}
prop.table(table(
  Idents(poolA_et_B),
  poolA_et_B$response
), margin = 2)
```

# analyse differentiel #PRE vs POST (par patient)

```{r}
Idents(poolA_et_B) <- "timepoint"

markers_pre_post <- FindMarkers(
  poolA_et_B,
  ident.1 = "post",
  ident.2 = "pre",
  subset.ident = "P1"
)
```

```{r}
top10 <- markers_pre_post %>%
  rownames_to_column("gene") %>%
  slice_max(order_by = avg_log2FC, n = 10)

DoHeatmap(poolA_et_B, features = top10$gene) + NoLegend()
```

```{r}
VlnPlot(poolA_et_B, features = top10$gene[1:10], group.by = "timepoint")
```

```{r}
FeaturePlot(poolA_et_B, features = top10$gene[1:10])
```

```{r}
DotPlot(poolA_et_B, features = top10$gene, group.by = "timepoint") + RotatedAxis()
```

```{r}
library(ggplot2)

markers_pre_post$gene <- rownames(markers_pre_post)

ggplot(markers_pre_post, aes(x = avg_log2FC, y = -log10(p_val_adj))) +
  geom_point() +
  theme_minimal() +
  geom_vline(xintercept = c(-0.5, 0.5), linetype = "dashed") +
  geom_hline(yintercept = 2) +
  ggtitle("Volcano plot PRE vs POST")
```

```{r}
# Top genes upregulated POST
top_post <- markers_pre_post %>%
  rownames_to_column("gene") %>%
  filter(avg_log2FC > 0) %>%
  slice_max(n = 20, order_by = avg_log2FC)

# Ajouter score 
#calculates an average expression "score" for a whole list of genes
poolA_et_B <- AddModuleScore(
  poolA_et_B,
  features = list(top_post$gene),
  name = "POST_signature"
)

FeaturePlot(poolA_et_B, features = "POST_signature1")
VlnPlot(poolA_et_B, features = "POST_signature1", group.by = "timepoint")
```

## responder vs refractory

```{r}
Idents(poolA_et_B) <- "response"

markers_resistance <- FindMarkers(
  poolA_et_B,
  ident.1 = "refractory",
  ident.2 = "responder"
)
```

```{r}
top_resistance <- markers_resistance %>%
  rownames_to_column("gene") %>%
  filter(avg_log2FC > 0) %>%
  slice_max(n = 20, order_by = avg_log2FC)

poolA_et_B <- AddModuleScore(
  poolA_et_B,
  features = list(top_resistance$gene),
  name = "Resistance_signature1"
)

FeaturePlot(poolA_et_B, features = "Resistance_signature1")
VlnPlot(poolA_et_B, features = "Resistance_signature1", group.by = "response")
```

## par cluster

```{r}
Idents(poolA_et_B) <- "seurat_clusters"

markers_cluster <- FindMarkers(
  subset(poolA_et_B, idents = 3),
  group.by = "response",
  ident.1 = "refractory",
  ident.2 = "responder"
)
```

## tableau de comptage

```{r}
table(poolA_et_B$patient, poolA_et_B$timepoint)
table(poolA_et_B$response, poolA_et_B$timepoint)
```

# save

```{r}
#poolA
saveRDS(poolA, file = "020_010_042_025_qc_drop_adt_integ_clean.rds")

#poolB
saveRDS(poolB, file = "008_034_095_017_qc_drop_adt_integ_clean.rds")

#poolA et B
saveRDS(poolA_et_B, file = "020_010_042_025_008_034_095_017_qc_drop_adt_integ_clean.rds")

```

# annotation

```{r}
library(SingleR)
library(caret)
library(patchwork)
library("ggplot2")
library('viridis')
library('pheatmap')

ref <- readRDS("O:/Equipes_PF_Services/EQ_Duprez/EQ_Duprez/Current Lab Members/Chellal Alyssa/Ref_annotation/AML_downsampled_ref_fixed2_with_umap.rds")

#poolA
ref_mat_poolA  <- GetAssayData(ref, layer = "data")
test_mat_poolA <- GetAssayData(poolA, layer = "data")

head(test_mat_poolA)
head(ref_mat_poolA)

common_genes_poolA <- intersect(rownames(ref_mat_poolA), rownames(test_mat_poolA))
length(common_genes_poolA)

pred_poolA <- SingleR(test = test_mat_poolA, ref = ref_mat_poolA, labels = ref$Cell.Type, de.method="wilcox", de.n=20)
colnames(pred_poolA)

head(sort(table(pred_poolA$labels), decreasing=TRUE))
table(pred_poolA$labels)
head(pred_poolA$labels)

pred_poolA$scores[1:10,]
plotScoreHeatmap(pred_poolA)

pred2_poolA <- SingleR(test = test_mat_poolA, ref = ref_mat_poolA, labels = ref$HSPC.Cell.Type, de.method="wilcox", de.n=20)

colnames(pred2_poolA)
table(pred2_poolA$labels)
head(sort(table(pred2_poolA$labels), decreasing=TRUE))

pred2_poolA$scores[1:10,]
plotScoreHeatmap(pred2_poolA)

poolA$SingleR_main_poolA  <- pred_poolA$labels
poolA$SingleR_fine_poolA  <- pred2_poolA$labels
poolA@meta.data[["SingleR_main_poolA"]]
poolA@meta.data[["SingleR_fine_poolA"]]

options(ggrepel.max.overlaps = Inf)

DimPlot(
  poolA,
  reduction = "umap",
  group.by = "SingleR_main_poolA",
  label = TRUE,
  repel = TRUE
) + theme_bw()

DimPlot(
  poolA,
  reduction = "umap",
  group.by = "SingleR_fine_poolA",
  label = TRUE,
  repel = TRUE
) + theme_bw()


#poolB
ref_mat_poolB  <- GetAssayData(ref, layer = "data")
test_mat_poolB <- GetAssayData(poolB, layer = "data")

head(test_mat_poolB)
head(ref_mat_poolB)

common_genes_poolB <- intersect(rownames(ref_mat_poolB), rownames(test_mat_poolB))
length(common_genes_poolB)

pred_poolB <- SingleR(test = test_mat_poolB, ref = ref_mat_poolB, labels = ref$Cell.Type, de.method="wilcox", de.n=20)
colnames(pred_poolB)

head(sort(table(pred_poolB$labels), decreasing=TRUE))
table(pred_poolB$labels)
head(pred_poolB$labels)

pred_poolB$scores[1:10,]
plotScoreHeatmap(pred_poolB)

pred2_poolB <- SingleR(test = test_mat_poolB, ref = ref_mat_poolB, labels = ref$HSPC.Cell.Type, de.method="wilcox", de.n=20)

colnames(pred2_poolB)
table(pred2_poolB$labels)
head(sort(table(pred2_poolB$labels), decreasing=TRUE))

pred2_poolB$scores[1:10,]
plotScoreHeatmap(pred2_poolB)

poolB$SingleR_main_poolB  <- pred_poolB$labels
poolB$SingleR_fine_poolB  <- pred2_poolB$labels
poolB@meta.data[["SingleR_main_poolB"]]
poolB@meta.data[["SingleR_fine_poolB"]]

options(ggrepel.max.overlaps = Inf)

DimPlot(
  poolB,
  reduction = "umap",
  group.by = "SingleR_main_poolB",
  label = TRUE,
  repel = TRUE
) + theme_bw()

DimPlot(
  poolB,
  reduction = "umap",
  group.by = "SingleR_fine_poolB",
  label = TRUE,
  repel = TRUE
) + theme_bw()


#poolA et B
ref_mat_poolA_et_B  <- GetAssayData(ref, layer = "data")
test_mat_poolA_et_B <- GetAssayData(poolA_et_B, layer = "data")

head(test_mat_poolA_et_B)
head(ref_mat_poolA_et_B)

common_genes_poolA_et_B <- intersect(rownames(ref_mat_poolA_et_B), rownames(test_mat_poolA_et_B))
length(common_genes_poolA_et_B)

pred_poolA_et_B <- SingleR(test = test_mat_poolA_et_B, ref = ref_mat_poolA_et_B, labels = ref$Cell.Type, de.method="wilcox", de.n=20)
colnames(pred_poolA_et_B)

head(sort(table(pred_poolA_et_B$labels), decreasing=TRUE))
table(pred_poolA_et_B$labels)
head(pred_poolA_et_B$labels)

pred_poolA_et_B$scores[1:10,]
plotScoreHeatmap(pred_poolA_et_B)

pred2_poolA_et_B <- SingleR(test = test_mat_poolA_et_B, ref = ref_mat_poolA_et_B, labels = ref$HSPC.Cell.Type, de.method="wilcox", de.n=20)

colnames(pred2_poolA_et_B)
table(pred2_poolA_et_B$labels)
head(sort(table(pred2_poolA_et_B$labels), decreasing=TRUE))

pred2_poolA_et_B$scores[1:10,]
plotScoreHeatmap(pred2_poolA_et_B)

poolA_et_B$SingleR_main_poolA_et_B  <- pred_poolA_et_B$labels
poolA_et_B$SingleR_fine_poolA_et_B  <- pred2_poolA_et_B$labels
poolA_et_B@meta.data[["SingleR_main_poolA_et_B"]]
poolA_et_B@meta.data[["SingleR_fine_poolA_et_B"]]

options(ggrepel.max.overlaps = Inf)

DimPlot(
  poolA_et_B,
  reduction = "umap",
  group.by = "SingleR_main_poolA_et_B",
  label = TRUE,
  repel = TRUE
) + theme_bw()

DimPlot(
  poolA_et_B,
  reduction = "umap",
  group.by = "SingleR_fine_poolA_et_B",
  label = TRUE,
  repel = TRUE
) + theme_bw()
```

```{r}
table(poolA_et_B$patient, poolA_et_B$timepoint)
table(poolA_et_B$response, poolA_et_B$timepoint)
```

```{r}
library(patchwork) # to put plots side by side

p1 <- DimPlot(poolA_et_B, reduction = "umap", group.by = "response") + ggtitle("Response")
p2 <- DimPlot(poolA_et_B, reduction = "umap", group.by = "SingleR_main_poolA_et_B", label = TRUE) + ggtitle("Clusters")

p1 + p2
```

```{r}
library(patchwork) 

p1 <- DimPlot(poolA_et_B, reduction = "umap", group.by = "response") + ggtitle("Response")
p2 <- DimPlot(poolA_et_B, reduction = "umap", group.by = "SingleR_fine_poolA_et_B", label = TRUE) + ggtitle("Clusters")

p1 + p2
```

```{r}
# Reset Idents to clusters to find the number
Idents(poolA_et_B) <- "seurat_clusters"

# Calculate the percentage of Refractory vs Responder per cluster
comp_table <- table(poolA_et_B$seurat_clusters, poolA_et_B$response)
prop_table <- prop.table(comp_table, margin = 1) * 100

# Sort to see the most "Refractory" clusters at the top
print(prop_table[order(prop_table[, "refractory"], decreasing = TRUE), ])
```

```{r}
# Créer une table croisée Cluster / Patient
patient_cluster_table <- table(poolA_et_B$seurat_clusters, poolA_et_B$patient)

# Convertir en proportions par patient (pour normaliser si un patient a plus de cellules)
patient_prop <- prop.table(patient_cluster_table, margin = 2) * 100

# Afficher les clusters pour vos patients réfractaires (P1 et P4)
print(patient_prop[, c("P1", "P4")])
```

```{r}
Idents(poolA_et_B) <- "response"

#cluster 6
DotPlot(poolA_et_B, features = c("LINC02506", "TNFRSF12A", "HLA-DRB5", "SPINK2", "EMP1", "ID1", "CXCL8", "ABO", "LMNA", "FHL2"))

#cluster 0
DotPlot(poolA_et_B, features = c("TKT", "SERPINB1", "HSPD1", "NPM1", "PDIA6", "TALAM1", "ATP5F1C", "TUBB", "OLA1", "CCT6A"))

#cluster 14
DotPlot(poolA_et_B, features = c("ENSG00000236304", "ENSG00000289621", "SPX", "PF4V1", "PPBP", "LY6G6F", "GP9", "PF4", "TREML1", "CLEC1B"))
```

```{r}
Idents(poolA_et_B) <- "timepoint"

#cluster 6
DotPlot(poolA_et_B, features = c("LINC02506", "TNFRSF12A", "HLA-DRB5", "SPINK2", "EMP1", "ID1", "CXCL8", "ABO", "LMNA", "FHL2"))

#cluster 0
DotPlot(poolA_et_B, features = c("TKT", "SERPINB1", "HSPD1", "NPM1", "PDIA6", "TALAM1", "ATP5F1C", "TUBB", "OLA1", "CCT6A"))

#cluster 14
DotPlot(poolA_et_B, features = c("ENSG00000236304", "ENSG00000289621", "SPX", "PF4V1", "PPBP", "LY6G6F", "GP9", "PF4", "TREML1", "CLEC1B"))
```

```{r}
Idents(poolA_et_B) <- "patient"

#cluster 6
DotPlot(poolA_et_B, features = c("LINC02506", "TNFRSF12A", "HLA-DRB5", "SPINK2", "EMP1", "ID1", "CXCL8", "ABO", "LMNA", "FHL2"))

#cluster 0
DotPlot(poolA_et_B, features = c("TKT", "SERPINB1", "HSPD1", "NPM1", "PDIA6", "TALAM1", "ATP5F1C", "TUBB", "OLA1", "CCT6A"))

#cluster 14
DotPlot(poolA_et_B, features = c("ENSG00000236304", "ENSG00000289621", "SPX", "PF4V1", "PPBP", "LY6G6F", "GP9", "PF4", "TREML1", "CLEC1B"))
```

```{r}
# Cross-tabulate Cluster number with SingleR Fine labels
table(poolA_et_B$seurat_clusters, poolA_et_B$SingleR_fine_poolA_et_B)
``` 

```{r}
# Create a contingency table of Clusters vs SingleR Fine Annotation
annotation_map <- table(poolA_et_B$seurat_clusters, poolA_et_B$SingleR_fine_poolA_et_B)

# Convert to proportions to see the "dominant" cell type for each cluster
annotation_prop <- prop.table(annotation_map, margin = 1) * 100

# View the map
print(annotation_prop)

library(ggplot2)

# This plot shows the distribution of cell types within each cluster
ggplot(poolA_et_B@meta.data, aes(x = seurat_clusters, fill = SingleR_fine_poolA_et_B)) +
  geom_bar(position = "fill") +
  theme_minimal() +
  RotatedAxis() +
  labs(y = "Proportion", title = "Biological Identity of Numerical Clusters")
```

```{r}
library(ggplot2)

ggplot(poolA_et_B@meta.data, aes(x = response, fill = SingleR_fine_poolA_et_B)) +
  geom_bar(position = "fill") +
  theme_minimal() +
  labs(y = "Proportion of Cells", title = "Cell Type Composition: Responder vs Refractory")
```

```{r}
# Install if needed: install.packages("openxlsx")
library(openxlsx)

# 1. Set Identity to Response
Idents(poolA_et_B) <- "response"

# 2. Run the Differential Expression (DE)
# We use logfc.threshold = 0.25 (approx 1.2x fold change) 
# and min.pct = 0.1 (gene must be in at least 10% of cells)
resistance_markers <- FindMarkers(
  poolA_et_B,
  ident.1 = "refractory",
  ident.2 = "responder",
  only.pos = TRUE 
)

# 3. Filter for Significance and top 50
# We want highly significant (p < 0.05) and strongest fold change
top50_resistance <- resistance_markers %>%
  rownames_to_column("gene") %>%
  filter(p_val_adj < 0.05) %>%
  slice_max(n = 50, order_by = avg_log2FC)

# 4. Save to Excel
write.xlsx(top50_resistance, file = "Top50_Refractory_Markers.xlsx")
```



Transcription Factor Inference (SCENIC or DoRothEA)
Genes don't just turn on by themselves; they are controlled by Transcription Factors (TFs) like SPI1 (PU.1) or RUNX1. Analyzing these tells you the "Master Regulator" of the refractory state.
```{r}
# Quick check: Look for TFs in your marker list
tfs_to_check <- c("SPI1", "RUNX1", "CEBPA", "MYC", "GATA2")
DotPlot(poolA_et_B, features = tfs_to_check, group.by = "response")
```

```{r}
Idents(poolA_et_B) <- "response"

# find refractory markers for every cluster compared to all remaining cells, report only the positive ones
aml.markers <- FindAllMarkers(poolA_et_B, only.pos = TRUE)
aml.markers %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1)

DotPlot(poolA_et_B, features = c("SLC35F1", "RPS24", "RPS26", "RPS4Y1", "DDX3Y", "UTY"))

top20_repsonse_aml <- aml.markers %>%
  group_by(cluster) %>%
  slice_max(n = 20, order_by = avg_log2FC)

DoHeatmap(poolA_et_B, features = top20_repsonse_aml$gene)

# 1. Get the list of genes you want to plot
genes_to_plot <- top20_repsonse_aml$gene

# 2. Scale ONLY these genes (this is fast)
# This adds them to the scale.data layer so DoHeatmap can find them
poolA_et_B <- ScaleData(poolA_et_B, features = genes_to_plot)

# 3. Now run the heatmap again
DoHeatmap(poolA_et_B, features = genes_to_plot, group.by = "response") +
  scale_fill_gradientn(colors = c("blue", "white", "red"))
```

# save final

```{r}
#poolA
#saveRDS(poolA, file = "020_010_042_025_qc_drop_adt_integ_clean_annot.rds")

#poolB
#saveRDS(poolB, file = "008_034_095_017_qc_drop_adt_integ_clean_annot.rds")

#poolA et B
#saveRDS(poolA_et_B, file = "020_010_042_025_008_034_095_017_qc_drop_adt_integ_clean_annot.rds")

```
