---
title: "scRNA seq AML 2ndaire"
format: html
editor: visual
---

# Setting up environment:

```{r}
# Clean environment
rm(list = ls(all.names = TRUE)) # will clear all objects including hidden objects
gc() # free up memory and report the memory usage
options(max.print = .Machine$integer.max, scipen = 999, stringsAsFactors = F, dplyr.summarise.inform = F) # avoid truncated output in R console and scientific notation
```

```{r}
# Loading relevant libraries 
library(tidyverse) # includes ggplot2, for data visualisation. dplyr, for data manipulation.
library(Seurat)
```

# import data

## poolA

```{r}
#020_010_D refractaire
inputMatrixDirOB1 <- "O:/Equipes_PF_Services/EQ_Duprez/EQ_Duprez/Current Lab Members/Chellal Alyssa/cellranger_output_poolA/poolA_output/outs/per_sample_outs/020_010_D/count/sample_filtered_feature_bc_matrix"

if(!dir.exists(inputMatrixDirOB1)){
  stop(paste("ERREUR : Le dossier n'existe pas :", inputMatrixDirOB1))
}

cat("Reading 10X matrix...\n")
OB1 <- Read10X(
  data.dir = inputMatrixDirOB1,
  gene.column = 2,        # utiliser 2 si features.tsv a gene symbol
  unique.features = TRUE,
  strip.suffix = FALSE
)

str(OB1) # Check the multiple modalities (list of matrixes) - we're interested in Gene expression
cts_OB1 <- OB1$`Gene Expression`
adt_OB1 <- OB1$`Antibody Capture`

#020_010_F
inputMatrixDirOB2 <- "O:/Equipes_PF_Services/EQ_Duprez/EQ_Duprez/Current Lab Members/Chellal Alyssa/cellranger_output_poolA/poolA_output/outs/per_sample_outs/020_010_F/count/sample_filtered_feature_bc_matrix"

if(!dir.exists(inputMatrixDirOB2)){
  stop(paste("ERREUR : Le dossier n'existe pas :", inputMatrixDirOB2))
}

cat("Reading 10X matrix...\n")
OB2 <- Read10X(
  data.dir = inputMatrixDirOB2,
  gene.column = 2,        # utiliser 2 si features.tsv a gene symbol
  unique.features = TRUE,
  strip.suffix = FALSE
)

str(OB2) # Check the multiple modalities (list of matrixes) - we're interested in Gene expression
cts_OB2 <- OB2$`Gene Expression`
adt_OB2 <- OB2$`Antibody Capture`

#042_025_D repondeur
inputMatrixDirOB3 <- "O:/Equipes_PF_Services/EQ_Duprez/EQ_Duprez/Current Lab Members/Chellal Alyssa/cellranger_output_poolA/poolA_output/outs/per_sample_outs/042_025_D/count/sample_filtered_feature_bc_matrix"

if(!dir.exists(inputMatrixDirOB3)){
  stop(paste("ERREUR : Le dossier n'existe pas :", inputMatrixDirOB3))
}

cat("Reading 10X matrix...\n")
OB3 <- Read10X(
  data.dir = inputMatrixDirOB3,
  gene.column = 2,        # utiliser 2 si features.tsv a gene symbol
  unique.features = TRUE,
  strip.suffix = FALSE
)

str(OB3) # Check the multiple modalities (list of matrixes) - we're interested in Gene expression
cts_OB3 <- OB3$`Gene Expression`
adt_OB3 <- OB3$`Antibody Capture`

#042_025_F
inputMatrixDirOB4 <- "O:/Equipes_PF_Services/EQ_Duprez/EQ_Duprez/Current Lab Members/Chellal Alyssa/cellranger_output_poolA/poolA_output/outs/per_sample_outs/042_025_F/count/sample_filtered_feature_bc_matrix"

if(!dir.exists(inputMatrixDirOB4)){
  stop(paste("ERREUR : Le dossier n'existe pas :", inputMatrixDirOB4))
}

cat("Reading 10X matrix...\n")
OB4 <- Read10X(
  data.dir = inputMatrixDirOB4,
  gene.column = 2,        # utiliser 2 si features.tsv a gene symbol
  unique.features = TRUE,
  strip.suffix = FALSE
)

str(OB4) # Check the multiple modalities (list of matrixes) - we're interested in Gene expression
cts_OB4 <- OB4$`Gene Expression`
adt_OB4 <- OB4$`Antibody Capture`

```

## poolB

```{r}
#008_034_D repondant

inputMatrixDirOB5 <- "O:/Equipes_PF_Services/EQ_Duprez/EQ_Duprez/Current Lab Members/Chellal Alyssa/cellranger_output_poolB/poolB_output/outs/per_sample_outs/008_034_D/count/sample_filtered_feature_bc_matrix"

if(!dir.exists(inputMatrixDirOB5)){
  stop(paste("ERREUR : Le dossier n'existe pas :", inputMatrixDirOB5))
}

cat("Reading 10X matrix...\n")
OB5 <- Read10X(
  data.dir = inputMatrixDirOB5,
  gene.column = 2,        # utiliser 2 si features.tsv a gene symbol
  unique.features = TRUE,
  strip.suffix = FALSE
)

str(OB5) # Check the multiple modalities (list of matrixes) - we're interested in Gene expression
cts_OB5 <- OB5$`Gene Expression`
adt_OB5 <- OB5$`Antibody Capture`

#008_034_F
inputMatrixDirOB6 <- "O:/Equipes_PF_Services/EQ_Duprez/EQ_Duprez/Current Lab Members/Chellal Alyssa/cellranger_output_poolB/poolB_output/outs/per_sample_outs/008_034_F/count/sample_filtered_feature_bc_matrix"

if(!dir.exists(inputMatrixDirOB6)){
  stop(paste("ERREUR : Le dossier n'existe pas :", inputMatrixDirOB6))
}

cat("Reading 10X matrix...\n")
OB6 <- Read10X(
  data.dir = inputMatrixDirOB6,
  gene.column = 2,        # utiliser 2 si features.tsv a gene symbol
  unique.features = TRUE,
  strip.suffix = FALSE
)

str(OB6) # Check the multiple modalities (list of matrixes) - we're interested in Gene expression
cts_OB6 <- OB6$`Gene Expression`
adt_OB6 <- OB6$`Antibody Capture`

#095_017_D refractaire
inputMatrixDirOB7 <- "O:/Equipes_PF_Services/EQ_Duprez/EQ_Duprez/Current Lab Members/Chellal Alyssa/cellranger_output_poolB/poolB_output/outs/per_sample_outs/095_017_D/count/sample_filtered_feature_bc_matrix"

if(!dir.exists(inputMatrixDirOB7)){
  stop(paste("ERREUR : Le dossier n'existe pas :", inputMatrixDirOB7))
}

cat("Reading 10X matrix...\n")
OB7 <- Read10X(
  data.dir = inputMatrixDirOB7,
  gene.column = 2,        # utiliser 2 si features.tsv a gene symbol
  unique.features = TRUE,
  strip.suffix = FALSE
)

str(OB7) # Check the multiple modalities (list of matrixes) - we're interested in Gene expression
cts_OB7 <- OB7$`Gene Expression`
adt_OB7 <- OB7$`Antibody Capture`

#095_017_F
inputMatrixDirOB8 <- "O:/Equipes_PF_Services/EQ_Duprez/EQ_Duprez/Current Lab Members/Chellal Alyssa/cellranger_output_poolB/poolB_output/outs/per_sample_outs/095_017_F/count/sample_filtered_feature_bc_matrix"

if(!dir.exists(inputMatrixDirOB8)){
  stop(paste("ERREUR : Le dossier n'existe pas :", inputMatrixDirOB8))
}

cat("Reading 10X matrix...\n")
OB8 <- Read10X(
  data.dir = inputMatrixDirOB8,
  gene.column = 2,        # utiliser 2 si features.tsv a gene symbol
  unique.features = TRUE,
  strip.suffix = FALSE
)

str(OB8) # Check the multiple modalities (list of matrixes) - we're interested in Gene expression
cts_OB8 <- OB8$`Gene Expression`
adt_OB8 <- OB8$`Antibody Capture`

```

# create seurat object

## poolA

```{r}
OB1 <- CreateSeuratObject(
  counts = cts_OB1,
  project = "AML_OB1",
  min.cells = 3,
  min.features = 200
)

# 1. Identify the cells that exist in BOTH the Seurat object and the ADT matrix
common_cells <- intersect(Cells(OB1), colnames(adt_OB1))

# 2. Subset the Seurat object to only include these common cells
# (This ensures the object doesn't have cells that are missing ADT data)
OB1 <- subset(OB1, cells = common_cells)

# 3. Subset and REORDER the ADT matrix to match the Seurat object exactly
adt_filtered <- adt_OB1[, Cells(OB1)]

# 4. Add the assay
OB1[["ADT"]] <- CreateAssay5Object(counts = adt_filtered)

saveRDS(OB1, file = "020_010_D_filtered_with_adt.rds")

head(rownames(OB1))

OB2 <- CreateSeuratObject(
  counts = cts_OB2,
  project = "AML_OB2",
  min.cells = 3,
  min.features = 200
)

# 1. Identify the cells that exist in BOTH the Seurat object and the ADT matrix
common_cells <- intersect(Cells(OB2), colnames(adt_OB2))

# 2. Subset the Seurat object to only include these common cells
# (This ensures the object doesn't have cells that are missing ADT data)
OB2 <- subset(OB2, cells = common_cells)

# 3. Subset and REORDER the ADT matrix to match the Seurat object exactly
adt_filtered <- adt_OB2[, Cells(OB2)]

# 4. Add the assay
OB2[["ADT"]] <- CreateAssay5Object(counts = adt_filtered)

saveRDS(OB2, file = "020_010_F_filtered_with_adt.rds")

head(rownames(OB2))

OB3 <- CreateSeuratObject(
  counts = cts_OB3,
  project = "AML_OB3",
  min.cells = 3,
  min.features = 200
)

# 1. Identify the cells that exist in BOTH the Seurat object and the ADT matrix
common_cells <- intersect(Cells(OB3), colnames(adt_OB3))

# 2. Subset the Seurat object to only include these common cells
# (This ensures the object doesn't have cells that are missing ADT data)
OB3 <- subset(OB3, cells = common_cells)

# 3. Subset and REORDER the ADT matrix to match the Seurat object exactly
adt_filtered <- adt_OB3[, Cells(OB3)]

# 4. Add the assay
OB3[["ADT"]] <- CreateAssay5Object(counts = adt_filtered)

saveRDS(OB3, file = "042_025_D_filtered_with_adt.rds")

head(rownames(OB3))

OB4 <- CreateSeuratObject(
  counts = cts_OB4,
  project = "AML_OB4",
  min.cells = 3,
  min.features = 200
)

# 1. Identify the cells that exist in BOTH the Seurat object and the ADT matrix
common_cells <- intersect(Cells(OB4), colnames(adt_OB4))

# 2. Subset the Seurat object to only include these common cells
# (This ensures the object doesn't have cells that are missing ADT data)
OB4 <- subset(OB4, cells = common_cells)

# 3. Subset and REORDER the ADT matrix to match the Seurat object exactly
adt_filtered <- adt_OB4[, Cells(OB4)]

# 4. Add the assay
OB4[["ADT"]] <- CreateAssay5Object(counts = adt_filtered)

saveRDS(OB4, file = "042_025_F_filtered_with_adt.rds")

head(rownames(OB4))

```

## poolB

```{r}
OB5 <- CreateSeuratObject(
  counts = cts_OB5,
  project = "AML_OB5",
  min.cells = 3,
  min.features = 200
)

# 1. Identify the cells that exist in BOTH the Seurat object and the ADT matrix
common_cells <- intersect(Cells(OB5), colnames(adt_OB5))

# 2. Subset the Seurat object to only include these common cells
# (This ensures the object doesn't have cells that are missing ADT data)
OB5 <- subset(OB5, cells = common_cells)

# 3. Subset and REORDER the ADT matrix to match the Seurat object exactly
adt_filtered <- adt_OB5[, Cells(OB5)]

# 4. Add the assay
OB5[["ADT"]] <- CreateAssay5Object(counts = adt_filtered)

saveRDS(OB5, file = "008_034_D_filtered_with_adt.rds")

head(rownames(OB5))

OB6 <- CreateSeuratObject(
  counts = cts_OB6,
  project = "AML_OB6",
  min.cells = 3,
  min.features = 200
)

# 1. Identify the cells that exist in BOTH the Seurat object and the ADT matrix
common_cells <- intersect(Cells(OB6), colnames(adt_OB6))

# 2. Subset the Seurat object to only include these common cells
# (This ensures the object doesn't have cells that are missing ADT data)
OB6 <- subset(OB6, cells = common_cells)

# 3. Subset and REORDER the ADT matrix to match the Seurat object exactly
adt_filtered <- adt_OB6[, Cells(OB6)]

# 4. Add the assay
OB6[["ADT"]] <- CreateAssay5Object(counts = adt_filtered)

saveRDS(OB6, file = "008_034_F_filtered_with_adt.rds")

head(rownames(OB6))

OB7 <- CreateSeuratObject(
  counts = cts_OB7,
  project = "AML_OB7",
  min.cells = 3,
  min.features = 200
)

# 1. Identify the cells that exist in BOTH the Seurat object and the ADT matrix
common_cells <- intersect(Cells(OB7), colnames(adt_OB7))

# 2. Subset the Seurat object to only include these common cells
# (This ensures the object doesn't have cells that are missing ADT data)
OB7 <- subset(OB7, cells = common_cells)

# 3. Subset and REORDER the ADT matrix to match the Seurat object exactly
adt_filtered <- adt_OB7[, Cells(OB7)]

# 4. Add the assay
OB7[["ADT"]] <- CreateAssay5Object(counts = adt_filtered)

saveRDS(OB7, file = "095_017_D_filtered_with_adt.rds")

head(rownames(OB7))

OB8 <- CreateSeuratObject(
  counts = cts_OB8,
  project = "AML_OB8",
  min.cells = 3,
  min.features = 200
)

# 1. Identify the cells that exist in BOTH the Seurat object and the ADT matrix
common_cells <- intersect(Cells(OB8), colnames(adt_OB8))

# 2. Subset the Seurat object to only include these common cells
# (This ensures the object doesn't have cells that are missing ADT data)
OB8 <- subset(OB8, cells = common_cells)

# 3. Subset and REORDER the ADT matrix to match the Seurat object exactly
adt_filtered <- adt_OB8[, Cells(OB8)]

# 4. Add the assay
OB8[["ADT"]] <- CreateAssay5Object(counts = adt_filtered)

saveRDS(OB8, file = "095_017_F_filtered_with_adt.rds")

head(rownames(OB8))

```

# merge RDS

## merge poolA

```{r}
# Merge OB1 (x) with the others (y)
poolA <- merge(
  x = OB1, 
  y = c(OB2, OB3, OB4), 
  add.cell.ids = c("OB1", "OB2", "OB3", "OB4"), 
  project = "PoolA"
)

# Check the new dimensions
# Should be ~38k genes and the sum of all cells
print(dim(poolA))

saveRDS(poolA, file = "020_010_042_025_filtered_with_adt.rds")

```

## merge poolB

```{r}
# Merge OB1 (x) with the others (y)
poolB <- merge(
  x = OB5, 
  y = c(OB6, OB7, OB8), 
  add.cell.ids = c("OB5", "OB6", "OB7", "OB8"), 
  project = "PoolB"
)

# Check the new dimensions
# Should be ~38k genes and the sum of all cells
print(dim(poolB))

saveRDS(poolB, file = "008_034_095_017_filtered_with_adt.rds")

```

## merge poolA + poolB

```{r}
# Merge poolA and poolB
# 'add.cell.ids' is critical here to avoid duplicate cell names across pools
poolA_et_B <- merge(
  x = poolA, 
  y = poolB, 
  add.cell.ids = c("PoolA", "PoolB"), 
  project = "Full_AML_Cohort"
)

# 1. Check the final cell count
ncol(poolA_et_B)
print(dim(poolA_et_B))
saveRDS(poolA_et_B, file = "020_010_042_025_008_034_095_017_filtered_with_adt.rds")

```

# Ajouter les metadata biologiques

```{r}
poolA_et_B$patient <- NA
poolA_et_B$response <- NA
poolA_et_B$timepoint <- NA
```

```{r}
# Assignation patient
poolA_et_B$patient[poolA_et_B$orig.ident %in% c("AML_OB1","AML_OB2")] <- "P1"
poolA_et_B$patient[poolA_et_B$orig.ident %in% c("AML_OB3","AML_OB4")] <- "P2"
poolA_et_B$patient[poolA_et_B$orig.ident %in% c("AML_OB5","AML_OB6")] <- "P3"
poolA_et_B$patient[poolA_et_B$orig.ident %in% c("AML_OB7","AML_OB8")] <- "P4"

# Assignation réponse
poolA_et_B$response[poolA_et_B$patient %in% c("P1","P4")] <- "refractory"
poolA_et_B$response[poolA_et_B$patient %in% c("P2","P3")] <- "responder"

# Assignation timepoint
poolA_et_B$timepoint[poolA_et_B$orig.ident %in% c("AML_OB1","AML_OB3","AML_OB5","AML_OB7")] <- "pre"
poolA_et_B$timepoint[poolA_et_B$orig.ident %in% c("AML_OB2","AML_OB4","AML_OB6","AML_OB8")] <- "post"

```

```{r}
# Vérifier
table(poolA_et_B$patient, poolA_et_B$timepoint)
table(poolA_et_B$response, poolA_et_B$timepoint)
```

# change l'ordre

```{r}
# Transformer 'patient' en facteur avec l'ordre souhaité
poolA_et_B$patient <- factor(poolA_et_B$patient, levels = c("P1", "P4", "P2", "P3"))
table(poolA_et_B$patient)

# Transformer 'timepoint' en facteur avec l'ordre souhaité
poolA_et_B$timepoint <- factor(poolA_et_B$timepoint, levels = c("pre", "post"))
table(poolA_et_B$timepoint)
```

# QC (quality control)

## mito

```{r}
#poolA
poolA[["percent.mt"]] <- PercentageFeatureSet(poolA, pattern = "^MT-")
VlnPlot(poolA, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"))
FeatureScatter(poolA, "nCount_RNA", "percent.mt")

#poolB
poolB[["percent.mt"]] <- PercentageFeatureSet(poolB, pattern = "^MT-")
VlnPlot(poolB, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"))
FeatureScatter(poolB, "nCount_RNA", "percent.mt")

#poolA et B
poolA_et_B[["percent.mt"]] <- PercentageFeatureSet(poolA_et_B, pattern = "^MT-")
VlnPlot(poolA_et_B, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"))
FeatureScatter(poolA_et_B, "nCount_RNA", "percent.mt")
FeatureScatter(poolA_et_B, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
```

## ribo

```{r}
#poolA
poolA[["percent.ribo"]] <- PercentageFeatureSet(poolA, pattern = "^RPS|^RPL")
VlnPlot(poolA, features = c("nFeature_RNA", "nCount_RNA", "percent.ribo"))
FeatureScatter(poolA, "nCount_RNA", "percent.ribo")

#pool B
poolB[["percent.ribo"]] <- PercentageFeatureSet(poolB, pattern = "^RPS|^RPL")
VlnPlot(poolB, features = c("nFeature_RNA", "nCount_RNA", "percent.ribo"))
FeatureScatter(poolB, "nCount_RNA", "percent.ribo")

#poolA et B
poolA_et_B[["percent.ribo"]] <- PercentageFeatureSet(poolA_et_B, pattern = "^RPS|^RPL")
VlnPlot(poolA_et_B, features = c("nFeature_RNA", "nCount_RNA", "percent.ribo"))
FeatureScatter(poolA_et_B, "nCount_RNA", "percent.ribo")
```

# filtering

garder nfeatures < 5000

```{r}
#poolA
poolA <- subset(
  poolA,
  subset = nFeature_RNA > 300 &
    nFeature_RNA < 5000 &
    percent.mt < 10
)
VlnPlot(poolA, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"))
FeatureScatter(poolA, "nCount_RNA", "percent.mt")

#poolB
poolB <- subset(
  poolB,
  subset = nFeature_RNA > 300 &
    nFeature_RNA < 5000 &
    percent.mt < 10
)
VlnPlot(poolB, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"))
FeatureScatter(poolB, "nCount_RNA", "percent.mt")

#poolA et B
poolA_et_B <- subset(
  poolA_et_B,
  subset = nFeature_RNA > 300 &
    nFeature_RNA < 5000 &
    percent.mt < 10
)

VlnPlot(poolA_et_B, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"))
FeatureScatter(poolA_et_B, "nCount_RNA", "percent.mt")
FeatureScatter(poolA_et_B, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
```
# RNA

## Normalisation

normalisation “LogNormalize” that normalizes the feature expression measurements for each cell by the total expression, multiplies this by a scale factor (10,000 by default), and log-transforms the result.

```{r}
#poolA
poolA <- NormalizeData(poolA) 

#poolB
poolB <- NormalizeData(poolB) 

#poolA et B
poolA_et_B <- NormalizeData(poolA_et_B) 
```

## Identification of highly-variable genes

```{r}
#poolA
poolA <- FindVariableFeatures(poolA, nfeatures = 2000)

# Identify the top 10 Highly Variable Genes (HVGs) poolA
top10_poolA <- head(VariableFeatures(poolA), 10)
top10_plot_poolA <- VariableFeaturePlot(poolA)
LabelPoints(plot = top10_plot_poolA, points = top10_poolA, repel = TRUE)

#poolB
poolB <- FindVariableFeatures(poolB, nfeatures = 2000)

# Identify the top 10 Highly Variable Genes (HVGs) poolB
top10_poolB <- head(VariableFeatures(poolB), 10)
top10_plot_poolB <- VariableFeaturePlot(poolB)
LabelPoints(plot = top10_plot_poolB, points = top10_poolB, repel = TRUE)

#poolA et B
poolA_et_B <- FindVariableFeatures(poolA_et_B, nfeatures = 2000)

# Identify the top 10 Highly Variable Genes (HVGs) poolA et B
top10_poolA_et_B <- head(VariableFeatures(poolA_et_B), 10)
top10_plot_poolA_et_B <- VariableFeaturePlot(poolA_et_B)
LabelPoints(plot = top10_plot_poolA_et_B, points = top10_poolA_et_B, repel = TRUE)
```

### nbr gene/cell

```{r}
# Summary table of genes per cell by Sample
VlnPlot(poolA_et_B, features = "nFeature_RNA", group.by = "orig.ident")

# To get the average number per pool:
#tapply(poolA_et_B$nFeature_RNA, poolA_et_B$pool, mean) [cite: 12]
```

### nbr de cell/condition

```{r}
poolA_et_B@meta.data %>%
  count(response) %>%
  ggplot(aes(x = response, y = n, fill = response)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  labs(
    x = "réponse au ttt",
    y = "Nombre de cellules"
  )
```

### nbr de cell/patient

```{r}
poolA_et_B@meta.data %>%
  count(patient) %>%
  ggplot(aes(x = patient, y = n, fill = patient)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  labs(
    x = "patient",
    y = "Nombre de cellules"
  )
```

## scaling

```{r}
#poolA
poolA <- ScaleData(poolA)
#view(poolA@assays$RNA)

#poolB
poolB <- ScaleData(poolB)
#view(poolB@assays$RNA)

#poolA et B
poolA_et_B <- ScaleData(poolA_et_B)
#view(poolA_et_B@assays$RNA)

```

## PCA

voir si on monte pas à 30

```{r}
#poolA
poolA <- RunPCA(poolA, features = VariableFeatures(poolA))

#choisir le nbr de PC

ElbowPlot(poolA, ndims = 50)

# Visualize PCA
VizDimLoadings(poolA, dims = 1:3, reduction = "pca")
DimPlot(poolA, reduction = "pca")

#poolB
poolB <- RunPCA(poolB, features = VariableFeatures(poolB))

#choisir le nbr de PC

ElbowPlot(poolB, ndims = 50)

# Visualize PCA
VizDimLoadings(poolB, dims = 1:3, reduction = "pca")
DimPlot(poolB, reduction = "pca")

#poolA et B
poolA_et_B <- RunPCA(poolA_et_B, features = VariableFeatures(poolA_et_B))

#choisir le nbr de PC

ElbowPlot(poolA_et_B, ndims = 50)

# Visualize PCA
VizDimLoadings(poolA_et_B, dims = 1:3, reduction = "pca")
DimPlot(poolA_et_B, reduction = "pca")
```

## Integration

```{r}
# (keeping layers separate)
poolA_et_B <- IntegrateLayers(
  object = poolA_et_B, 
  method = CCAIntegration, # Good for different donors
  orig.reduction = "pca", 
  new.reduction = "integrated.cca",
  verbose = FALSE
)
```

## clustering

tester plsr resolution 

```{r}
#poolA
poolA <-  FindNeighbors(poolA, dim = 1:30)
poolA <- FindClusters(poolA, resolution = 0.4)

#poolB
poolB <-  FindNeighbors(poolB, dim = 1:30)
poolB <- FindClusters(poolB, resolution = 0.4)

#poolA et B
poolA_et_B <-  FindNeighbors(poolA_et_B, dim = 1:30)
poolA_et_B <- FindClusters(poolA_et_B, resolution = 0.4)

#poolA et B integrated
poolA_et_B <-  FindNeighbors(poolA_et_B, reduction = "integrated.cca", dim = 1:30)
poolA_et_B <- FindClusters(poolA_et_B, resolution = 0.4, cluster.name = "cca_clusters")
```

## UMAP

```{r}
#poolA
poolA <- RunUMAP(poolA, dims = 1:30)
DimPlot(poolA, reduction = "umap", label = TRUE)

#poolB
poolB <- RunUMAP(poolB, dims = 1:30)
DimPlot(poolB, reduction = "umap", label = TRUE)

#poolA et B
poolA_et_B <- RunUMAP(poolA_et_B, dims = 1:30)
DimPlot(poolA_et_B, reduction = "umap", label = TRUE)

##poolA et B integrated
poolA_et_B <- RunUMAP(poolA_et_B, reduction = "integrated.cca", dims = 1:30, reduction.name = "umap.cca")
umap_cluster <- DimPlot(poolA_et_B, reduction = "umap.cca", label = TRUE)
ggsave(umap_cluster, "umap_cluster.jpeg")
```

## t-sne

```{r}
#poolA
poolA <- RunTSNE(poolA, dims = 1:20)
DimPlot(poolA, reduction = "tsne", label = TRUE)

#poolB
poolB <- RunTSNE(poolB, dims = 1:20)
DimPlot(poolB, reduction = "tsne", label = TRUE)

#poolA et B
poolA_et_B <- RunTSNE(poolA_et_B, dims = 1:20)
DimPlot(poolA_et_B, reduction = "tsne", label = TRUE)

#poolA et B integrated
poolA_et_B <- RunTSNE(poolA_et_B, reduction = "integrated.cca", dims = 1:20, reduction.name = "tnse.cca")
DimPlot(poolA_et_B,  reduction = "tnse.cca", label = TRUE)
```

### umap 1 couleur/patient

```{r}
Dimplot(poolA_et_B, reduction = "umap.cca", split.by = "patient")
```

### differente resolution

```{r}
#res 0.5
poolA_et_B <- FindClusters(poolA_et_B, resolution = 0.5, cluster.name = "cca_clusters")
poolA_et_B <- RunUMAP(poolA_et_B, reduction = "integrated.cca", dims = 1:30, reduction.name = "umap.cca")
umap_cluster <- DimPlot(poolA_et_B, reduction = "umap.cca", label = TRUE)
ggsave(umap_cluster, "umap_cluster.jpeg")
#poolA et B integrated
poolA_et_B <- RunTSNE(poolA_et_B, reduction = "integrated.cca", dims = 1:20, reduction.name = "tnse.cca")
DimPlot(poolA_et_B,  reduction = "tnse.cca", label = TRUE)

#res 0.6
poolA_et_B <- FindClusters(poolA_et_B, resolution = 0.6, cluster.name = "cca_clusters")
poolA_et_B <- RunUMAP(poolA_et_B, reduction = "integrated.cca", dims = 1:30, reduction.name = "umap.cca")
umap_cluster <- DimPlot(poolA_et_B, reduction = "umap.cca", label = TRUE)
ggsave(umap_cluster, "umap_cluster.jpeg")
#poolA et B integrated
poolA_et_B <- RunTSNE(poolA_et_B, reduction = "integrated.cca", dims = 1:20, reduction.name = "tnse.cca")
DimPlot(poolA_et_B,  reduction = "tnse.cca", label = TRUE)

#res 0.7
poolA_et_B <- FindClusters(poolA_et_B, resolution = 0.7, cluster.name = "cca_clusters")
poolA_et_B <- RunUMAP(poolA_et_B, reduction = "integrated.cca", dims = 1:30, reduction.name = "umap.cca")
umap_cluster <- DimPlot(poolA_et_B, reduction = "umap.cca", label = TRUE)
ggsave(umap_cluster, "umap_cluster.jpeg")
#poolA et B integrated
poolA_et_B <- RunTSNE(poolA_et_B, reduction = "integrated.cca", dims = 1:20, reduction.name = "tnse.cca")
DimPlot(poolA_et_B,  reduction = "tnse.cca", label = TRUE)

#res 0.8
poolA_et_B <- FindClusters(poolA_et_B, resolution = 0.8, cluster.name = "cca_clusters")
poolA_et_B <- RunUMAP(poolA_et_B, reduction = "integrated.cca", dims = 1:30, reduction.name = "umap.cca")
umap_cluster <- DimPlot(poolA_et_B, reduction = "umap.cca", label = TRUE)
ggsave(umap_cluster, "umap_cluster.jpeg")
#poolA et B integrated
poolA_et_B <- RunTSNE(poolA_et_B, reduction = "integrated.cca", dims = 1:20, reduction.name = "tnse.cca")
DimPlot(poolA_et_B,  reduction = "tnse.cca", label = TRUE)

#res 0.9
poolA_et_B <- FindClusters(poolA_et_B, resolution = 0.9, cluster.name = "cca_clusters")
poolA_et_B <- RunUMAP(poolA_et_B, reduction = "integrated.cca", dims = 1:30, reduction.name = "umap.cca")
umap_cluster <- DimPlot(poolA_et_B, reduction = "umap.cca", label = TRUE)
ggsave(umap_cluster, "umap_cluster.jpeg")
#poolA et B integrated
poolA_et_B <- RunTSNE(poolA_et_B, reduction = "integrated.cca", dims = 1:20, reduction.name = "tnse.cca")
DimPlot(poolA_et_B,  reduction = "tnse.cca", label = TRUE)

#res 1
poolA_et_B <- FindClusters(poolA_et_B, resolution = 1, cluster.name = "cca_clusters")
poolA_et_B <- RunUMAP(poolA_et_B, reduction = "integrated.cca", dims = 1:30, reduction.name = "umap.cca")
umap_cluster <- DimPlot(poolA_et_B, reduction = "umap.cca", label = TRUE)
ggsave(umap_cluster, "umap_cluster.jpeg")
#poolA et B integrated
poolA_et_B <- RunTSNE(poolA_et_B, reduction = "integrated.cca", dims = 1:20, reduction.name = "tnse.cca")
DimPlot(poolA_et_B,  reduction = "tnse.cca", label = TRUE)

#res 1.1
poolA_et_B <- FindClusters(poolA_et_B, resolution = 1.1, cluster.name = "cca_clusters")
poolA_et_B <- RunUMAP(poolA_et_B, reduction = "integrated.cca", dims = 1:30, reduction.name = "umap.cca")
umap_cluster <- DimPlot(poolA_et_B, reduction = "umap.cca", label = TRUE)
ggsave(umap_cluster, "umap_cluster.jpeg")
#poolA et B integrated
poolA_et_B <- RunTSNE(poolA_et_B, reduction = "integrated.cca", dims = 1:20, reduction.name = "tnse.cca")
DimPlot(poolA_et_B,  reduction = "tnse.cca", label = TRUE)

#res 1.2
poolA_et_B <- FindClusters(poolA_et_B, resolution = 1.2, cluster.name = "cca_clusters")
poolA_et_B <- RunUMAP(poolA_et_B, reduction = "integrated.cca", dims = 1:30, reduction.name = "umap.cca")
umap_cluster <- DimPlot(poolA_et_B, reduction = "umap.cca", label = TRUE)
ggsave(umap_cluster, "umap_cluster.jpeg")
#poolA et B integrated
poolA_et_B <- RunTSNE(poolA_et_B, reduction = "integrated.cca", dims = 1:20, reduction.name = "tnse.cca")
DimPlot(poolA_et_B,  reduction = "tnse.cca", label = TRUE)
```


## batch effect check

```{r}

# Check if samples (OB1-OB8) are mixing or separated
p1 <- DimPlot(poolA_et_B, reduction = "umap.cca", group.by = "orig.ident") + ggtitle("UMAP by Sample ID")

# Check if there is a "Pool" effect (PoolA vs PoolB)
# Note: Since you merged poolA and poolB with add.cell.ids, 
# you might need to create a 'pool' column if it's not already in metadata.
# Safer way to assign Pool metadata
poolA_et_B$pool <- ifelse(poolA_et_B$orig.ident %in% c("AML_OB1", "AML_OB2", "AML_OB3", "AML_OB4"), "PoolA", "PoolB")
poolA_et_B$pool <- ifelse(grepl("PoolA", rownames(poolA_et_B@meta.data)), "PoolA", "PoolB")

p2 <- DimPlot(poolA_et_B, reduction = "umap.cca", group.by = "pool") + ggtitle("UMAP by Pool")

p1 + p2

umap_per_patient <- DimPlot(poolA_et_B, reduction = "umap.cca", split.by = "orig.ident", ncol = 4)
ggsave(umap_per_patient, "umap_per_patient.jpeg")
```

## join layer

```{r}
# --- poolA ---
# Join layers first
poolA <- JoinLayers(poolA)

# --- poolB ---
# Join layers first
poolB <- JoinLayers(poolB)

# --- poolA et B ---
# Join layers first
poolA_et_B <- JoinLayers(poolA_et_B)

```

## find marker

```{r}
#poolA
markers_poolA <-  FindAllMarkers(
  poolA,
  only.pos = TRUE,
  min.pct = 0.25,
  logfc.threshold = 0.25
)

#poolB
markers_poolB <-  FindAllMarkers(
  poolB,
  only.pos = TRUE,
  min.pct = 0.25,
  logfc.threshold = 0.25
)

#poolA et B
markers_poolA_et_B <-  FindAllMarkers(
  poolA_et_B,
  only.pos = TRUE,
  min.pct = 0.25,
  logfc.threshold = 0.25
)

write.csv(markers_poolA_et_B, "markers_poolA_et_B_filtered.csv")
```

```{r}
library(dplyr)

top10 <- markers_poolA_et_B %>%
  group_by(cluster) %>%
  slice_max(n = 10, order_by = avg_log2FC)

DoHeatmap(poolA_et_B, features = top10$gene)
```

```{r}
VlnPlot(poolA_et_B, features = top10$gene)
RidgePlot(poolA_et_B, features = top10$gene)
```

```{r}
unique(poolA_et_B$orig.ident)
```

## sex classification

```{r}
library(org.Hs.eg.db)
library(cellXY)


# 1. Extraire les counts bruts
# JoinLayers est crucial ici pour avoir une seule matrice
#poolA_et_B <- JoinLayers(poolA_et_B)
counts_bruts <- GetAssayData(poolA_et_B, assay = "RNA", layer = "counts")

# 2. Conversion en matrice standard (Dense Matrix)
# C'est l'étape qui règle l'erreur "S4 to vector"
counts_matrix <- as.matrix(counts_bruts)

# 3. Vérification des noms de gènes
# cellXY a besoin de SYMBOLS (ex: XIST, pas ENSG000...)
# Si vos gènes sont déjà des Symbols, passez à l'étape 4.
# Sinon, utilisez ce bloc pour convertir :
# ann <- select(org.Hs.eg.db, keys=rownames(counts_matrix), 
#               columns=c("SYMBOL"), keytype="ENSEMBL")
# m <- match(rownames(counts_matrix), ann$ENSEMBL)
# rownames(counts_matrix) <- ann$SYMBOL[m]

# 4. Lancer la classification
# On donne la matrice 'counts_matrix' directement, pas l'objet Seurat
sex_preds <- classifySex(counts_matrix, genome = "Hs")

# 5. Vérifier et Intégrer
table(sex_preds$prediction)

# Ajouter au metadata Seurat
poolA_et_B$predicted_sex <- sex_preds$prediction

# 6. Le Boxplot de validation (comme dans l'exemple)
boxplot(counts_matrix["XIST", ] ~ poolA_et_B$predicted_sex, 
        main="Validation par XIST", ylab="Expression Log", col=c("pink", "lightblue"))
```

```{r}
library(ggplot2)

df_sex <- as.data.frame(table(poolA_et_B$patient, poolA_et_B$predicted_sex))
colnames(df_sex) <- c("Patient", "Sexe", "Nombre_de_Cellules")

ggplot(df_sex, aes(x = Patient, y = Nombre_de_Cellules, fill = Sexe)) +
  geom_bar(stat = "identity") + # "fill" pour avoir des pourcentages (0 à 100%)
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("Female" = "pink", "Male" = "lightblue")) +
  theme_minimal() +
  labs(title = "Répartition du Sexe par Patient", 
       y = "Proportion de cellules", 
       x = "Identifiant Patient")
```

```{r}
library(ggplot2)

# EXtraction de la colonne sex en vecteur simple
sex_vector <- as.character(poolA_et_B$predicted_sex)

# creation d'un count table et convertion en data frame
sex_counts <- as.data.frame(table(sex_vector))
colnames(sex_counts) <- c("Sexe", "Nombre_de_cellules")

ggplot(sex_counts, aes(x = Sexe, y = Nombre_de_cellules, fill = Sexe)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("Female" = "pink", "Male" = "lightblue")) +
  theme_minimal() +
  labs(title = "Répartition Globale du Sexe", 
       subtitle = "Nombre total de cellules identifiées",
       x = "Sexe Prédit", 
       y = "Nombre de cellules")
```

### filtration des gene sex

met 2h30 environ

```{r}
# Comparer Réfractaire vs Répondeur en ignorant l'effet du sexe
# On utilise l'argument 'latent.vars' pour corriger statistiquement le biais
clean_markers <- FindMarkers(
  poolA_et_B, 
  group.by = "response", 
  ident.1 = "refractory", 
  ident.2 = "responder",
  latent.vars = "predicted_sex", # CECI est la clé pour "effacer" le bruit du sexe
  test.use = "LR" # Nécessaire quand on utilise latent.vars
)

# Regardez le nouveau top 10
head(clean_markers, 10)

# Mise à jour du Heatmap
DoHeatmap(poolA_et_B, 
          features = clean_markers, 
          group.by = "response", 
          size = 4) + 
  scale_fill_gradientn(colors = c("blue", "white", "red")) +
  ggtitle("Signature de Résistance (Corrigée du Sexe)")
```

## Cell cycle cyclone

```{r}
library(org.Hs.eg.db)
library(dplyr)
library(ggplot2)

# ==============================================================================
# 1. CONVERSION DES IDS ENSEMBL -> SYMBOLS (SANS "LOC")
# ==============================================================================

# Extraction des IDs originaux
ensembl_ids <- rownames(poolA_et_B)

# Mapping strict vers SYMBOL (HGNC)
mapping <- mapIds(org.Hs.eg.db, 
                  keys = ensembl_ids, 
                  column = "SYMBOL", 
                  keytype = "ENSEMBL", 
                  multiVals = "first")

# On garde le Symbole s'il existe, sinon on garde l'ID Ensembl (évite les trous)
clean_names <- mapping
clean_names[is.na(clean_names)] <- ensembl_ids[is.na(clean_names)]

# Suppression des doublons (Seurat exige des rownames uniques)
clean_names <- clean_names[!duplicated(clean_names)]
poolA_et_B <- subset(poolA_et_B, features = names(clean_names))
rownames(poolA_et_B) <- clean_names

# ==============================================================================
# 2. PRÉPARATION ET SCORING DU CYCLE CELLULAIRE
# ==============================================================================

# Normalisation indispensable avant le scoring
pool_final <- NormalizeData(pool_final)

# Forcer les majuscules pour correspondre aux listes cc.genes de Seurat
rownames(pool_final) <- toupper(rownames(pool_final))

# Identification des gènes de cycle présents dans l'objet
s_genes <- intersect(cc.genes$s.genes, rownames(pool_final))
g2m_genes <- intersect(cc.genes$g2m.genes, rownames(pool_final))

# Calcul du score (CellCycleScoring)
poolA_et_B <- CellCycleScoring(
  pool_final, 
  s.features = s_genes, 
  g2m.features = g2m_genes, 
  set.ident = TRUE
)

head(poolA_et_B@meta.data[, c("S.Score", "G2M.Score", "Phase")])

# Visualisation de la quiescence (G1) vs Prolifération (S/G2M)
VlnPlot(poolA_et_B, features = c("S.Score", "G2M.Score"), group.by = "response")
```

```{r}
library(dplyr)
library(ggplot2)

# On force l'utilisation de dplyr::count pour éviter les conflits
df_plot <- poolA_et_B@meta.data %>%
  dplyr::count(Phase, response) %>% # Attention : Phase avec un P majuscule
  group_by(response) %>%
  mutate(percentage = n / sum(n) * 100)

# Création du graphique en barres empilées (Proportions)
ggplot(df_plot, aes(x = response, y = percentage, fill = Phase)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Répartition des Phases du Cycle Cellulaire",
       y = "Pourcentage de cellules (%)",
       x = "Groupe de réponse")
```

```{r}
poolA_et_B@meta.data %>%
  count(patient ,Phase) %>%
  ggplot(aes(x = patient, y = n, fill = Phase)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_bw() +
  labs(
    x = "samples",
    y = "nbr de cell",
    fill = "cell_cycle"
  )
```

## cell-cycle seurat

```{r}
# Read in the expression matrix The first row is a header row, the first column is rownames
exp.mat <- read.table(file = "/brahms/shared/vignette-data/cell_cycle_vignette_files/nestorawa_forcellcycle_expressionMatrix.txt",
    header = TRUE, as.is = TRUE, row.names = 1)

# A list of cell cycle markers, from Tirosh et al, 2015, is loaded with Seurat.  We can
# segregate this list into markers of G2/M phase and markers of S phase
s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes

poolA_et_B <- <- CellCycleScoring(poolA_et_B, s.features = s.genes, g2m.features = g2m.genes, set.ident = TRUE)

Dimplot(poolA_et_B, )
```

## évolution intra-patient

```{r}
DimPlot(poolA_et_B, reduction = "umap.cca", group.by = "timepoint", split.by = "patient")

DimPlot(poolA_et_B, group.by = "response", split.by = "patient")
```

### proportions par cluster

```{r}
prop.table(table(
  Idents(poolA_et_B),
  poolA_et_B$timepoint,
  poolA_et_B$patient
), margin = 2)
```

### comparer responder vs refractory

```{r}
DimPlot(poolA_et_B, reduction = "umap.cca", group.by = "response")
```

```{r}

#évolution des timepoints par groupe
umap_timepoint_per_response <- DimPlot(poolA_et_B, 
        reduction = "umap.cca", 
        group.by = "timepoint", 
        split.by = "response") + 
  theme_bw() +
  ggtitle("Trajectoire des cellules par groupe de réponse")
ggsave(umap_timepoint_per_response, "umap_timepoint_per_response.jpeg")
```

```{r}
# Transformer 'timepoint' en facteur avec l'ordre souhaité
poolA_et_B$timepoint <- factor(poolA_et_B$timepoint, levels = c("pre", "post"))

umap_response_per_timepoint <- DimPlot(poolA_et_B, reduction = "umap.cca", group.by = "response", split.by = "timepoint")
ggsave(umap_response_per_timepoint, "umap_response_per_timepoint.jpeg")
```

```{r}
library(patchwork) # to put plots side by side

p1 <- DimPlot(poolA_et_B, reduction = "umap.cca", group.by = "response") + ggtitle("Response")
p2 <- DimPlot(poolA_et_B, reduction = "umap.cca", group.by = "seurat_clusters", label = TRUE) + ggtitle("Clusters")

p1 + p2
```

## proportions de cell

```{r}
prop.table(table(
  Idents(poolA_et_B),
  poolA_et_B$response
), margin = 2)
```

## analyse differentiel #PRE vs POST (par patient)

```{r}
Idents(poolA_et_B) <- "timepoint"

markers_pre_post <- FindMarkers(
  poolA_et_B,
  ident.1 = "post",
  ident.2 = "pre",
  subset.ident = "P1"
)

```

```{r}
top10 <- markers_pre_post %>%
  rownames_to_column("gene") %>%
  slice_max(order_by = avg_log2FC, n = 10)

DoHeatmap(poolA_et_B, features = top10$gene) + NoLegend()
```

```{r}
VlnPlot(poolA_et_B, features = top10$gene[1:10], group.by = "timepoint")
```

```{r}
FeaturePlot(poolA_et_B, features = top10$gene[1:10])
```

```{r}
DotPlot(poolA_et_B, features = top10$gene, group.by = "timepoint") + RotatedAxis()
```

```{r}
library(ggplot2)

markers_pre_post$gene <- rownames(markers_pre_post)

ggplot(markers_pre_post, aes(x = avg_log2FC, y = -log10(p_val_adj))) +
  geom_point() +
  theme_minimal() +
  geom_vline(xintercept = c(-0.5, 0.5), linetype = "dashed") +
  geom_hline(yintercept = 2) +
  ggtitle("Volcano plot PRE vs POST")
```

```{r}
# Top genes upregulated POST
top_post <- markers_pre_post %>%
  rownames_to_column("gene") %>%
  filter(avg_log2FC > 0) %>%
  slice_max(n = 20, order_by = avg_log2FC)

# Ajouter score 
#calculates an average expression "score" for a whole list of genes
poolA_et_B <- AddModuleScore(
  poolA_et_B,
  features = list(top_post$gene),
  name = "POST_signature"
)

FeaturePlot(poolA_et_B, features = "POST_signature1")
VlnPlot(poolA_et_B, features = "POST_signature1", group.by = "timepoint")
```

### responder vs refractory

```{r}
# Install if needed: install.packages("openxlsx")
library(openxlsx)

# 1. Set Identity to Response
Idents(poolA_et_B) <- "response"

# 2. Run the Differential Expression (DE)
# We use logfc.threshold = 0.25 (approx 1.2x fold change) 
# and min.pct = 0.1 (gene must be in at least 10% of cells)
resistance_markers <- FindMarkers(
  poolA_et_B,
  ident.1 = "refractory",
  ident.2 = "responder",
  only.pos = TRUE 
)

# 3. Filter for Significance and top 50
# We want highly significant (p < 0.05) and strongest fold change
top50_resistance <- resistance_markers %>%
  rownames_to_column("gene") %>%
  filter(p_val_adj < 0.05) %>%
  slice_max(n = 50, order_by = avg_log2FC)

# 4. Save to Excel
write.xlsx(top50_resistance, file = "Top50_Refractory_Markers_filtered.xlsx")
```

```{r}
top20_resistance <- markers_resistance %>%
  rownames_to_column("gene") %>%
  filter(avg_log2FC > 0) %>%
  slice_max(n = 20, order_by = avg_log2FC)

poolA_et_B <- AddModuleScore(
  poolA_et_B,
  features = list(top_resistance$gene),
  name = "Resistance_signature1"
)

FeaturePlot(poolA_et_B, features = "Resistance_signature1")
VlnPlot(poolA_et_B, features = "Resistance_signature1", group.by = "response")
```

```{r}
Idents(poolA_et_B) <- "response"

DotPlot(poolA_et_B, features = c("PF4V1", "LY6G6F", "THY1", "HBG2", "EPCAM", "ENSG00000289621", "ENSG00000283235", "CMTM5", "TUSC3", "ENSG00000290906"))
```

MAIS

```{r}
Idents(poolA_et_B) <- "patient"

DotPlot(poolA_et_B, features = c("PF4V1", "LY6G6F", "THY1", "HBG2", "EPCAM", "ENSG00000289621", "ENSG00000283235", "CMTM5", "TUSC3", "ENSG00000290906"))
```

tester findmarker p1 vs p4 puis supprimer tt les genes nn communs entre p1 et p4 comme ca on garde les gene marker reponse ET qui se trouve dans p1 et p4

## gene commun p1 et p4
```{r}
# ==============================================================================
# 3. EXTRACTION DE LA SIGNATURE COMMUNE (P1 & P4)
# ==============================================================================

# Marqueurs pour P1 (Réfractaire) vs Répondeurs (P2, P3)
markers_P1 <- FindMarkers(poolA_et_B, ident.1 = "P1", group.by = "patient", 
                          ident.2 = c("P2", "P3"), only.pos = TRUE)

# Marqueurs pour P4 (Réfractaire) vs Répondeurs (P2, P3)
markers_P4 <- FindMarkers(poolA_et_B, ident.1 = "P4", group.by = "patient", 
                          ident.2 = c("P2", "P3"), only.pos = TRUE)

# Intersection : Gènes significatifs (p_val_adj < 0.05) communs aux deux
sig_P1 <- rownames(markers_P1[markers_P1$p_val_adj < 0.05, ])
sig_P4 <- rownames(markers_P4[markers_P4$p_val_adj < 0.05, ])
genes_communs <- intersect(sig_P1, sig_P4)

# Création du tableau final trié par puissance d'expression (Log2FC)
liste_resistance <- markers_P1[genes_communs, ] %>% arrange(desc(avg_log2FC))

# ==============================================================================
# 4. VISUALISATION FINALE
# ==============================================================================

# Afficher les 10 gènes les plus discriminants
top30_genes <- head(rownames(liste_resistance), 30)

DotPlot(poolA_et_B, features = top30_genes, group.by = "patient") + 
  ggtitle("Top 30 : Signature de Résistance Partagée (P1 & P4)")

# Export de la liste en CSV pour ton rapport
write.csv(liste_resistance, "Signature_Resistance_AML_Commune_filtered.csv")
```

### par cluster

```{r}
Idents(poolA_et_B) <- "seurat_clusters"

markers_cluster <- FindMarkers(
  subset(poolA_et_B, idents = 3),
  group.by = "response",
  ident.1 = "refractory",
  ident.2 = "responder"
)
```

### tableau de comptage

```{r}
table(poolA_et_B$patient, poolA_et_B$timepoint)
table(poolA_et_B$response, poolA_et_B$timepoint)
```

# save

```{r}
#poolA
saveRDS(poolA, file = "020_010_042_025_filtered_qc_drop__with_adt_integ_clean.rds")

#poolB
saveRDS(poolB, file = "008_034_095_017_filtered_qc_drop_with_adt_integ_clean.rds")

#poolA et B
saveRDS(poolA_et_B, file = "020_010_042_025_008_034_095_017_filtered_qc_drop_with_adt_integ_clean_03_04_2026.rds")

```

# annotation

##chargement library + atlas
```{r}
library(SingleR)
library(caret)
library(patchwork)
library("ggplot2")
library('viridis')
library('pheatmap')

ref <- readRDS("O:/Equipes_PF_Services/EQ_Duprez/EQ_Duprez/Current Lab Members/Chellal Alyssa/Ref_annotation/AML_downsampled_ref_fixed2_with_umap.rds")
```

```{r}
#poolA
ref_mat_poolA  <- GetAssayData(ref, layer = "data")
test_mat_poolA <- GetAssayData(poolA, layer = "data")

head(test_mat_poolA)
head(ref_mat_poolA)

common_genes_poolA <- intersect(rownames(ref_mat_poolA), rownames(test_mat_poolA))
length(common_genes_poolA)

pred_poolA <- SingleR(test = test_mat_poolA, ref = ref_mat_poolA, labels = ref$Cell.Type, de.method="wilcox", de.n=20)
colnames(pred_poolA)

head(sort(table(pred_poolA$labels), decreasing=TRUE))
table(pred_poolA$labels)
head(pred_poolA$labels)

pred_poolA$scores[1:10,]
plotScoreHeatmap(pred_poolA)

pred2_poolA <- SingleR(test = test_mat_poolA, ref = ref_mat_poolA, labels = ref$HSPC.Cell.Type, de.method="wilcox", de.n=20)

colnames(pred2_poolA)
table(pred2_poolA$labels)
head(sort(table(pred2_poolA$labels), decreasing=TRUE))

pred2_poolA$scores[1:10,]
plotScoreHeatmap(pred2_poolA)

poolA$SingleR_main_poolA  <- pred_poolA$labels
poolA$SingleR_fine_poolA  <- pred2_poolA$labels
poolA@meta.data[["SingleR_main_poolA"]]
poolA@meta.data[["SingleR_fine_poolA"]]

options(ggrepel.max.overlaps = Inf)

DimPlot(
  poolA,
  reduction = "umap.cca",
  group.by = "SingleR_main_poolA",
  label = TRUE,
  repel = TRUE
) + theme_bw()

DimPlot(
  poolA,
  reduction = "umap.cca",
  group.by = "SingleR_fine_poolA",
  label = TRUE,
  repel = TRUE
) + theme_bw()


#poolB
ref_mat_poolB  <- GetAssayData(ref, layer = "data")
test_mat_poolB <- GetAssayData(poolB, layer = "data")

head(test_mat_poolB)
head(ref_mat_poolB)

common_genes_poolB <- intersect(rownames(ref_mat_poolB), rownames(test_mat_poolB))
length(common_genes_poolB)

pred_poolB <- SingleR(test = test_mat_poolB, ref = ref_mat_poolB, labels = ref$Cell.Type, de.method="wilcox", de.n=20)
colnames(pred_poolB)

head(sort(table(pred_poolB$labels), decreasing=TRUE))
table(pred_poolB$labels)
head(pred_poolB$labels)

pred_poolB$scores[1:10,]
plotScoreHeatmap(pred_poolB)

pred2_poolB <- SingleR(test = test_mat_poolB, ref = ref_mat_poolB, labels = ref$HSPC.Cell.Type, de.method="wilcox", de.n=20)

colnames(pred2_poolB)
table(pred2_poolB$labels)
head(sort(table(pred2_poolB$labels), decreasing=TRUE))

pred2_poolB$scores[1:10,]
plotScoreHeatmap(pred2_poolB)

poolB$SingleR_main_poolB  <- pred_poolB$labels
poolB$SingleR_fine_poolB  <- pred2_poolB$labels
poolB@meta.data[["SingleR_main_poolB"]]
poolB@meta.data[["SingleR_fine_poolB"]]

options(ggrepel.max.overlaps = Inf)

DimPlot(
  poolB,
  reduction = "umap.cca",
  group.by = "SingleR_main_poolB",
  label = TRUE,
  repel = TRUE
) + theme_bw()

DimPlot(
  poolB,
  reduction = "umap.cca",
  group.by = "SingleR_fine_poolB",
  label = TRUE,
  repel = TRUE
) + theme_bw()
```

## annotation pool A et B

```{r}
#poolA et B
ref_mat_poolA_et_B  <- GetAssayData(ref, layer = "data")
test_mat_poolA_et_B <- GetAssayData(poolA_et_B, layer = "data")

head(test_mat_poolA_et_B)
head(ref_mat_poolA_et_B)

common_genes_poolA_et_B <- intersect(rownames(ref_mat_poolA_et_B), rownames(test_mat_poolA_et_B))
length(common_genes_poolA_et_B)

pred_poolA_et_B <- SingleR(test = test_mat_poolA_et_B, ref = ref_mat_poolA_et_B, labels = ref$Cell.Type, de.method="wilcox", de.n=20)
colnames(pred_poolA_et_B)

head(sort(table(pred_poolA_et_B$labels), decreasing=TRUE))
table(pred_poolA_et_B$labels)
head(pred_poolA_et_B$labels)

pred_poolA_et_B$scores[1:10,]
plotScoreHeatmap(pred_poolA_et_B)

pred2_poolA_et_B <- SingleR(test = test_mat_poolA_et_B, ref = ref_mat_poolA_et_B, labels = ref$HSPC.Cell.Type, de.method="wilcox", de.n=20)

colnames(pred2_poolA_et_B)
table(pred2_poolA_et_B$labels)
head(sort(table(pred2_poolA_et_B$labels), decreasing=TRUE))

pred2_poolA_et_B$scores[1:10,]
plotScoreHeatmap(pred2_poolA_et_B)

poolA_et_B$SingleR_main_poolA_et_B  <- pred_poolA_et_B$labels
poolA_et_B$SingleR_fine_poolA_et_B  <- pred2_poolA_et_B$labels
poolA_et_B@meta.data[["SingleR_main_poolA_et_B"]]
poolA_et_B@meta.data[["SingleR_fine_poolA_et_B"]]

options(ggrepel.max.overlaps = Inf)

annot_main <- DimPlot(
  poolA_et_B,
  reduction = "umap.cca",
  group.by = "SingleR_main_poolA_et_B",
  label = TRUE,
  repel = TRUE
) + theme_bw()

annot_main <- DimPlot(
  poolA_et_B,
  reduction = "umap.cca",
  group.by = "SingleR_fine_poolA_et_B",
  label = TRUE,
  repel = TRUE
) + theme_bw()

ggsave(annot_main, "annot_main.jpeg")
ggsave(annot_fine, "annot_fine.jpeg")
```

## find marker par cell type

```{r}
#poolA et B 
#cell type
Idents(poolA_et_B) <- "SingleR_main_poolA_et_B"
markers_poolA_et_B_annot_main <-  FindAllMarkers(
  poolA_et_B,
  only.pos = TRUE,
  min.pct = 0.25,
  logfc.threshold = 0.25
)

top30_main <- markers_poolA_et_B_annot_main %>%
  group_by(cluster) %>%
  slice_max(n = 30, order_by = avg_log2FC)

write.csv(markers_poolA_et_B_annot_main, "markers_poolA_et_B_annot_main_filtered.csv")
write.csv(top30_main, "markers_poolA_et_B_annot_main_top30_filtered.csv")

#hspc cell type
Idents(poolA_et_B) <- "SingleR_fine_poolA_et_B"
markers_poolA_et_B_annot_fine <-  FindAllMarkers(
  poolA_et_B,
  only.pos = TRUE,
  min.pct = 0.25,
  logfc.threshold = 0.25
)

top30_fine <- markers_poolA_et_B_annot_fine %>%
  group_by(cluster) %>%
  slice_max(n = 30, order_by = avg_log2FC)

write.csv(markers_poolA_et_B_annot_fine, "markers_poolA_et_B_annot_fine_filtered.csv")
write.csv(top30_fine, "markers_poolA_et_B_annot_fine_filtered.csv")
```

## nbr cell par celltype

```{r}
count_per_cell_type <- ggplot(cell_counts, aes(x = SingleR_fine_poolA_et_B, y = Count, fill = SingleR_fine_poolA_et_B)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Cell Counts per Cell Type", x = "Cell Type", y = "Number of Cells")

ggsave(count_per_cell_type, "cellcount_per_cell_type.jpeg")
```

```{r}
table(poolA_et_B$patient, poolA_et_B$timepoint)
table(poolA_et_B$response, poolA_et_B$timepoint)
```

```{r}
library(patchwork) # to put plots side by side

p1 <- DimPlot(poolA_et_B, reduction = "umap.cca", group.by = "response") + ggtitle("Response")
p2 <- DimPlot(poolA_et_B, reduction = "umap.cca", group.by = "SingleR_main_poolA_et_B", label = TRUE) + ggtitle("Clusters")

p1 + p2
```

```{r}
library(patchwork) 

p1 <- DimPlot(poolA_et_B, reduction = "umap.cca", group.by = "response") + ggtitle("Response")
p2 <- DimPlot(poolA_et_B, reduction = "umap.cca", group.by = "SingleR_fine_poolA_et_B", label = TRUE) + ggtitle("Clusters")

p1 + p2
```

## split par timepoint

```{r}
mes_couleurs = c("B" = "purple", "lateEry" = "yellow", "HSPC" = "Red", "HSC" = "Blue", "T" = "Green")

Idents(poolA_et_B) <- "SingleR_main_poolA_et_B"

DimPlot(poolA_et_B, reduction = "umap.cca", split.by = "timepoint",  label = TRUE, repel = TRUE, cols = mes_couleurs) + theme_bw()

Idents(poolA_et_B) <- "SingleR_fine_poolA_et_B"

DimPlot(poolA_et_B,reduction = "umap.cca", split.by = "timepoint",  label = TRUE, repel = TRUE, cols = mes_couleurs) + theme_bw()
```

## split par reponse

```{r}
Idents(poolA_et_B) <- "SingleR_main_poolA_et_B"

DimPlot(poolA_et_B, reduction = "umap.cca", split.by = "response",  label = TRUE, repel = TRUE, cols = mes_couleurs) + theme_bw()

Idents(poolA_et_B) <- "SingleR_fine_poolA_et_B"

DimPlot(poolA_et_B,reduction = "umap.cca", split.by = "response",  label = TRUE, repel = TRUE, cols = mes_couleurs) + theme_bw()

```

## split par patient

```{r}
mes_couleurs = c("B" = "purple", "lateEry" = "yellow", "HSPC" = "Red", "HSC" = "Blue", "T" = "Green")

Idents(poolA_et_B) <- "SingleR_main_poolA_et_B"

DimPlot(poolA_et_B, reduction = "umap.cca", split.by = "patient",  label = TRUE, repel = TRUE) + theme_bw()

Idents(poolA_et_B) <- "SingleR_fine_poolA_et_B"

DimPlot(poolA_et_B, reduction = "umap.cca", split.by = "patient",  label = TRUE, repel = TRUE, cols = mes_couleurs) + theme_bw()

```

```{r}
Idents(poolA_et_B) <- "SingleR_main_poolA_et_B"
DotPlot(poolA_et_B, features = markers_poolA_et_B_annot_main, cols = c("blue", "red", "green", "purple"), dot.scale = 8, split.by = "patient") +
    RotatedAxis()

library(ggplot2)
library(cowplot)
theme_set(theme_cowplot())

aggregate_aml <- AggregateExpression(poolA_et_B, group.by = c("SingleR_main_poolA_et_B", "patient"), return.seurat = TRUE)

p1 <- CellScatter(aggregate_aml, "HSPC_P2", "HSPC_P3")
p2 <- LabelPoints(plot = p1, repel = TRUE)

p2

aggregate_aml <- AggregateExpression(poolA_et_B, group.by = c("SingleR_main_poolA_et_B", "response"), return.seurat = TRUE)

p3 <- CellScatter(aggregate_aml, highlight = genes.to.label)
p4 <- LabelPoints(plot = p1, points = genes.to.label, repel = TRUE)

p4
```


```{r}
plot <- DimPlot(poolA_et_B, reduction = "umap.cca")
select.cells <- CellSelector(plot = plot)

head(select.cells)

Idents(poolA_et_B, cells = select.cells) <- "NewCells"

# Now, we find markers that are specific to the new cells, and find clear DC markers
newcells.markers <- FindMarkers(poolA_et_B, ident.1 = "NewCells", min.diff.pct = 0.3,
    only.pos = TRUE)
head(newcells.markers)

write.csv(newcells.markers, "gene_marker_cluster_resistant_filtered.csv")

#creation tableau croisé
table(Idents(poolA_et_B), poolA_et_B$SingleR_main_poolA_et_B)

#Exporter vos marqueurs AVEC les types cellulaires dominants
# On extrait les types cellulaires de SingleR uniquement pour vos cellules sélectionnées
types_dans_selection <- poolA_et_B$SingleR_main_poolA_et_B[select.cells]

# On peut voir le type majoritaire
type_majoritaire <- names(which.max(table(types_dans_selection)))

# Ajouter cette info dans votre CSV (via un commentaire ou une colonne dédiée)
newcells.markers$pred_cell_type <- type_majoritaire

write.csv(newcells.markers, "gene_marker_cluster_resistant_with_type_filtered.csv")

#visualisation
# Prendre les 5 meilleurs gènes de votre liste de marqueurs
top10_markers <- rownames(newcells.markers)[1:10]

# Afficher l'expression de ces gènes à travers les types cellulaires de SingleR
DotPlot(poolA_et_B, 
        features = top10_markers, 
        group.by = "SingleR_main_poolA_et_B") + RotatedAxis()

#creation tableau croisé
table(Idents(poolA_et_B), poolA_et_B$SingleR_fine_poolA_et_B)

# Pour voir les gènes utilisés pour distinguer les "Monocytes" des autres par exemple :
genes_utilises <- metadata$de.genes$Monocytes
head(genes_utilises)

#Exporter vos marqueurs AVEC les types cellulaires dominants
# On extrait les types cellulaires de SingleR uniquement pour vos cellules sélectionnées
types_dans_selection <- poolA_et_B$SingleR_fine_poolA_et_B[select.cells]

# On peut voir le type majoritaire
type_majoritaire <- names(which.max(table(types_dans_selection)))

# Ajouter cette info dans votre CSV (via un commentaire ou une colonne dédiée)
newcells.markers$pred_cell_type <- type_majoritaire

write.csv(newcells.markers, "gene_marker_cluster_resistant_with_type_filtered.csv")

#visualisation
# Prendre les 5 meilleurs gènes de votre liste de marqueurs
top5_markers <- rownames(newcells.markers)[1:5]

# Afficher l'expression de ces gènes à travers les types cellulaires de SingleR
DotPlot(poolA_et_B, 
        features = top5_markers, 
        group.by = "SingleR_fine_poolA_et_B") + RotatedAxis()
```

```{r}
Idents(poolA_et_B) <- "SingleR_main_poolA_et_B"

marker_stromal <- FindMarkers(poolA_et_B, ident.1 = "Stromal", only.pos = TRUE, min.pct = 0.25)

marker_B <- FindMarkers(poolA_et_B, ident.1 = "B", only.pos = TRUE, min.pct = 0.25)

marker_CMP <- FindMarkers(poolA_et_B, ident.1 = "CMP", only.pos = TRUE, min.pct = 0.25)
```

```{r}
Idents(poolA_et_B) <- "SingleR_fine_poolA_et_B"

marker_GMP_like <- FindMarkers(poolA_et_B, ident.1 = "GMP-like", only.pos = TRUE, min.pct = 0.25)

marker_B <- FindMarkers(poolA_et_B, ident.1 = "B", only.pos = TRUE, min.pct = 0.25)

marker_late_ery <- FindMarkers(poolA_et_B, ident.1 = "lateEry", only.pos = TRUE, min.pct = 0.25)

DotPlot(poolA_et_B, features = marker_late_ery)
```

```{r}
# Reset Idents to clusters to find the number
Idents(poolA_et_B) <- "seurat_clusters"

# Calculate the percentage of Refractory vs Responder per cluster
comp_table <- table(poolA_et_B$seurat_clusters, poolA_et_B$response)
prop_table <- prop.table(comp_table, margin = 1) * 100

# Sort to see the most "Refractory" clusters at the top
print(prop_table[order(prop_table[, "refractory"], decreasing = TRUE), ])
```

```{r}
# Créer une table croisée Cluster / Patient
patient_cluster_table <- table(poolA_et_B$seurat_clusters, poolA_et_B$patient)

# Convertir en proportions par patient (pour normaliser si un patient a plus de cellules)
patient_prop <- prop.table(patient_cluster_table, margin = 2) * 100

# Afficher les clusters pour vos patients réfractaires (P1 et P4)
print(patient_prop[, c("P1", "P4")])
```

```{r}
Idents(poolA_et_B) <- "response"

#cluster 6
DotPlot(poolA_et_B, features = c("LINC02506", "TNFRSF12A", "HLA-DRB5", "SPINK2", "EMP1", "ID1", "CXCL8", "ABO", "LMNA", "FHL2"))

#cluster 0
DotPlot(poolA_et_B, features = c("TKT", "SERPINB1", "HSPD1", "NPM1", "PDIA6", "TALAM1", "ATP5F1C", "TUBB", "OLA1", "CCT6A"))

#cluster 14
DotPlot(poolA_et_B, features = c("ENSG00000236304", "ENSG00000289621", "SPX", "PF4V1", "PPBP", "LY6G6F", "GP9", "PF4", "TREML1", "CLEC1B"))
```

```{r}
Idents(poolA_et_B) <- "timepoint"

#cluster 6
DotPlot(poolA_et_B, features = c("LINC02506", "TNFRSF12A", "HLA-DRB5", "SPINK2", "EMP1", "ID1", "CXCL8", "ABO", "LMNA", "FHL2"))

#cluster 0
DotPlot(poolA_et_B, features = c("TKT", "SERPINB1", "HSPD1", "NPM1", "PDIA6", "TALAM1", "ATP5F1C", "TUBB", "OLA1", "CCT6A"))

#cluster 14
DotPlot(poolA_et_B, features = c("ENSG00000236304", "ENSG00000289621", "SPX", "PF4V1", "PPBP", "LY6G6F", "GP9", "PF4", "TREML1", "CLEC1B"))
```

```{r}
Idents(poolA_et_B) <- "patient"

#cluster 6
DotPlot(poolA_et_B, features = c("LINC02506", "TNFRSF12A", "HLA-DRB5", "SPINK2", "EMP1", "ID1", "CXCL8", "ABO", "LMNA", "FHL2"))

#cluster 0
DotPlot(poolA_et_B, features = c("TKT", "SERPINB1", "HSPD1", "NPM1", "PDIA6", "TALAM1", "ATP5F1C", "TUBB", "OLA1", "CCT6A"))

#cluster 14
DotPlot(poolA_et_B, features = c("ENSG00000236304", "ENSG00000289621", "SPX", "PF4V1", "PPBP", "LY6G6F", "GP9", "PF4", "TREML1", "CLEC1B"))
```

```{r}
# Cross-tabulate Cluster number with SingleR Fine labels
table(poolA_et_B$seurat_clusters, poolA_et_B$SingleR_fine_poolA_et_B)
```

```{r}
# Create a contingency table of Clusters vs SingleR Fine Annotation
annotation_map <- table(poolA_et_B$seurat_clusters, poolA_et_B$SingleR_fine_poolA_et_B)

# Convert to proportions to see the "dominant" cell type for each cluster
annotation_prop <- prop.table(annotation_map, margin = 1) * 100

# View the map
print(annotation_prop)

library(ggplot2)

# This plot shows the distribution of cell types within each cluster
ggplot(poolA_et_B@meta.data, aes(x = seurat_clusters, fill = SingleR_fine_poolA_et_B)) +
  geom_bar(position = "fill") +
  theme_minimal() +
  RotatedAxis() +
  labs(y = "Proportion", title = "Biological Identity of Numerical Clusters")
```

```{r}
library(ggplot2)

ggplot(poolA_et_B@meta.data, aes(x = response, fill = SingleR_fine_poolA_et_B)) +
  geom_bar(position = "fill") +
  theme_minimal() +
  labs(y = "Proportion of Cells", title = "Cell Type Composition: Responder vs Refractory")
```

Transcription Factor Inference (SCENIC or DoRothEA) Genes don't just turn on by themselves; they are controlled by Transcription Factors (TFs) like SPI1 (PU.1) or RUNX1. Analyzing these tells you the "Master Regulator" of the refractory state.

```{r}
# Quick check: Look for TFs in your marker list
tfs_to_check <- c("SPI1", "RUNX1", "CEBPA", "MYC", "GATA2")
DotPlot(poolA_et_B, features = tfs_to_check, group.by = "response")
```

```{r}
Idents(poolA_et_B) <- "response"

# find refractory markers for every cluster compared to all remaining cells, report only the positive ones
aml.markers <- FindAllMarkers(poolA_et_B, only.pos = TRUE)
aml.markers %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1)

DotPlot(poolA_et_B, features = c("SLC35F1", "RPS24", "RPS26", "RPS4Y1", "DDX3Y", "UTY"))

top20_repsonse_aml <- aml.markers %>%
  group_by(cluster) %>%
  slice_max(n = 20, order_by = avg_log2FC)

DoHeatmap(poolA_et_B, features = top20_repsonse_aml$gene)

# 1. Get the list of genes you want to plot
genes_to_plot <- top20_repsonse_aml$gene

# 2. Scale ONLY these genes (this is fast)
# This adds them to the scale.data layer so DoHeatmap can find them
poolA_et_B <- ScaleData(poolA_et_B, features = genes_to_plot)

# 3. Now run the heatmap again
DoHeatmap(poolA_et_B, features = genes_to_plot, group.by = "response") +
  scale_fill_gradientn(colors = c("blue", "white", "red"))
```

# annotation 2e atlas

```{r}

ref2 <- readRDS("O:/Equipes_PF_Services/EQ_Duprez/OpenAccess/AML_scRNAseq/Atlas/BoneMarrowMap_Annotated_Dataset_2025_ZengDick.rds")
str(ref2)

#poolA et B
ref2_mat_poolA_et_B  <- GetAssayData(ref2, layer = "data")
test2_mat_poolA_et_B <- GetAssayData(poolA_et_B, layer = "data")

head(test2_mat_poolA_et_B)
head(ref2_mat_poolA_et_B)

common_genes2_poolA_et_B <- intersect(rownames(ref2_mat_poolA_et_B), rownames(test2_mat_poolA_et_B))
length(common_genes2_poolA_et_B)

pred3_poolA_et_B <- SingleR(test = test2_mat_poolA_et_B, ref = ref2_mat_poolA_et_B, labels = ref2$CellType, de.method="wilcox", de.n=20)
colnames(pred3_poolA_et_B)

head(sort(table(pred3_poolA_et_B$labels), decreasing=TRUE))
table(pred3_poolA_et_B$labels)
head(pred3_poolA_et_B$labels)

pred3_poolA_et_B$scores[1:10,]
plotScoreHeatmap(pred3_poolA_et_B)


poolA_et_B$SingleR2_main_poolA_et_B  <- pred3_poolA_et_B$labels
poolA_et_B@meta.data[["SingleR2_main_poolA_et_B"]]

options(ggrepel.max.overlaps = Inf)

annot2 <- DimPlot(
  poolA_et_B,
  reduction = "umap.cca",
  group.by = "SingleR2_main_poolA_et_B",
  label = TRUE,
  repel = TRUE
) + theme_bw()

ggsave("annot2.jpeg", annot2,width = 20, height = 10 )

#poolA et B
Idents(poolA_et_B) <- "SingleR2_main_poolA_et_B"
markers_poolA_et_B_annot2_main <-  FindAllMarkers(
  poolA_et_B,
  only.pos = TRUE,
  min.pct = 0.25,
  logfc.threshold = 0.25
)

top30_main2 <- markers_poolA_et_B_annot2_main %>%
  group_by(cluster) %>%
  slice_max(n = 30, order_by = avg_log2FC)

write.csv(markers_poolA_et_B_annot2_main, "markers_poolA_et_B_annot2_main_filtered.csv")
write.csv(top30_main2, "markers_poolA_et_B_annot2_main_top30_filtered.csv")

#poolA et B : broad
ref2_mat_poolA_et_B_broad  <- GetAssayData(ref2, layer = "data")
test2_mat_poolA_et_B_broad <- GetAssayData(poolA_et_B, layer = "data")

head(test2_mat_poolA_et_B_broad)
head(ref2_mat_poolA_et_B_broad)

common_genes2_poolA_et_B_broad <- intersect(rownames(ref2_mat_poolA_et_B_broad), rownames(test2_mat_poolA_et_B_broad))
length(common_genes2_poolA_et_B_broad)

pred3_poolA_et_B_broad <- SingleR(test = test2_mat_poolA_et_B_broad, ref = ref2_mat_poolA_et_B_broad, labels = ref2$CellType_Broad, de.method="wilcox", de.n=20)
colnames(pred3_poolA_et_B_broad)

head(sort(table(pred3_poolA_et_B_broad$labels), decreasing=TRUE))
table(pred3_poolA_et_B_broad$labels)
head(pred3_poolA_et_B_broad$labels)

pred3_poolA_et_B_broad$scores[1:10,]
plotScoreHeatmap(pred3_poolA_et_B_broad)


poolA_et_B$SingleR2_main_poolA_et_B_broad  <- pred3_poolA_et_B_broad$labels
poolA_et_B@meta.data[["SingleR2_main_poolA_et_B_broad"]]

options(ggrepel.max.overlaps = Inf)

annot2_broad <- DimPlot(
  poolA_et_B,
  reduction = "umap.cca",
  group.by = "SingleR2_main_poolA_et_B_broad",
  label = TRUE,
  repel = TRUE
) + theme_bw()

ggsave("annot2_broad.jpeg", annot2_broad,width = 20, height = 10 )

#poolA et B
Idents(poolA_et_B) <- "SingleR2_main_poolA_et_B_broad"
markers_poolA_et_B_annot2_main_broad <-  FindAllMarkers(
  poolA_et_B,
  only.pos = TRUE,
  min.pct = 0.25,
  logfc.threshold = 0.25
)

top30_main2_broad <- markers_poolA_et_B_annot2_main_broad %>%
  group_by(cluster) %>%
  slice_max(n = 30, order_by = avg_log2FC)

write.csv(markers_poolA_et_B_annot2_main_broad, "markers_poolA_et_B_annot2_main_filtered_broad.csv")
write.csv(top30_main2_broad, "markers_poolA_et_B_annot2_main_top30_filtered_broad.csv")
```

# ADT

```{r}
DefaultAssay(poolA_et_B)

DefaultAssay(poolA_et_B) <- "ADT"
DefaultAssay(poolA_et_B)
```

## Normalisation

normalisation “LogNormalize” that normalizes the feature expression measurements for each cell by the total expression, multiplies this by a scale factor (10,000 by default), and log-transforms the result.

```{r}
#poolA
poolA <- NormalizeData(poolA) 

#poolB
poolB <- NormalizeData(poolB) 

#poolA et B
poolA_et_B <- NormalizeData(poolA_et_B) 
```

## Identification of highly-variable genes

```{r}
#poolA
poolA <- FindVariableFeatures(poolA, nfeatures = 2000)

# Identify the top 10 Highly Variable Genes (HVGs) poolA
top10_poolA <- head(VariableFeatures(poolA), 10)
top10_plot_poolA <- VariableFeaturePlot(poolA)
LabelPoints(plot = top10_plot_poolA, points = top10_poolA, repel = TRUE)

#poolB
poolB <- FindVariableFeatures(poolB, nfeatures = 2000)

# Identify the top 10 Highly Variable Genes (HVGs) poolB
top10_poolB <- head(VariableFeatures(poolB), 10)
top10_plot_poolB <- VariableFeaturePlot(poolB)
LabelPoints(plot = top10_plot_poolB, points = top10_poolB, repel = TRUE)

#poolA et B
poolA_et_B <- FindVariableFeatures(poolA_et_B, nfeatures = 2000)

# Identify the top 10 Highly Variable Genes (HVGs) poolA et B
top10_poolA_et_B <- head(VariableFeatures(poolA_et_B), 10)
top10_plot_poolA_et_B <- VariableFeaturePlot(poolA_et_B)
LabelPoints(plot = top10_plot_poolA_et_B, points = top10_poolA_et_B, repel = TRUE)
```

### nbr gene/cell

```{r}
# Summary table of genes per cell by Sample
VlnPlot(poolA_et_B, features = "nFeature_RNA", group.by = "orig.ident")

# To get the average number per pool:
#tapply(poolA_et_B$nFeature_RNA, poolA_et_B$pool, mean) [cite: 12]
```

### nbr de cell/condition

```{r}
poolA_et_B@meta.data %>%
  count(response) %>%
  ggplot(aes(x = response, y = n, fill = response)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  labs(
    x = "réponse au ttt",
    y = "Nombre de cellules"
  )
```

### nbr de cell/condition

```{r}
poolA_et_B@meta.data %>%
  count(response) %>%
  ggplot(aes(x = response, y = n, fill = response)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  labs(
    x = "réponse au ttt",
    y = "Nombre de cellules"
  )
```

## scaling

```{r}
#poolA
poolA <- ScaleData(poolA)
#view(poolA@assays$RNA)

#poolB
poolB <- ScaleData(poolB)
#view(poolB@assays$RNA)

#poolA et B
poolA_et_B <- ScaleData(poolA_et_B)
#view(poolA_et_B@assays$RNA)

```

## PCA

voir si on monte pas à 30

```{r}

#poolA et B
poolA_et_B <- RunPCA(poolA_et_B, features = VariableFeatures(poolA_et_B))

#choisir le nbr de PC

ElbowPlot(poolA_et_B, ndims = 50)

# Visualize PCA
VizDimLoadings(poolA_et_B, dims = 1:3, reduction = "pca")
DimPlot(poolA_et_B, reduction = "pca")
```

## Integration

```{r}
# (keeping layers separate)
poolA_et_B <- IntegrateLayers(
  object = poolA_et_B, 
  method = CCAIntegration, # Good for different donors
  orig.reduction = "pca", 
  new.reduction = "integrated.cca",
  verbose = FALSE
)
```

## clustering

tester plsr resolution 

```{r}
#poolA et B
poolA_et_B <-  FindNeighbors(poolA_et_B, dim = 1:30)
poolA_et_B <- FindClusters(poolA_et_B, resolution = 0.4)

#poolA et B integrated
poolA_et_B <-  FindNeighbors(poolA_et_B, reduction = "integrated.cca", dim = 1:30)
poolA_et_B <- FindClusters(poolA_et_B, resolution = 0.4, cluster.name = "cca_clusters")
```

## UMAP

```{r}
##poolA et B integrated
poolA_et_B <- RunUMAP(poolA_et_B, reduction = "integrated.cca", dims = 1:30, reduction.name = "umap.cca")
umap_cluster <- DimPlot(poolA_et_B, reduction = "umap.cca", label = TRUE)
ggsave(umap_cluster, "umap_cluster.jpeg")
```

## t-sne

```{r}
#poolA
poolA <- RunTSNE(poolA, dims = 1:20)
DimPlot(poolA, reduction = "tsne", label = TRUE)

#poolB
poolB <- RunTSNE(poolB, dims = 1:20)
DimPlot(poolB, reduction = "tsne", label = TRUE)

#poolA et B
poolA_et_B <- RunTSNE(poolA_et_B, dims = 1:20)
DimPlot(poolA_et_B, reduction = "tsne", label = TRUE)

#poolA et B integrated
poolA_et_B <- RunTSNE(poolA_et_B, reduction = "integrated.cca", dims = 1:20, reduction.name = "tnse.cca")
DimPlot(poolA_et_B,  reduction = "tnse.cca", label = TRUE)
```

### differente resolution

```{r}
#res 0.5
poolA_et_B <- FindClusters(poolA_et_B, resolution = 0.5, cluster.name = "cca_clusters")
poolA_et_B <- RunUMAP(poolA_et_B, reduction = "integrated.cca", dims = 1:30, reduction.name = "umap.cca")
umap_cluster <- DimPlot(poolA_et_B, reduction = "umap.cca", label = TRUE)
ggsave(umap_cluster, "umap_cluster.jpeg")
#poolA et B integrated
poolA_et_B <- RunTSNE(poolA_et_B, reduction = "integrated.cca", dims = 1:20, reduction.name = "tnse.cca")
DimPlot(poolA_et_B,  reduction = "tnse.cca", label = TRUE)

#res 0.6
poolA_et_B <- FindClusters(poolA_et_B, resolution = 0.6, cluster.name = "cca_clusters")
poolA_et_B <- RunUMAP(poolA_et_B, reduction = "integrated.cca", dims = 1:30, reduction.name = "umap.cca")
umap_cluster <- DimPlot(poolA_et_B, reduction = "umap.cca", label = TRUE)
ggsave(umap_cluster, "umap_cluster.jpeg")
#poolA et B integrated
poolA_et_B <- RunTSNE(poolA_et_B, reduction = "integrated.cca", dims = 1:20, reduction.name = "tnse.cca")
DimPlot(poolA_et_B,  reduction = "tnse.cca", label = TRUE)

#res 0.7
poolA_et_B <- FindClusters(poolA_et_B, resolution = 0.7, cluster.name = "cca_clusters")
poolA_et_B <- RunUMAP(poolA_et_B, reduction = "integrated.cca", dims = 1:30, reduction.name = "umap.cca")
umap_cluster <- DimPlot(poolA_et_B, reduction = "umap.cca", label = TRUE)
ggsave(umap_cluster, "umap_cluster.jpeg")
#poolA et B integrated
poolA_et_B <- RunTSNE(poolA_et_B, reduction = "integrated.cca", dims = 1:20, reduction.name = "tnse.cca")
DimPlot(poolA_et_B,  reduction = "tnse.cca", label = TRUE)

#res 0.8
poolA_et_B <- FindClusters(poolA_et_B, resolution = 0.8, cluster.name = "cca_clusters")
poolA_et_B <- RunUMAP(poolA_et_B, reduction = "integrated.cca", dims = 1:30, reduction.name = "umap.cca")
umap_cluster <- DimPlot(poolA_et_B, reduction = "umap.cca", label = TRUE)
ggsave(umap_cluster, "umap_cluster.jpeg")
#poolA et B integrated
poolA_et_B <- RunTSNE(poolA_et_B, reduction = "integrated.cca", dims = 1:20, reduction.name = "tnse.cca")
DimPlot(poolA_et_B,  reduction = "tnse.cca", label = TRUE)

#res 0.9
poolA_et_B <- FindClusters(poolA_et_B, resolution = 0.9, cluster.name = "cca_clusters")
poolA_et_B <- RunUMAP(poolA_et_B, reduction = "integrated.cca", dims = 1:30, reduction.name = "umap.cca")
umap_cluster <- DimPlot(poolA_et_B, reduction = "umap.cca", label = TRUE)
ggsave(umap_cluster, "umap_cluster.jpeg")
#poolA et B integrated
poolA_et_B <- RunTSNE(poolA_et_B, reduction = "integrated.cca", dims = 1:20, reduction.name = "tnse.cca")
DimPlot(poolA_et_B,  reduction = "tnse.cca", label = TRUE)

#res 1
poolA_et_B <- FindClusters(poolA_et_B, resolution = 1, cluster.name = "cca_clusters")
poolA_et_B <- RunUMAP(poolA_et_B, reduction = "integrated.cca", dims = 1:30, reduction.name = "umap.cca")
umap_cluster <- DimPlot(poolA_et_B, reduction = "umap.cca", label = TRUE)
ggsave(umap_cluster, "umap_cluster.jpeg")
#poolA et B integrated
poolA_et_B <- RunTSNE(poolA_et_B, reduction = "integrated.cca", dims = 1:20, reduction.name = "tnse.cca")
DimPlot(poolA_et_B,  reduction = "tnse.cca", label = TRUE)

#res 1.1
poolA_et_B <- FindClusters(poolA_et_B, resolution = 1.1, cluster.name = "cca_clusters")
poolA_et_B <- RunUMAP(poolA_et_B, reduction = "integrated.cca", dims = 1:30, reduction.name = "umap.cca")
umap_cluster <- DimPlot(poolA_et_B, reduction = "umap.cca", label = TRUE)
ggsave(umap_cluster, "umap_cluster.jpeg")
#poolA et B integrated
poolA_et_B <- RunTSNE(poolA_et_B, reduction = "integrated.cca", dims = 1:20, reduction.name = "tnse.cca")
DimPlot(poolA_et_B,  reduction = "tnse.cca", label = TRUE)

#res 1.2
poolA_et_B <- FindClusters(poolA_et_B, resolution = 1.2, cluster.name = "cca_clusters")
poolA_et_B <- RunUMAP(poolA_et_B, reduction = "integrated.cca", dims = 1:30, reduction.name = "umap.cca")
umap_cluster <- DimPlot(poolA_et_B, reduction = "umap.cca", label = TRUE)
ggsave(umap_cluster, "umap_cluster.jpeg")
#poolA et B integrated
poolA_et_B <- RunTSNE(poolA_et_B, reduction = "integrated.cca", dims = 1:20, reduction.name = "tnse.cca")
DimPlot(poolA_et_B,  reduction = "tnse.cca", label = TRUE)
```

## batch effect check

```{r}

# Check if samples (OB1-OB8) are mixing or separated
p1 <- DimPlot(poolA_et_B, reduction = "umap.cca", group.by = "orig.ident") + ggtitle("UMAP by Sample ID")

# Check if there is a "Pool" effect (PoolA vs PoolB)
# Note: Since you merged poolA and poolB with add.cell.ids, 
# you might need to create a 'pool' column if it's not already in metadata.
# Safer way to assign Pool metadata
poolA_et_B$pool <- ifelse(poolA_et_B$orig.ident %in% c("AML_OB1", "AML_OB2", "AML_OB3", "AML_OB4"), "PoolA", "PoolB")
poolA_et_B$pool <- ifelse(grepl("PoolA", rownames(poolA_et_B@meta.data)), "PoolA", "PoolB")

p2 <- DimPlot(poolA_et_B, reduction = "umap.cca", group.by = "pool") + ggtitle("UMAP by Pool")

p1 + p2

umap_per_patient <- DimPlot(poolA_et_B, reduction = "umap.cca", split.by = "orig.ident", ncol = 4)
ggsave(umap_per_patient, "umap_per_patient.jpeg")
```

## join layer

```{r}
# --- poolA ---
# Join layers first
poolA <- JoinLayers(poolA)

# --- poolB ---
# Join layers first
poolB <- JoinLayers(poolB)

# --- poolA et B ---
# Join layers first
poolA_et_B <- JoinLayers(poolA_et_B)

```

## find marker

```{r}
#poolA
markers_poolA <-  FindAllMarkers(
  poolA,
  only.pos = TRUE,
  min.pct = 0.25,
  logfc.threshold = 0.25
)

#poolB
markers_poolB <-  FindAllMarkers(
  poolB,
  only.pos = TRUE,
  min.pct = 0.25,
  logfc.threshold = 0.25
)

#poolA et B
markers_poolA_et_B <-  FindAllMarkers(
  poolA_et_B,
  only.pos = TRUE,
  min.pct = 0.25,
  logfc.threshold = 0.25
)

write.csv(markers_poolA_et_B, "markers_poolA_et_B_filtered.csv")
```

```{r}
library(dplyr)

top10 <- markers_poolA_et_B %>%
  group_by(cluster) %>%
  slice_max(n = 10, order_by = avg_log2FC)

DoHeatmap(poolA_et_B, features = top10$gene)
```

## évolution intra-patient

```{r}
DimPlot(poolA_et_B, reduction = "umap.cca", group.by = "timepoint", split.by = "patient")

DimPlot(poolA_et_B, group.by = "response", split.by = "patient")
```

### proportions par cluster

```{r}
prop.table(table(
  Idents(poolA_et_B),
  poolA_et_B$timepoint,
  poolA_et_B$patient
), margin = 2)
```

### comparer responder vs refractory

```{r}
DimPlot(poolA_et_B, reduction = "umap.cca", group.by = "response")
```

```{r}

#évolution des timepoints par groupe
umap_timepoint_per_response <- DimPlot(poolA_et_B, 
        reduction = "umap.cca", 
        group.by = "timepoint", 
        split.by = "response") + 
  theme_bw() +
  ggtitle("Trajectoire des cellules par groupe de réponse")
ggsave(umap_timepoint_per_response, "umap_timepoint_per_response.jpeg")
```

```{r}
# Transformer 'timepoint' en facteur avec l'ordre souhaité
poolA_et_B$timepoint <- factor(poolA_et_B$timepoint, levels = c("pre", "post"))

umap_response_per_timepoint <- DimPlot(poolA_et_B, reduction = "umap.cca", group.by = "response", split.by = "timepoint")
ggsave(umap_response_per_timepoint, "umap_response_per_timepoint.jpeg")
```

```{r}
library(patchwork) # to put plots side by side

p1 <- DimPlot(poolA_et_B, reduction = "umap.cca", group.by = "response") + ggtitle("Response")
p2 <- DimPlot(poolA_et_B, reduction = "umap.cca", group.by = "seurat_clusters", label = TRUE) + ggtitle("Clusters")

p1 + p2
```

## proportions de cell

```{r}
prop.table(table(
  Idents(poolA_et_B),
  poolA_et_B$response
), margin = 2)
```

## gene commun p1 et p4
```{r}
# ==============================================================================
# 3. EXTRACTION DE LA SIGNATURE COMMUNE (P1 & P4)
# ==============================================================================

# Marqueurs pour P1 (Réfractaire) vs Répondeurs (P2, P3)
markers_P1 <- FindMarkers(poolA_et_B, ident.1 = "P1", group.by = "patient", 
                          ident.2 = c("P2", "P3"), only.pos = TRUE)

# Marqueurs pour P4 (Réfractaire) vs Répondeurs (P2, P3)
markers_P4 <- FindMarkers(poolA_et_B, ident.1 = "P4", group.by = "patient", 
                          ident.2 = c("P2", "P3"), only.pos = TRUE)

# Intersection : Gènes significatifs (p_val_adj < 0.05) communs aux deux
sig_P1 <- rownames(markers_P1[markers_P1$p_val_adj < 0.05, ])
sig_P4 <- rownames(markers_P4[markers_P4$p_val_adj < 0.05, ])
genes_communs <- intersect(sig_P1, sig_P4)

# Création du tableau final trié par puissance d'expression (Log2FC)
liste_resistance <- markers_P1[genes_communs, ] %>% arrange(desc(avg_log2FC))

# ==============================================================================
# 4. VISUALISATION FINALE
# ==============================================================================

# Afficher les 10 gènes les plus discriminants
top30_genes <- head(rownames(liste_resistance), 30)

DotPlot(poolA_et_B, features = top30_genes, group.by = "patient") + 
  ggtitle("Top 30 : Signature de Résistance Partagée (P1 & P4)")

# Export de la liste en CSV pour ton rapport
write.csv(liste_resistance, "Signature_Resistance_AML_Commune_filtered.csv")
```

### par cluster

```{r}
Idents(poolA_et_B) <- "seurat_clusters"

markers_cluster <- FindMarkers(
  subset(poolA_et_B, idents = 3),
  group.by = "response",
  ident.1 = "refractory",
  ident.2 = "responder"
)
```

### tableau de comptage

```{r}
table(poolA_et_B$patient, poolA_et_B$timepoint)
table(poolA_et_B$response, poolA_et_B$timepoint)
```

## split par timepoint

```{r}
mes_couleurs = c("B" = "purple", "lateEry" = "yellow", "HSPC" = "Red", "HSC" = "Blue", "T" = "Green")

Idents(poolA_et_B) <- "SingleR_main_poolA_et_B"

DimPlot(poolA_et_B, reduction = "umap.cca", split.by = "timepoint",  label = TRUE, repel = TRUE, cols = mes_couleurs) + theme_bw()

Idents(poolA_et_B) <- "SingleR_fine_poolA_et_B"

DimPlot(poolA_et_B,reduction = "umap.cca", split.by = "timepoint",  label = TRUE, repel = TRUE, cols = mes_couleurs) + theme_bw()
```

## split par reponse

```{r}
Idents(poolA_et_B) <- "SingleR_main_poolA_et_B"

DimPlot(poolA_et_B, reduction = "umap.cca", split.by = "response",  label = TRUE, repel = TRUE, cols = mes_couleurs) + theme_bw()

Idents(poolA_et_B) <- "SingleR_fine_poolA_et_B"

DimPlot(poolA_et_B,reduction = "umap.cca", split.by = "response",  label = TRUE, repel = TRUE, cols = mes_couleurs) + theme_bw()

```

## split par patient

```{r}
mes_couleurs = c("B" = "purple", "lateEry" = "yellow", "HSPC" = "Red", "HSC" = "Blue", "T" = "Green")

Idents(poolA_et_B) <- "SingleR_main_poolA_et_B"

DimPlot(poolA_et_B, reduction = "umap.cca", split.by = "patient",  label = TRUE, repel = TRUE) + theme_bw()

Idents(poolA_et_B) <- "SingleR_fine_poolA_et_B"

DimPlot(poolA_et_B, reduction = "umap.cca", split.by = "patient",  label = TRUE, repel = TRUE, cols = mes_couleurs) + theme_bw()

```

# save final

```{r}
#poolA
#saveRDS(poolA, file = "020_010_042_025_qc_drop_adt_integ_clean_annot.rds")

#poolB
#saveRDS(poolB, file = "008_034_095_017_qc_drop_adt_integ_clean_annot.rds")

#poolA et B
saveRDS(poolA_et_B, file = "020_010_042_025_008_034_095_017_filtered_qc_drop_with_adt_integ_clean_annot_07_04_2026.rds")

```

# julien rds

```{r}
data <- readRDS("O:/E8YK5F~N/E59W7E~8/O3WXU3~7/AE1A36~8/0KZFNR~B/I4C02N~9/AYNOK6~C/AQ9ID3~B.RDS")

str(data)
dim(data)
colnames(data)

Dim
```
